% Initial environment

%\vfill\eject
\chapter{Standard procedures}
\label{initialenv}
\label{builtinchapter}

%\mainindex{initial environment}
%\mainindex{top level environment}

This chapter describes Scheme's built-in procedures.  The
\defining{initial environment} starts out with a number of variables
bound to locations containing useful values, most of which are
primitive procedures that manipulate data.  For example, the variable
\ide{abs} is bound to (a location initially containing) a procedure of
one argument that computes the absolute value of a number, and the
variable \ide{+} is bound to a procedure that computes sums.

The \defining{top level environment} consists of the bindings of the
initial environment plus any bindings created by top level definitions
(see~\ref{topleveldefinitions}).  A conforming program may use a top
level definition to bind {\em any} variable; the program may
subsequently alter such a binding by an assignment
(see~\ref{assignments}).  A conforming implementation shall not permit
a top level definition of any variable, nor any subsequent assignments
to that variable, to modify the behavior of the procedures specified
in this Standard.\label{safebehavior}

\begin{rationale}
Portable programs should be allowed to redefine any variable that is
bound in the top level environment, including those that are bound in
the initial environment.  To provide a guarantee of consistent
behavior, such redefinitions don't change the meaning of procedures in
the initial environment.  For example, if a program redefines the
variable \ide{car}, the behavior of the procedure \ide{cadr} does not
change.
\end{rationale}

A conforming program shall not assign to any variable in the top level
environment unless the program has previously provided a top level
definition of that variable.  Nevertheless, a conforming
implementation shall permit assignments to alter {\em any} bindings in
the top level environment.

\begin{rationale}
Implementations should be allowed to depend on the bindings of the
initial environment.  These bindings can be altered by program
assignments, so the behavior of such assignments is implementation
dependent.  For example, in a conforming implementation, if a program
executes {\cf (set!~car 'chevy)} without previously redefining
\ide{car}, a subsequent reference to \ide{car} will return {\cf
chevy}, and \ide{cadr} may signal an error if invoked.
Implementations may specify useful (but non-portable) behavior for
programs that alter these initial bindings.
\end{rationale}


\section{Booleans}
\label{booleansection}

The standard boolean objects for true and false are written as
\schtrue{} and \schfalse.\sharpindex{t}\sharpindex{f}  What really
matters, though, are the objects that the Scheme conditional
expressions (\ide{if}, \ide{cond}, \ide{and}, \ide{or}, \ide{do})
treat as true\index{true} or false\index{false}.  Of all the standard
Scheme values, only \schfalse{} counts as false in conditional
expressions.  All other standard Scheme values, including \schtrue{},
pairs, the empty list, symbols, numbers, strings, vectors, and
procedures, count as true.

\begin{note}
Programmers accustomed to other dialects of Lisp should be aware that
Scheme distinguishes both \schfalse{} and the empty list from the symbol
\ide{nil}.
\end{note}

\vest Boolean constants evaluate to themselves, so they don't need to be quoted
in programs.

\begin{scheme}
\schtrue         \ev  \schtrue
\schfalse        \ev  \schfalse
'\schfalse       \ev  \schfalse%
\end{scheme}


\begin{entry}{%
\proto{not}{ obj}{procedure}}

\ide{Not} returns \schtrue{} if \var{obj} is false, and returns
\schfalse{} otherwise.

\begin{scheme}
(not \schtrue)   \ev  \schfalse
(not 3)          \ev  \schfalse
(not (list 3))   \ev  \schfalse
(not \schfalse)  \ev  \schtrue
(not '())        \ev  \schfalse
(not (list))     \ev  \schfalse
(not 'nil)       \ev  \schfalse%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{boolean?}{ obj}{procedure}}

\ide{Boolean?} returns \schtrue{} if \var{obj} is either \schtrue{} or
\schfalse{} and returns \schfalse{} otherwise.

\begin{scheme}
(boolean? \schfalse)  \ev  \schtrue
(boolean? 0)          \ev  \schfalse
(boolean? '())        \ev  \schfalse%
\end{scheme}

\end{entry}

 
\section{Equivalence predicates}
\label{equivalencesection}

A \defining{predicate} is a procedure that always returns a boolean
value (\schtrue{} or \schfalse).  An \defining{equivalence predicate} is
the computational analogue of a mathematical equivalence relation (it is
symmetric, reflexive, and transitive).  Of the equivalence predicates
described in this section, \ide{eq?}\ is the finest or most
discriminating, and \ide{equal?}\ is the coarsest.  \ide{Eqv?}\ is
slightly less discriminating than \ide{eq?}.  \todo{Pitman doesn't like
this paragraph.  Lift the discussion from the Maclisp manual.  Explain
why there's more than one predicate.}


\begin{entry}{%
\proto{eqv?}{ \vari{obj} \varii{obj}}{procedure}}

The \ide{eqv?} procedure defines a useful equivalence relation on objects.
Briefly, it returns \schtrue{} if \vari{obj} and \varii{obj} should
normally be regarded as the same object.  This relation is left slightly
open to interpretation, but the following partial specification of
\ide{eqv?} holds for all implementations of Scheme.

The \ide{eqv?} procedure returns \schtrue{} if:

\begin{itemize}
\item \vari{obj} and \varii{obj} are both \schtrue{} or both \schfalse.

\item \vari{obj} and \varii{obj} are both symbols as defined in this
Standard and

\begin{scheme}
(string=? (symbol->string \vari{obj})
          (symbol->string \varii{obj}))
    \ev  \schtrue%
\end{scheme}

\item \vari{obj} and \varii{obj} are both numbers, are numerically
equal (see~\ref{numbersection}, \ide{=}), and are either both
exact\index{exact} or both inexact\index{inexact}.

\item \vari{obj} and \varii{obj} are both characters and are the same
character according to the \ide{char=?} procedure
(see~\ref{charactersection}).

\item both \vari{obj} and \varii{obj} are the empty list.

\item \vari{obj} and \varii{obj} are pairs, vectors, or strings that denote the
same locations in the store (see~\ref{storagemodel}).

\item \vari{obj} and \varii{obj} are procedures whose location tags are
equal (see~\ref{lambda}).
\end{itemize}

The \ide{eqv?} procedure returns \schfalse{} if:

\begin{itemize}
\item \vari{obj} and \varii{obj} are of different types
(see~\ref{disjointness}).

\item one of \vari{obj} and \varii{obj} is \schtrue{} but the other is
\schfalse{}.

\item \vari{obj} and \varii{obj} are symbols but

\begin{scheme}
(string=? (symbol->string \vari{obj})
          (symbol->string \varii{obj}))
    \ev  \schfalse%
\end{scheme}

\item one of \vari{obj} and \varii{obj} is an exact number but the other
is an inexact number.

\item \vari{obj} and \varii{obj} are numbers for which the \ide{=}
procedure returns \schfalse{}.

\item \vari{obj} and \varii{obj} are characters for which the \ide{char=?}
procedure returns \schfalse{}.

\item one of \vari{obj} and \varii{obj} is the empty list but the other
is not.

\item \vari{obj} and \varii{obj} are pairs, vectors, or strings that denote
distinct locations.

\item \vari{obj} and \varii{obj} are procedures that would behave differently
(return a different value or have different side effects) for some arguments.

\end{itemize}

\begin{scheme}
(eqv? 'a 'a)                     \ev  \schtrue
(eqv? 'a 'b)                     \ev  \schfalse
(eqv? 2 2)                       \ev  \schtrue
(eqv? '() '())                   \ev  \schtrue
(eqv? 100000000 100000000)       \ev  \schtrue
(eqv? (cons 1 2) (cons 1 2))     \ev  \schfalse
(eqv? (lambda () 1)
      (lambda () 2))             \ev  \schfalse
(eqv? \#f 'nil)                  \ev  \schfalse
(let ((p (lambda (x) x)))
  (eqv? p p))                    \ev  \schtrue%
\end{scheme}

The following examples illustrate cases in which the above rules do
not fully specify the behavior of \ide{eqv?}.  All that can be said
about such cases is that the value returned by \ide{eqv?} shall be a
boolean.

\begin{scheme}
(eqv? "" "")             \ev  \unspecified
(eqv? '\#() '\#())         \ev  \unspecified
(eqv? (lambda (x) x)
      (lambda (x) x))    \ev  \unspecified
(eqv? (lambda (x) x)
      (lambda (y) y))    \ev  \unspecified%
\end{scheme}

The next set of examples shows the use of \ide{eqv?}\ with procedures
that have local state.  {\cf Gen-counter} must return a distinct
procedure every time, since each procedure has its own internal counter.
{\cf Gen-loser}, however, returns equivalent procedures each time, since
the local state does not affect the value or side effects of the
procedures.

\begin{scheme}
(define gen-counter
  (lambda ()
    (let ((n 0))
      (lambda () (set! n (+ n 1)) n))))
(let ((g (gen-counter)))
  (eqv? g g))           \ev  \schtrue
(eqv? (gen-counter) (gen-counter))
                        \ev  \schfalse
(define gen-loser
  (lambda ()
    (let ((n 0))
      (lambda () (set! n (+ n 1)) 27))))
(let ((g (gen-loser)))
  (eqv? g g))           \ev  \schtrue
(eqv? (gen-loser) (gen-loser))
                        \ev  \unspecified

(letrec ((f (lambda () (if (eqv? f g) 'both 'f)))
         (g (lambda () (if (eqv? f g) 'both 'g)))
  (eqv? f g))
                        \ev  \unspecified

(letrec ((f (lambda () (if (eqv? f g) 'f 'both)))
         (g (lambda () (if (eqv? f g) 'g 'both)))
  (eqv? f g))
                        \ev  \schfalse%
\end{scheme}

Objects of distinct types shall never be regarded as the same object.

\begin{scheme}
(eqv? '() \schfalse)    \ev  \unspecified%
\end{scheme}

Since it is an error\index{error} to modify constant objects (those returned by
literal expressions), implementations are permitted, though not
required, to share structure between constants where appropriate.  Thus
the value of \ide{eqv?} on constants is sometimes
implementation-dependent.

\begin{scheme}
(let ((x '(a)))
  (eqv? x x))                    \ev  \schtrue
(eqv? '(a) '(a))                 \ev  \unspecified
(eqv? "a" "a")                   \ev  \unspecified
(eqv? '(b) (cdr '(a b)))         \ev  \unspecified%
\end{scheme}

\begin{rationale} 
The above definition of \ide{eqv?} allows implementations latitude in
their treatment of procedures and literals:  implementations are free
either to detect or to fail to detect that two procedures or two literals
are equivalent to each other, and can decide whether or not to
merge representations of equivalent objects by using the same pointer or
bit pattern to represent both.
\end{rationale}

\end{entry}


\begin{entry}{%
\proto{eq?}{ \vari{obj} \varii{obj}}{procedure}}

\ide{Eq?}\ is similar to \ide{eqv?}\ except that in some cases it is
capable of discerning distinctions finer than those detectable by
\ide{eqv?}.

\vest \ide{Eq?}\ and \ide{eqv?}\ are guaranteed to have the same
behavior on symbols, booleans, the empty list, pairs, and non-empty
strings and vectors.  \ide{Eq?}'s behavior on numbers and characters is
implementation-dependent, but it will always return either true or
false, and will return true only when \ide{eqv?}\ would also return
true.  \ide{Eq?} may also behave differently from \ide{eqv?} on empty
vectors and empty strings.

\begin{scheme}
(eq? 'a 'a)                     \ev  \schtrue
(eq? '(a) '(a))                 \ev  \unspecified
(eq? (list 'a) (list 'a))       \ev  \schfalse
(eq? "a" "a")                   \ev  \unspecified
(eq? "" "")                     \ev  \unspecified
(eq? '() '())                   \ev  \schtrue
(eq? 2 2)                       \ev  \unspecified
(eq? \#\backwhack{}A \#\backwhack{}A) \ev  \unspecified
(eq? car car)                   \ev  \schtrue
(let ((n (+ 2 3)))
  (eq? n n))      \ev  \unspecified
(let ((x '(a)))
  (eq? x x))      \ev  \schtrue
(let ((x '\#()))
  (eq? x x))      \ev  \schtrue
(let ((p (lambda (x) x)))
  (eq? p p))      \ev  \schtrue%
\end{scheme}

\todo{Needs to be explained better above.  How can this be made to be
not confusing?  A table maybe?}

\begin{rationale} It will usually be possible to implement \ide{eq?}\ much
more efficiently than \ide{eqv?}, for example, as a simple pointer
comparison instead of as some more complicated operation.  One reason is
that it may not be possible to compute \ide{eqv?}\ of two numbers in
constant time, whereas \ide{eq?}\ implemented as pointer comparison will
always finish in constant time.  \ide{Eq?}\ may be used like \ide{eqv?}\
in applications using procedures to implement objects with state since
it obeys the same constraints as \ide{eqv?}.  \end{rationale}

\end{entry}


\begin{entry}{%
\proto{equal?}{ \vari{obj} \varii{obj}}{procedure}}

\ide{Equal?} recursively compares the contents of pairs, vectors, and
strings, applying \ide{eqv?} on other objects such as numbers and symbols.
A rule of thumb is that objects are generally \ide{equal?} if they print
the same.  \ide{Equal?}\ may fail to terminate if its arguments are
circular data structures.

\begin{scheme}
(equal? 'a 'a)                  \ev  \schtrue
(equal? '(a) '(a))              \ev  \schtrue
(equal? '(a (b) c)
        '(a (b) c))             \ev  \schtrue
(equal? "abc" "abc")            \ev  \schtrue
(equal? 2 2)                    \ev  \schtrue
(equal? (make-vector 5 'a)
        (make-vector 5 'a))     \ev  \schtrue
(equal? (lambda (x) x)
        (lambda (y) y))  \ev  \unspecified%
\end{scheme}

\end{entry}


\section{Pairs and lists}
\label{listsection}

A \defining{pair} (sometimes called a \defining{dotted pair}) is a
record structure with two fields called the car and cdr fields (for
historical reasons).  Pairs are created by the procedure \ide{cons}.
The car and cdr fields are accessed by the procedures \ide{car} and
\ide{cdr}.  The car and cdr fields are assigned by the procedures
\ide{set-car!}\ and \ide{set-cdr!}.

Pairs are used primarily to represent lists.  A list can
be defined recursively as either the empty list\index{empty list} or a pair whose
cdr is a list.  More precisely, the set of lists is defined as the smallest
set \var{X} such that

\begin{itemize}
\item The empty list is in \var{X}.
\item If \var{list} is in \var{X}, then any pair whose cdr field contains
      \var{list} is also in \var{X}.
\end{itemize}

The objects in the car fields of successive pairs of a list are the
elements of the list.  For example, a two-element list is a pair whose car
is the first element and whose cdr is a pair whose car is the second element
and whose cdr is the empty list.  The length of a list is the number of
elements, which is the same as the number of pairs.

The empty list\mainindex{empty list} is a special object of its own type
(see~\ref{disjointness}); it has no elements and its length is zero.

\begin{note}
The above definitions imply that all lists have finite length and are
terminated by the empty list.
\end{note}

The most general notation (external representation) for Scheme pairs is
the ``dotted'' notation \hbox{\cf (\vari{c} .\ \varii{c})} where
\mainindex{dotted notation}\mainschindex{.}
\vari{c} is the value of the car field and \varii{c} is the value of the
cdr field.  For example {\cf (4 .\ 5)} is a pair whose car is 4 and whose
cdr is 5.  Note that {\cf (4 .\ 5)} is the external representation of a
pair, not an expression that evaluates to a pair.

A more streamlined notation can be used for lists: the elements of the
list are simply enclosed in parentheses and separated by spaces.  The
empty list\index{empty list} is written {\tt()} .  For example,

\begin{scheme}
(a b c d e)%
\end{scheme}

and

\begin{scheme}
(a . (b . (c . (d . (e . ())))))%
\end{scheme}

are equivalent notations for a list of symbols.

A chain of pairs not ending in the empty list is called an
\defining{improper list}.  Note that an improper list is not a list.
The list and dotted notations can be combined to represent
improper lists:

\begin{scheme}
(a b c . d)%
\end{scheme}

is equivalent to

\begin{scheme}
(a . (b . (c . d)))%
\end{scheme}

Whether a given pair is a list depends upon what is stored in the cdr
field.  When the \ide{set-cdr!} procedure is used, an object can be a
list one moment and not the next:

\begin{scheme}
(define x (list 'a 'b 'c))
(define y x)
y                       \ev  (a b c)
(list? y)               \ev  \schtrue
(set-cdr! x 4)          \ev  \unspecified
x                       \ev  (a . 4)
(eqv? x y)              \ev  \schtrue
y                       \ev  (a . 4)
(list? y)               \ev  \schfalse
(set-cdr! x x)          \ev  \unspecified
(list? x)               \ev  \schfalse%
\end{scheme}

%It is often convenient to speak of a homogeneous list of objects
%of some particular data type, as for example \hbox{\cf (1 2 3)} is a list of
%integers.  To be more precise, suppose \var{D} is some data type.  (Any
%predicate defines a data type consisting of those objects of which the
%predicate is true.)  Then
%
%\begin{itemize}
%\item The empty list is a list of \var{D}.
%\item If \var{list} is a list of \var{D}, then any pair whose cdr is
%      \var{list} and whose car is an element of the data type \var{D} is also a
%      list of \var{D}.
%\item There are no other lists of \var{D}.
%\end{itemize}

Within literal expressions and representations of objects read by the
\ide{read} procedure, the forms \singlequote\hyper{datum}\schindex{'},
\backquote\hyper{datum}, {\tt,}\hyper{datum}\schindex{,}, and
{\tt,@}\hyper{datum}\schindex{,@} denote two-ele\-ment lists whose
first elements are the symbols \ide{quote}, \ide{quasiquote},
\hbox{\ide{unquote}}, and
\ide{unquote-splicing}, respectively.  The second element in each case
is \hyper{datum}.  This convention is supported so that arbitrary Scheme
programs may be represented as lists.  \todo{Can or need this be stated
more carefully?} That is, according to Scheme's grammar, every
\meta{expression} is also a \meta{datum} (see~\ref{datum}).
Among other things, this permits the use of the \ide{read} procedure to
parse Scheme programs.  See~\ref{externalreps}. 
 

\begin{entry}{%
\proto{pair?}{ obj}{procedure}}

\ide{Pair?} returns \schtrue{} if \var{obj} is a pair, and otherwise
returns \schfalse.

\begin{scheme}
(pair? '(a . b))        \ev  \schtrue
(pair? '(a b c))        \ev  \schtrue
(pair? '())             \ev  \schfalse
(pair? '\#(a b))         \ev  \schfalse%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{cons}{ \vari{obj} \varii{obj}}{procedure}}

Returns a newly allocated pair whose car is \vari{obj} and whose cdr is
\varii{obj}.  The pair is guaranteed to be different (in the sense of
\ide{eqv?}) from every existing object.

\begin{scheme}
(cons 'a '())           \ev  (a)
(cons '(a) '(b c d))    \ev  ((a) b c d)
(cons "a" '(b c))       \ev  ("a" b c)
(cons 'a 3)             \ev  (a . 3)
(cons '(a b) 'c)        \ev  ((a b) . c)%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{car}{ pair}{procedure}}

\nodomain{\var{Pair} must be a pair.}
Returns the contents of the car field of \var{pair}.  Note that it is an
error\index{error} to take the car of the empty list\index{empty list}.

\begin{scheme}
(car '(a b c))          \ev  a
(car '((a) b c d))      \ev  (a)
(car '(1 . 2))          \ev  1
(car '())               \ev  \scherror%
\end{scheme}
 
\end{entry}


\begin{entry}{%
\proto{cdr}{ pair}{procedure}}

\nodomain{\var{Pair} must be a pair.}
Returns the contents of the cdr field of \var{pair}.
Note that it is an error\index{error} to take the cdr of the empty list.

\begin{scheme}
(cdr '((a) b c d))      \ev  (b c d)
(cdr '(1 . 2))          \ev  2
(cdr '())               \ev  \scherror%
\end{scheme}
 
\end{entry}


\begin{entry}{%
\proto{set-car!}{ pair obj}{procedure}}

\nodomain{\var{Pair} must be a pair.}  
Stores \var{obj} in the car field of \var{pair}.
The value returned by \ide{set-car!}\ is unspecified\index{unspecified}.  % <!>
%This procedure can be very confusing if used indiscriminately.

\begin{scheme}
(define (f) (list 'not-a-constant-list))
(define (g) '(constant-list))
(set-car! (f) 3)             \ev  \unspecified
(set-car! (g) 3)             \ev  \scherror ; constant list%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{set-cdr!}{ pair obj}{procedure}}

\nodomain{\var{Pair} must be a pair.}
Stores \var{obj} in the cdr field of \var{pair}.
The value returned by \ide{set-cdr!}\ is unspecified\index{unspecified}.  % <!>
%This procedure can be very confusing if used indiscriminately.

\end{entry}

\setbox0\hbox{\tt(cadr \var{pair})}
\setbox1\hbox{procedure}


\begin{entry}{%
\proto{caar}{ pair}{procedure}
\proto{cadr}{ pair}{procedure}
\pproto{\hbox to 1\wd0 {\hfil$\vdots$\hfil}}{\hbox to 1\wd1 {\hfil$\vdots$\hfil}}
\proto{cdddar}{ pair}{procedure}
\proto{cddddr}{ pair}{procedure}}

These procedures are compositions of \ide{car} and \ide{cdr}, where
for example \ide{caddr} could be defined by

\begin{scheme}
(define caddr (lambda (x) (car (cdr (cdr x))))){\rm.}%
\end{scheme}

Arbitrary compositions, up to four deep, are provided.  There are
twenty-eight of these procedures in all.

\end{entry}


\begin{entry}{%
\proto{null?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is the empty list\index{empty list},
otherwise returns \schfalse.

\end{entry}

\begin{entry}{%
\proto{list?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is a list, otherwise returns \schfalse{}.
By definition, all lists have finite length and are terminated by
the empty list.  This procedure must return an answer even for
circular structures.

\begin{scheme}
(list? '(a b c))                \ev  \schtrue
(list? '())                     \ev  \schtrue
(list? '(a . b))                \ev  \schfalse
(let ((x (list 'a)))
  (set-cdr! x x)
  (list? x))                    \ev  \schfalse%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{list}{ \var{obj} \dotsfoo}{procedure}}

Returns a list of its arguments.

\begin{scheme}
(list 'a (+ 3 4) 'c)            \ev  (a 7 c)
(list)                          \ev  ()%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{length}{ list}{procedure}}

\nodomain{\var{List} must be a list.}
Returns the length of \var{list}.

\begin{scheme}
(length '(a b c))               \ev  3
(length '(a (b) (c d e)))       \ev  3
(length '())                    \ev  0%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{append}{ list \dotsfoo}{procedure}}

\nodomain{All \var{list}s must be lists.}
Returns a list consisting of the elements of the first \var{list}
followed by the elements of the other \var{list}s.

\begin{scheme}
(append '(x) '(y))              \ev  (x y)
(append '(a) '(b c d))          \ev  (a b c d)
(append '(a (b)) '((c)))        \ev  (a (b) (c))
(append)                        \ev  ()%
\end{scheme}

The resulting list is always newly allocated, except that it shares
structure with the last \var{list} argument.  The last argument may
actually be any object; an improper list results if the last argument is not a
proper list.  \todo{This is pretty awkward.  I should get Bartley to fix this.}

\begin{scheme}
(append '(a b) '(c . d))        \ev  (a b c . d)
(append '() 'a)                 \ev  a%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{reverse}{ list}{procedure}}

\nodomain{\var{List} must be a list.}
Returns a newly allocated list consisting of the elements of \var{list}
in reverse order.

\begin{scheme}
(reverse '(a b c))              \ev  (c b a)
(reverse '(a (b c) d (e (f))))  \lev  ((e (f)) d (b c) a)%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{list-ref}{ list \vr{k}}{procedure}}

Returns the \vr{k}th element of \var{list}, using zero-origin indexing.

\begin{scheme}
(list-ref '(a b c d) 2)                 \ev  c
(list-ref '(a b c d)
          (inexact->exact (round 1.8))) \lev  c%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{memq}{ obj list}{procedure}
\proto{memv}{ obj list}{procedure}
\proto{member}{ obj list}{procedure}}

These procedures return the first pair of \var{list} whose car is
\var{obj}; the returned pair is always one from which
\var{list} is composed.
If \var{obj} does not occur in \var{list}, \schfalse{} (n.b.:~not the empty list) is
returned.  \ide{Memq} uses \ide{eq?}\ to compare \var{obj} with the elements of
\var{list}, while \ide{memv} uses \ide{eqv?}\ and \ide{member} uses \ide{equal?}.

\begin{scheme}
(memq 'a '(a b c))              \ev  (a b c)
(memq 'b '(a b c))              \ev  (b c)
(memq 'a '(b c d))              \ev  \schfalse
(memq (list 'a) '(b (a) c))     \ev  \schfalse
(member (list 'a)
        '(b (a) c))             \ev  ((a) c)
(memq 101 '(100 101 102))       \ev  \unspecified
(memv 101 '(100 101 102))       \ev  (101 102)%
\end{scheme} 
 
\end{entry}


\begin{entry}{%
\proto{assq}{ obj alist}{procedure}
\proto{assv}{ obj alist}{procedure}
\proto{assoc}{ obj alist}{procedure}}

\domain{\var{Alist} (for ``association list'') must be a list of
pairs.}  These procedures find the first pair in \var{alist} whose car field is \var{obj},
and returns that pair; the returned pair is always an {\em element} of
\var{alist}, {\em not} one of the pairs from which \var{alist} is composed.
If no pair in \var{alist} has \var{obj} as its
car, \schfalse{} is returned.  \ide{Assq} uses \ide{eq?}\ to compare \var{obj}
with the car fields of the pairs in \var{alist}, while \ide{assv} uses
\ide{eqv?}\ and \ide{assoc} uses \ide{equal?}.

\begin{scheme}
(define e '((a 1) (b 2) (c 3)))
(assq 'a e)     \ev  (a 1)
(assq 'b e)     \ev  (b 2)
(assq 'd e)     \ev  \schfalse
(assq (list 'a) '(((a)) ((b)) ((c))))
                \ev  \schfalse
(assoc (list 'a) '(((a)) ((b)) ((c))))   
                           \ev  ((a))
(assq 5 '((2 3) (5 7) (11 13)))    
                           \ev  \unspecified
(assv 5 '((2 3) (5 7) (11 13)))    
                           \ev  (5 7)%
\end{scheme}


\begin{rationale}
Although they are frequently used as predicates,
\ide{memq}, \ide{memv}, \ide{member}, \ide{assq}, \ide{assv}, and \ide{assoc} do not
have question marks in their names because they return useful values rather
than just \schtrue{} or \schfalse{}.
\end{rationale}
\end{entry}


\section{Symbols}
\label{symbolsection}

Symbols are objects whose usefulness rests on the fact that two
symbols are identical (in the sense of \ide{eqv?}) if and only if their
names are spelled the same way.  This is exactly the property needed to
represent identifiers\index{identifier} in programs, and so most
implementations of Scheme use them internally for that purpose.  Symbols
are useful for many other applications; for instance, they may be used
the way enumerated values are used in Pascal.

\vest The rules for writing a symbol are exactly the same as the rules for
writing an identifier; see~\ref{syntaxsection}
and~\ref{identifiersyntax}.

\vest It is guaranteed that any symbol that has been returned as part of
a literal expression, or read using the \ide{read} procedure, and
subsequently written out using the \ide{write} procedure, will read back
in as the identical symbol (in the sense of \ide{eqv?}).  The
\ide{string\coerce{}symbol} procedure, however, can create symbols for
which this write/read invariance may not hold because their names
contain special characters or letters in the non-standard case.


\begin{entry}{%
\proto{symbol?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is a symbol, otherwise returns \schfalse.

\begin{scheme}
(symbol? 'foo)          \ev  \schtrue
(symbol? (car '(a b)))  \ev  \schtrue
(symbol? "bar")         \ev  \schfalse
(symbol? 'nil)          \ev  \schtrue
(symbol? '())           \ev  \schfalse
(symbol? \schfalse)     \ev  \schfalse%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{symbol->string}{ symbol}{procedure}}

Returns the name of \var{symbol} as a string.  If the symbol was part of
an object returned as the value of a literal expression
(see~\ref{literalsection}) or by a call to the \ide{read} procedure,
and its name contains alphabetic characters, then the string returned
will contain characters in the implementation's preferred standard
case---some implementations will prefer upper case, others lower case.
If the symbol was returned by \ide{string\coerce{}symbol}, the case of
characters in the string returned will be the same as the case in the
string that was passed to \ide{string\coerce{}symbol}.  It is an error\index{error}
to apply mutation procedures like \ide{string-set!} to strings returned
by this procedure.

The following examples assume that the implementation's standard case is
lower case:

\begin{scheme}
(symbol->string 'flying-fish)     
                                  \ev  "flying-fish"
(symbol->string 'Martin)          \ev  "martin"
(symbol->string
   (string->symbol "Malvina"))     
                                  \ev  "Malvina"%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{string->symbol}{ string}{procedure}}

Returns the symbol whose name is \var{string}.  This procedure can
create symbols with names containing special characters or letters in
the non-standard case, but it is usually a bad idea to create such
symbols because in some implementations of Scheme they cannot be read as
themselves.  See \ide{symbol\coerce{}string}.

The following examples assume that the implementation's standard case is
lower case:

\begin{scheme}
(eq? 'mISSISSIppi 'mississippi)  \lev  \schtrue
(string->symbol "mISSISSIppi")  \lev%
  {\rm{}the symbol with name} "mISSISSIppi"
(eq? 'bitBlt (string->symbol "bitBlt"))     \lev  \schfalse
(eq? 'JollyWog
     (string->symbol
       (symbol->string 'JollyWog)))  \lev  \schtrue
(string=? "K. Harper, M.D."
          (symbol->string
            (string->symbol "K. Harper, M.D.")))  \lev  \schtrue%
\end{scheme}

\end{entry}


\section{Numbers}
\label{numbersection}
\index{number}

Numerical computation has traditionally been neglected by the Lisp
community.  Until Common Lisp there was no carefully thought out
strategy for organizing numerical computation, and with the exception of
the MacLisp system \cite{Pitman83} little effort was made to
execute numerical code efficiently.  This Standard recognizes the excellent work
of the Common Lisp committee and accepts many of their recommendations.
In some ways this Standard simplifies and generalizes their proposals in a manner
consistent with the purposes of Scheme.

It is important to distinguish between the mathematical numbers, the
Scheme numbers that attempt to model them, the machine representations
used to implement the Scheme numbers, and notations used to write numbers.
This Standard uses the types \type{number}, \type{complex}, \type{real},
\type{rational}, and \type{integer} to refer to both mathematical numbers
and Scheme numbers.  Machine representations such as fixed point and
floating point are referred to by names such as \type{fixnum} and
\type{flonum}.

%%R4%% I did some reorganizing here to move the discussion of mathematical
% numbers before the discussion of the Scheme numbers, hoping that this
% would help to motivate the discussion of representation independence.

\subsection{Numerical types}
\label{numericaltypes}
\index{numerical types}

%%R4%% A Scheme system provides data of type \type{number}, which is the most
%general numerical type supported by that system.
%\type{Number} is
%likely to be a complicated union type implemented in terms of
%\type{fixnum}s, \type{bignum}s, \type{flonum}s, and so forth, but this
%should not be apparent to a naive user.  What the user should see is
%that the usual operations on numbers produce the mathematically
%expected results, within the limits of the implementation.

%%R4%%  I rewrote the following paragraph to make the various levels of
% the tower into subsets of each other, instead of relating them by
% injections.  I think the injections tended to put people in the frame
% of mind of thinking about coercions between non-overlapping numeric
% types in mainstream programming languages.

\vest Mathematically, numbers may be arranged into a tower of subtypes
%%R4%% with injections relating adjacent levels of the tower:
in which each level is a subset of the level above it:
$$\begin{tabular}{l}
\tupe{number} \\
\tupe{complex} \\
\tupe{real} \\
\tupe{rational} \\
\tupe{integer} 
\end{tabular} $$

For example, 3 is an integer.  Therefore 3 is also a rational,
a real, and a complex.  The same is true of the Scheme numbers
that model 3.  For Scheme numbers, these types are defined by the
predicates \ide{number?}, \ide{complex?}, \ide{real?}, \ide{rational?},
and \ide{integer?}.

There is no simple relationship between a number's type and its
representation inside a computer.  Although most implementations of
Scheme will offer at least two different representations of 3, these
different representations denote the same integer.

%%R4%% I moved "Implementations of Scheme are not required to implement
% the whole tower..." to the subsection on implementation restrictions.

Scheme's numerical operations treat numbers as abstract data, as
independent of their representation as possible.  Although an implementation
of Scheme may use fixnum, flonum, and perhaps other representations for
numbers, this should not be apparent to a casual programmer writing
simple programs.

It is necessary, however, to distinguish between numbers that are
represented exactly and those that may not be.  For example, indexes
into data structures must be known exactly, as must some polynomial
coefficients in a symbolic algebra system.  On the other hand, the
results of measurements are inherently inexact, and irrational numbers
may be approximated by rational and therefore inexact approximations.
In order to catch uses of inexact numbers where exact numbers are
required, Scheme explicitly distinguishes exact from inexact numbers.
This distinction is orthogonal to the dimension of type.

\subsection{Exactness}

%%R4%% I tried to direct the following paragraph away from philosophizing
% about the exactness of mathematical numbers, and toward philosophizing
% about the exactness of Scheme numbers.

\mainindex{exactness} \label{exactly}
Scheme numbers are either \type{exact} or \type{inexact}.  A number is
\tupe{exact} if it was written as an exact constant or was derived from
\tupe{exact} numbers using only \tupe{exact} operations.  A number is
\tupe{inexact} if it was written as an inexact constant,
%%R4%% models a quantity (e.g., a measurement) known only approximately,
if it was
derived using \tupe{inexact} ingredients, or if it was derived using
\tupe{inexact} operations. Thus \tupe{inexact}ness is a contagious
property of a number.
%%R4%% The rest of this paragraph (from R3RS) has been dropped.

\vest If two implementations produce \tupe{exact} results for a
computation that did not involve \tupe{inexact} intermediate results,
the two ultimate results will be mathematically equivalent.  This is
generally not true of computations involving \tupe{inexact} numbers
since approximate methods such as floating point arithmetic may be used,
but it is the duty of each implementation to make the result as close as
practical to the mathematically ideal result.

\vest Rational operations such as \ide{+} should always produce
\tupe{exact} results when given \tupe{exact} arguments.
%%R4%%If an implementation is
%unable to represent an \tupe{exact} result (for example, if it does not
%support infinite precision integers and rationals)
If the operation is unable to produce an \tupe{exact} result,
then it may either report the violation of an implementation restriction
or it may silently coerce its
result to an \tupe{inexact} value.
%%R4%%Such a coercion may cause an error later.
See~\ref{restrictions}.

\vest With the exception of \ide{inexact->exact}, the operations described in
this section shall generally return inexact results when given any inexact
arguments.  An operation may, however, return an \tupe{exact} result if it can
prove that the value of the result is unaffected by the inexactness of its
arguments.  For example, multiplication of any number by an \tupe{exact} zero
may produce an \tupe{exact} zero result, even if the other argument is
\tupe{inexact}.

\subsection{Implementation restrictions}

\index{implementation restriction}\label{restrictions}

\vest Implementations of Scheme are not required to implement the whole
tower of subtypes given in~\ref{numericaltypes},
but they shall implement a coherent subset consistent with both the
purposes of the implementation and the spirit of the Scheme language.
For example, an implementation in which all numbers are \tupe{real}
may still be quite useful.

\vest Implementations may also support only a limited range of numbers of
any type, subject to the requirements of this section.  The supported
range for \tupe{exact} numbers of any type may be different from the
supported range for \tupe{inexact} numbers of that type.  For example,
an implementation that uses flonums to represent all its
\tupe{inexact} \tupe{real} numbers may
support an effectively unbounded range of \tupe{exact} \tupe{integer}s
and \tupe{rational}s
while limiting the range of \tupe{inexact} \tupe{real}s (and therefore
the range of \tupe{inexact} \tupe{integer}s and \tupe{rational}s)
to the dynamic range of the flonum format.
Furthermore
the gaps between the representable \tupe{inexact} \tupe{integer}s and
\tupe{rational}s are
likely to be very large in such an implementation as the limits of this
range are approached.

\vest An implementation of Scheme shall support exact integers
throughout the range of numbers that may be used for indexes of
lists, vectors, and strings or that may result from computing the length of a
list, vector, or string.  The \ide{length}, \ide{vector-length},
and \ide{string-length} procedures shall return an exact
integer, and it is an error\index{error} to use anything but an exact integer as an
index.  Furthermore any integer constant within the index range, if
expressed by an exact integer syntax, will indeed be read as an exact
integer, regardless of any implementation restrictions that may apply
outside this range.  Finally, the procedures listed below shall
return an exact integer result provided all their arguments are exact integers
and the mathematically expected result is representable as an exact integer
within the implementation:

\begin{scheme}
+            -             *
quotient     remainder     modulo
max          min           abs
numerator    denominator   gcd
lcm          floor         ceiling
truncate     round         rationalize
expt         make-polar    make-rectangular
real-part    imag-part     magnitude
angle        inexact->exact%
\end{scheme}

\vest Implementations should support
\tupe{exact} \tupe{integer}s and \tupe{exact} \tupe{rational}s of
practically unlimited size and precision, and to implement the
above procedures and the \ide{/} procedure in
such a way that they always return \tupe{exact} results when given \tupe{exact}
arguments.  If one of these procedures is unable to deliver an \tupe{exact}
result when given \tupe{exact} arguments, then it may either report a
violation of an
implementation restriction or it may silently coerce its result to an
\tupe{inexact} number.  Such a coercion may cause an error\index{error} later.

%%R4%% I moved this stuff here.
% It seems to me that the only thing that this requires is that
% implementations that support inexact numbers have to have both
% exact and inexact representations for the integers 0 through 15.
% If that's what it's saying, I'd rather say it that way.
% On the other hand, letting the limit be as small as 15 sounds a
% tad silly, though I think I understand how that number was arrived at.
% (Or is 35 the number?)
%
%Implementations are encouraged, but not required, to support \tupe{inexact}
%numbers.  For any implementation that supports \tupe{inexact} numbers,
%there is a subset of the integers for which there are both \tupe{exact} and
%\tupe{inexact} representations.  This subset shall include all non-negative
%integers up to some limit specified by the implementation.  This limit
%shall be 16 or greater.  The
%\ide{exact\coerce{}inexact} and \ide{inexact\coerce{}exact}
%procedures implement the natural one-to-one correspondence between
%the \tupe{inexact} and \tupe{exact} integers within this range.

\vest An implementation may use floating point and other approximate 
representation strategies for \tupe{inexact} numbers.
%%R4%% The following sentence seemed a bit condescending as well as
% awkward.  It didn't seem to be very enforceable, so I flushed it.
%
%This is not to
%say that implementors need not use the best known algorithms for
%\tupe{inexact} computations---only that approximate methods of high
%quality are allowed.
%
This Standard recommends, but does not require, that the IEEE 32-bit
and 64-bit floating point standards be followed by implementations that use
flonum representations, and that implementations using
other representations should match or exceed the precision achievable
using these floating point standards~\cite{IEEE}.

\vest In particular, implementations that use flonum representations
shall follow these rules: A \tupe{flonum} result
shall be represented with at least as much precision as is used to express any of
the inexact arguments to that operation.
Potentially inexact operations such as \ide{sqrt}, when applied to \tupe{exact}
arguments, should produce \tupe{exact} answers whenever possible (for example the
square root of an \tupe{exact} 4 should be an \tupe{exact} 2).
If, however, an
\tupe{exact} number is operated upon so as to produce an \tupe{inexact} result
(as by \ide{sqrt}), and if the result is represented as a \tupe{flonum}, then
the most precise \tupe{flonum} format available shall be used; but if the result
is represented in some other way then the representation shall have at least as
much precision as the most precise \tupe{flonum} format available.

A particular implementation need not implement all of the procedures
defined in this section.  For example, an implementation in which all
numbers are \tupe{real} need not provide \ide{make-rectangular}.
However, all implementations shall provide the following procedures:

\begin{scheme}
number?      complex?      real?
rational?    integer?      exact?
inexact?     =             <
>            <=            >=
zero?        positive?     negative?
odd?         even?         max
min          +             *
-            abs           quotient
remainder    modulo        gcd
lcm          floor         ceiling
truncate     round         number->string
string->number%
\end{scheme}

Although Scheme allows a variety of written
%%R4%% representations of 
notations for
numbers, any particular implementation may support only some of them
(see the note associated with \ide{string->number}).
%%R4%%
For example, an implementation in which all numbers are \tupe{real}
need not support the rectangular and polar notations for complex
numbers.

If an implementation encounters an \tupe{exact} numerical constant that
it cannot represent as an \tupe{exact} number, then it may either report a
violation of an implementation restriction or it may silently represent the
constant by an \tupe{inexact} number.


\subsection{Syntax of numerical constants}
\label{numbernotations}

%@@@@LOSE@@@@

%%R4%%  I removed the following paragraph in an attempt to tighten up
% this subsection.  Except for its first sentence, which I moved to
% the subsection on implementation restrictions, I think its content
% is implied by the rest of the section.
%
%Although Scheme allows a variety of written representations of numbers,
%any particular implementation may support only some of them.
%These syntaxes are intended to be purely notational; any kind of number
%may be written in any form that the user deems convenient.  Of course,
%writing 1/7 as a limited-precision decimal fraction will not express the
%number exactly, but this approximate form of expression may be just what
%the user wants to see.

The syntax of the written representations for numbers is described
formally in~\ref{numbersyntax}.
%%R4%%  See~\ref{numberformats} for many examples.

A number may be written in binary, octal, decimal, or
hexadecimal by the use of a radix prefix.  The radix prefixes are {\cf
\#b}\sharpindex{b} (binary), {\cf \#o}\sharpindex{o} (octal), {\cf
\#d}\sharpindex{d} (decimal), and {\cf \#x}\sharpindex{x} (hexadecimal).  With
no radix prefix, a number is assumed to be expressed in decimal.

A
%%R4%%
% simple
numerical constant may be specified to be either \tupe{exact} or
\tupe{inexact} by a prefix.  The prefixes are {\cf \#e}\sharpindex{e}
for \tupe{exact}, and {\cf \#i}\sharpindex{i} for \tupe{inexact}.  An exactness
prefix may appear before or after any radix prefix that is used.  If
the written representation of a number has no exactness prefix, the
constant may be either \tupe{inexact} or \tupe{exact}.  It is
\tupe{inexact} if it contains a decimal point, an
exponent, or a ``\sharpsign'' character in the place of a digit,
otherwise it is \tupe{exact}.
%%R4%%  With our new syntax, the following sentence is redundant:
%
%The written representation of a
%compound number, such as a ratio or a complex, is exact if and only if
%all of its constituents are exact.

In systems with \tupe{inexact} numbers
of varying precisions it may be useful to specify
the precision of a constant.  For this purpose, numerical constants
may be written with an exponent marker that indicates the
desired precision of the \tupe{inexact}
representation.  The letters {\cf s}, {\cf f},
{\cf d}, and {\cf l}
specify the use of \var{short}, \var{single},
\var{double}, and \var{long} precision, respectively.  (When fewer
than four internal
%%R4%%\tupe{flonum}
\tupe{inexact}
representations exist, the four size
specifications are mapped onto those available.
\index{s (in number notation)}
\index{f (in number notation)}
\index{d (in number notation)}
\index{l (in number notation)}
For example, an
implementation with two internal representations may map short and
single together and long and double together.)  In addition, the
exponent marker {\cf e}
\index{e (in number notation)}
specifies the default precision for the
implementation.  The default precision has at least as much precision
as \var{double}, but
implementations may wish to allow this default to be set by the user.

\begin{scheme}
3.14159265358979F0
       {\rm Round to single ---} 3.141593
0.6L0
       {\rm Extend to long ---} .600000000000000%
\end{scheme}


\subsection{Numerical operations}

The reader is referred to~\ref{typeconventions} for a summary
of the naming conventions used to specify restrictions on the types of
arguments to numerical routines.
%%R4%% The following sentence has already been said twice, and the
% term "exactness-preserving" is no longer defined by the Report.
%
%  Remember that
%an exactness-preserving operation may coerce its result to inexact if the
%implementation is unable to represent it exactly.
The examples used in this section assume that any numerical constant written
using an \tupe{exact} notation is indeed represented as an \tupe{exact}
number.  Some examples also assume that certain numerical constants written
using an \tupe{inexact} notation can be represented without loss of
accuracy; the \tupe{inexact} constants were chosen so that this is
likely to be true in implementations that use flonums to represent
inexact numbers.

\todo{Scheme provides the usual set of operations for manipulating
numbers, etc.}

\begin{entry}{%
\proto{number?}{ obj}{procedure}
\proto{complex?}{ obj}{procedure}
\proto{real?}{ obj}{procedure}
\proto{rational?}{ obj}{procedure}
\proto{integer?}{ obj}{procedure}}

These numerical type predicates can be applied to any kind of
argument, including non-numbers.  They return \schtrue{} if the object is
of the named type, and otherwise they return \schfalse{}.
In general, if a type predicate is true of a number then all higher
type predicates are also true of that number.  Consequently, if a type
predicate is false of a number, then all lower type predicates are
also false of that number.
%%R4%% The new section on implementation restrictions subsumes: 
% Not every system
%supports all of these types; for example, it is entirely possible to have a
%Scheme system that has only \tupe{integer}s.  Nonetheless every implementation
%of Scheme shall have all of these predicates.

If \vr{z} is an inexact complex number, then {\cf (real? \vr{z})} is true if
and only if {\cf (zero? (imag-part \vr{z}))} is true.  If \vr{x} is an inexact
real number, then {\cf (integer? \vr{x})} is true if and only if
{\cf (= \vr{x} (round \vr{x}))}.

\begin{scheme}
(complex? 3+4i)         \ev  \schtrue
(complex? 3)            \ev  \schtrue
(real? 3)               \ev  \schtrue
(real? -2.5+0.0i)       \ev  \schtrue
(real? \#e1e10)          \ev  \schtrue
(rational? 6/10)        \ev  \schtrue
(rational? 6/3)         \ev  \schtrue
(integer? 3+0i)         \ev  \schtrue
(integer? 3.0)          \ev  \schtrue
(integer? 8/4)          \ev  \schtrue%
\end{scheme}

\begin{note}
The behavior of these type predicates on \tupe{inexact} numbers
is unreliable, since any inaccuracy may affect the result.
\end{note}

\begin{note}
In many implementations the \ide{rational?} procedure will be the same
as \ide{real?}, and the \ide{complex?} procedure will be the same as
\ide{number?}, but unusual implementations may be able to represent
some irrational numbers exactly or may extend the number system to
support some kind of non-complex numbers.
\end{note}

\end{entry}

\begin{entry}{%
\proto{exact?}{ \vr{z}}{procedure}
\proto{inexact?}{ \vr{z}}{procedure}}

These numerical predicates provide tests for the exactness of a
quantity.  For any Scheme number, precisely one of these predicates
is true.

\end{entry}


\begin{entry}{%
\proto{=}{ \vri{z} \vrii{z} \vriii{z} \dotsfoo}{procedure}
\proto{<}{ \vri{x} \vrii{x} \vriii{x} \dotsfoo}{procedure}
\proto{>}{ \vri{x} \vrii{x} \vriii{x} \dotsfoo}{procedure}
\proto{<=}{ \vri{x} \vrii{x} \vriii{x} \dotsfoo}{procedure}
\proto{>=}{ \vri{x} \vrii{x} \vriii{x} \dotsfoo}{procedure}}

%- Some implementations allow these procedures to take many arguments, to 
%- facilitate range checks.  
These procedures return \schtrue{} if their arguments are (respectively):
equal, monotonically increasing, monotonically decreasing,
monotonically nondecreasing, or monotonically nonincreasing.

These predicates are required to be transitive.

\begin{note}
The traditional implementations of these predicates in Lisp-like
languages are not transitive.
\end{note}

\begin{note}
While it is not an error to compare \tupe{inexact} numbers using these
predicates, the results may be unreliable because a small inaccuracy
may affect the result; this is especially true of \ide{=} and \ide{zero?}.
When in doubt, consult a numerical analyst.
\end{note}

\end{entry}

\begin{entry}{%
\proto{zero?}{ \vr{z}}{procedure}
\proto{positive?}{ \vr{x}}{procedure}
\proto{negative?}{ \vr{x}}{procedure}
\proto{odd?}{ \vr{n}}{procedure}
\proto{even?}{ \vr{n}}{procedure}}

These numerical predicates test a number for a particular property,
returning \schtrue{} or \schfalse.  See note above.

\end{entry}

\begin{entry}{%
\proto{max}{ \vri{x} \vrii{x} \dotsfoo}{procedure}
\proto{min}{ \vri{x} \vrii{x} \dotsfoo}{procedure}}

These procedures return the maximum or minimum of their arguments.

\begin{scheme}
(max 3 4)              \ev  4    ; exact
(max 3.9 4)            \ev  4.0  ; inexact%
\end{scheme}

\begin{note}
If any argument is inexact, then the result will also be inexact (unless
the procedure can prove that the inaccuracy is not large enough to affect the
result, which is possible only in unusual implementations).  If \ide{min} or
\ide{max} is used to compare numbers of mixed exactness, and the numerical
value of the result cannot be represented as an inexact number without loss of
accuracy, then the procedure may report a violation of an implementation
restriction.
\end{note}

\end{entry}


\begin{entry}{%
\proto{+}{ \vri{z} \dotsfoo}{procedure}
\proto{*}{ \vri{z} \dotsfoo}{procedure}}

These procedures return the sum or product of their arguments.  With
no arguments, \ide{+} returns {\cf 0} and \ide{*} returns {\cf 1}.
%- These procedures are exactness preserving.

\begin{scheme}
(+ 3 4)                 \ev  7
(+ 3)                   \ev  3
(+)                     \ev  0
(* 4)                   \ev  4
(*)                     \ev  1%
\end{scheme} 
 
\end{entry}


\begin{entry}{%
\proto{-}{ \vri{z} \vrii{z}}{procedure}
\protox{-}{ \vr{z}}{procedure}
\proto{/}{ \vri{z} \vrii{z}}{procedure}
\protox{/}{ \vr{z}}{procedure}}

With two arguments, these procedures return the difference or
quotient of their arguments.  With one argument,
however, they return the additive or multiplicative inverse of their argument.
%- These procedures are exactness preserving, except that division may
%- coerce its result to inexact in implementations that do not support
%- \tupe{ratnum}s. 

\begin{scheme}
(- 3 4)                 \ev  -1
(- 3)                   \ev  -3
(/ 3 4)                 \ev  3/4
(/ 3)                   \ev  1/3%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{abs}{ x}{procedure}}

\ide{Abs} returns the magnitude of its argument.  
%- \ide{Abs} is exactness preserving when its argument is real.
\begin{scheme}
(abs -7)                \ev  7
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{quotient}{ \vri{n} \vrii{n}}{procedure}
\proto{remainder}{ \vri{n} \vrii{n}}{procedure}
\proto{modulo}{ \vri{n} \vrii{n}}{procedure}}

These 
%- exactness-preserving 
procedures implement number-theoretic (integer)
division: For positive integers \vri{n} and \vrii{n}, if \vriii{n} and
\vriv{n} are integers such that $\vri{n}=\vrii{n}\vriii{n}+\vriv{n}$ and
$0\leq \vriv{n}<\vrii{n}$, then

\begin{scheme}
(quotient \vri{n} \vrii{n})        \ev  \vriii{n}
(remainder \vri{n} \vrii{n})       \ev  \vriv{n}
(modulo \vri{n} \vrii{n})          \ev  \vriv{n}%
\end{scheme}

For integers \vri{n} and \vrii{n} with \vrii{n} not equal to 0,

\begin{scheme}
(= \vri{n} (+ \=(* \vrii{n} (quotient \vri{n} \vrii{n}))
              \>(remainder \vri{n} \vrii{n})))
                                   \ev  \schtrue%
\end{scheme}

provided all numbers involved in that computation are exact.

The value returned by \ide{quotient} always has the sign of the product of
its arguments.  \ide{Remainder} and \ide{modulo} differ on negative
arguments---the
\ide{remainder} is either zero or has the sign of the dividend, the \ide{modulo}
always has the sign of the divisor:

\begin{scheme}
(modulo 13 4)           \ev  1
(remainder 13 4)        \ev  1

(modulo -13 4)          \ev  3
(remainder -13 4)       \ev  -1

(modulo 13 -4)          \ev  -3
(remainder 13 -4)       \ev  1

(modulo -13 -4)         \ev  -1
(remainder -13 -4)      \ev  -1

(remainder -13 -4.0)    \ev  -1.0  ; inexact%
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{gcd}{ \vri{n} \dotsfoo}{procedure}
\proto{lcm}{ \vri{n} \dotsfoo}{procedure}}

These procedures return the greatest common divisor or least common
multiple of their arguments.  The result is always non-negative.
With no arguments, \ide{gcd} returns {\cf 0} and \ide{lcm} returns
{\cf 1}.
%- These procedures are exactness preserving.

%%R4%% I added the inexact example.
\begin{scheme}
(gcd 32 -36)            \ev  4
(gcd)                   \ev  0
(lcm 32 -36)            \ev  288
(lcm 32.0 -36)          \ev  288.0  ; inexact
(lcm)                   \ev  1%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{numerator}{ \vr{q}}{procedure}
\proto{denominator}{ \vr{q}}{procedure}}

These procedures return the numerator or denominator of their
argument; the result is computed as if the argument was represented as
a fraction in lowest terms.  The denominator is always positive.  The
denominator of 0 is defined to be 1.
%- The remarks about denominators are new.
%- Clearly, they are exactness-preserving procedures.
\todo{More description and examples needed.}
\begin{scheme}
(numerator (/ 6 4))  \ev  3
(denominator (/ 6 4))  \ev  2
(denominator
  (exact->inexact (/ 6 4))) \ev 2.0%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{floor}{ x}{procedure}
\proto{ceiling}{ x}{procedure}
\proto{truncate}{ x}{procedure}
\proto{round}{ x}{procedure}
}

These procedures return integers.
\vest \ide{Floor} returns the largest integer not larger than \vr{x}.
\ide{Ceiling} returns the smallest integer not smaller than~\vr{x}.
\ide{Truncate} returns the integer closest to \vr{x} whose absolute
value is not larger than the absolute value of \vr{x}.  \ide{Round} returns the
closest integer to \vr{x}, rounding to even when \vr{x} is halfway between two
integers.

\begin{rationale}
\ide{Round} rounds to even for consistency with the rounding
modes required by the IEEE floating point standard.
\end{rationale}

\begin{note}
If the argument to one of these procedures is inexact, then the result
will also be inexact.  If an exact value is needed, the
result should be passed to the \ide{inexact->exact} procedure.
\end{note}

\begin{scheme}
(floor -4.3)          \ev  -5.0
(ceiling -4.3)        \ev  -4.0
(truncate -4.3)       \ev  -4.0
(round -4.3)          \ev  -4.0

(floor 3.5)           \ev  3.0
(ceiling 3.5)         \ev  4.0
(truncate 3.5)        \ev  3.0
(round 3.5)           \ev  4.0  ; inexact

(round 7/2)           \ev  4    ; exact
(round 7)             \ev  7%
\end{scheme}

\end{entry}

\begin{entry}{%
\proto{rationalize}{ x y}{procedure}
%- \proto{rationalize}{ x}{procedure}
}

\ide{Rationalize} returns the {\em simplest} rational number
differing from \vr{x} by no more than \vr{y}.  A rational number $r_1$ is
{\em simpler} \mainindex{simplest rational} than another rational number
$r_2$ if $r_1 = p_1/q_1$ and $r_2 = p_2/q_2$ (both in lowest terms) and $|p_1|
\leq |p_2|$ and $|q_1| \leq |q_2|$.  Thus $3/5$ is simpler than $4/7$.
Although not all rationals are comparable in this ordering (consider $2/7$
and $3/5$) any interval contains a rational number that is simpler than
every other rational number in that interval (the simpler $2/5$ lies
between $2/7$ and $3/5$).  Note that $0 = 0/1$ is the simplest rational of
all.

\begin{scheme}
(rationalize
  (inexact->exact .3) 1/10)  \ev 1/3    ; exact
(rationalize .3 1/10)        \ev \#i1/3  ; inexact%
\end{scheme}

\end{entry}

\begin{entry}{%
\proto{exp}{ \vr{z}}{procedure}
\proto{log}{ \vr{z}}{procedure}
\proto{sin}{ \vr{z}}{procedure}
\proto{cos}{ \vr{z}}{procedure}
\proto{tan}{ \vr{z}}{procedure}
\proto{asin}{ \vr{z}}{procedure}
\proto{acos}{ \vr{z}}{procedure}
\proto{atan}{ \vr{z}}{procedure}
\protox{atan}{ \vr{y} \vr{x}}{procedure}}

These procedures are part of every implementation that supports
%%R4%%
general
real numbers; they compute the usual transcendental functions.  \ide{Log}
computes the natural logarithm of \vr{z} (not the base ten logarithm).
\ide{Asin}, \ide{acos}, and \ide{atan} compute arcsine ($\sin^{-1}$),
arccosine ($\cos^{-1}$), and arctangent ($\tan^{-1}$), respectively.
The two-argument variant of \ide{atan} computes {\tt (angle
(make-rectangular \vr{x} \vr{y}))} (see below), even in implementations
that don't support general complex numbers.

In general, the mathematical functions log, arcsine, arccosine, and
arctangent are multiply defined.
For nonzero real $x$, the value of $\log x$ is defined to be
the one whose imaginary part lies in the range $-\pi$ (exclusive) to
$\pi$ (inclusive).  $\log 0$ is undefined.  The value of $\log z$ when
\vr{z} is complex is defined according to the formula
$$\log z = \log {\rm magnitude}(z) + i \: {\rm angle} (z)$$
With $\log$ defined this way, the values of $\sin^{-1} z$, $\cos^{-1} z$,
and $\tan^{-1} z$ are according to the following formul\ae:
$$\sin^{-1} z = -i \log (i z + \sqrt{1 - z^2})$$
$$\cos^{-1} z = \pi / 2 - \sin^{-1} z$$
$$\tan^{-1} z = (\log (1 + i z) - \log (1 - i z)) / (2 i)$$

The above specification follows~\cite{CLtL2}, which in turn
cites~\cite{Penfield81}; refer to these sources for more detailed
discussion of branch cuts, boundary conditions, and implementation of
these functions.  When it is possible these procedures produce a real
result from a real argument.

%%R4%%
\todo{The cited references are likely to change their branch cuts
soon to allow for the possibility of distinct positive and negative
zeroes, as in IEEE floating point.  We may not want to follow those
changes, since we may want a complex number with zero imaginary part
(whether positive or negative zero) to be treated as a real.  I don't
think there are any better standards for complex arithmetic than the
ones cited, so we're really on our own here.}

\end{entry}


\begin{entry}{%
\proto{sqrt}{ \vr{z}}{procedure}}

Returns the principal square root of \vr{z}.  The result will have
either positive real part, or zero real part and non-negative imaginary
part.
\end{entry}


\begin{entry}{%
\proto{expt}{ \vri{z} \vrii{z}}{procedure}}

Returns \vri{z} raised to the power \vrii{z}:
$${z_1}^{z_2} = e^{z_2 \log {z_1}}$$
$0^0$ is defined to be equal to 1.
\end{entry}

%- \begin{entry}{%- 
%- \proto{approximate}{ z x}{procedure}}
%- 
%- Returns an approximation to \vr{z} in a representation whose precision is
%- the same as that 
%- of the representation of \vr{x}, which must be an inexact number.  The
%- result is always inexact.
%- 
%- \begin{scheme}
%- (approximate 3.1415926535 1F10)
%-         \ev  3.14159F0
%- (approximate 3.1415926535 \#I65535)
%-         \ev \#I3
%- (approximate 3.14F0 1L8)
%-         \ev  3.14L0
%- (approximate 3.1415926535F0 1L8)
%-         \ev  3.14159L0
%- \end{scheme}
%- \end{entry}



\begin{entry}{%
\proto{make-rectangular}{ \vri{x} \vrii{x}}{procedure}
\proto{make-polar}{ \vriii{x} \vriv{x}}{procedure}
\proto{real-part}{ \vr{z}}{procedure}
\proto{imag-part}{ \vr{z}}{procedure}
\proto{magnitude}{ \vr{z}}{procedure}
\proto{angle}{ \vr{z}}{procedure}}

These procedures are part of every implementation that supports
%%R4%%
general
complex numbers.  Suppose \vri{x}, \vrii{x}, \vriii{x}, and \vriv{x} are
real numbers and \vr{z} is a complex number such that
 $$ \vr{z} = \vri{x} + \vrii{x}\hbox{$i$}
 = \vriii{x} \cdot e^{{\displaystyle{\hbox{$i$}} \vriv{x}}}$$
Then \ide{make-rectangular} and \ide{make-polar} return \vr{z},
\ide{real-part} returns \vri{x}, \ide{imag-part} returns \vrii{x},
\ide{magnitude} returns \vriii{x}, and \ide{angle} returns \vriv{x}.
In the case of \ide{angle}, whose value is not uniquely determined by
the preceding rule, the value returned will be the one in the range
$-\pi$ (exclusive) to $\pi$ (inclusive).

\begin{rationale}
\ide{Magnitude} is the same as \ide{abs} for a real argument,
but \ide{abs} shall be present in all implementations, whereas
\ide{magnitude} need only be present in implementations that support
general complex numbers.
\end{rationale}

\end{entry}


\begin{entry}{%
\proto{exact->inexact}{ \vr{z}}{procedure}
\proto{inexact->exact}{ \vr{z}}{procedure}}

\ide{Exact\coerce{}inexact} returns an \tupe{inexact} representation of \vr{z}.
The value returned is the
\tupe{inexact} number that is numerically closest to the argument.  
%%R4%%For
%\tupe{exact} arguments which have no reasonably close \tupe{inexact} equivalent,
%it is permissible to signal an error.
If an \tupe{exact} argument has no reasonably close \tupe{inexact} equivalent,
then a violation of an implementation restriction may be reported.

\ide{Inexact\coerce{}exact} returns an \tupe{exact} representation of
\vr{z}.  The value returned is the \tupe{exact} number that is numerically
closest to the argument.
%%R4%%  For \tupe{inexact} arguments which have no
%reasonably close \tupe{exact} equivalent, it is permissible to signal
%an error.
If an \tupe{inexact} argument has no reasonably close \tupe{exact} equivalent,
then a violation of an implementation restriction may be reported.

%%R%% I moved this to the section on implementation restrictions.
%For any implementation that supports \tupe{inexact} quantities,
%there is a subset of the integers for which there are both \tupe{exact} and
%\tupe{inexact} representations.  This subset shall include the non-negative
%integers up to a limit specified by the implementation.  The limit
%shall be big enough to represent all digits in reasonable radices, and
%may correspond to some natural word size for the implementation.  For
%such integers, these procedures implement the natural one-to-one
%correspondence between the representations.

These procedures implement the natural one-to-one correspondence between
\tupe{exact} and \tupe{inexact} integers throughout an
implementation-dependent range.  See~\ref{restrictions}.

\end{entry}

\medskip

\subsection{Numerical input and output}

\begin{entry}{%
\proto{number->string}{ number}{procedure}
\protox{number->string}{ number radix}{procedure}}

\vr{Radix} must be an exact integer, either 2, 8, 10, or 16.  If omitted,
\vr{radix} defaults to 10.
The procedure \ide{number\coerce{}string} takes a
number and a radix and returns as a string an external representation of
the given number in the given radix such that
\begin{scheme}
(let ((number \vr{number})
      (radix \vr{radix}))
  (eqv? number
        (string->number (number->string number
                                        radix)
                        radix)))
\end{scheme}
is true.  It is an error\index{error} if no possible result makes this expression true.

If \vr{number} is inexact, the radix is 10, and the above expression
can be satisfied by a result that contains a decimal point,
then the result contains a decimal point and is expressed using the
minimum number of digits (exclusive of exponent and trailing
zeroes) needed to make the above expression true~\cite{Heuristic};
otherwise the format of the result is unspecified\index{unspecified}.

The result returned by \ide{number\coerce{}string}
never contains an explicit radix prefix.

\begin{note}
The error case can occur only when \vr{number} is not a complex number
or is a complex number with a non-rational real or imaginary part.
\end{note}

\begin{rationale}
If \vr{number} is an inexact number represented using flonums, and
the radix is 10, then the above expression is normally satisfied by
a result containing a decimal point.  The unspecified case
allows for infinities, NaNs, and non-flonum representations.
\end{rationale}

\end{entry}


\begin{entry}{%
\proto{string->number}{ string}{procedure}
\protox{string->number}{ string radix}{procedure}}

%%R4%% I didn't include the (string->number string radix exactness)
% case, since I haven't heard any resolution of the coding to be used
% for the third argument.

Returns a number of the maximally precise representation expressed by the
given \vr{string}.  \vr{Radix} must be an exact integer, either 2, 8, 10,
or 16.  If supplied, \vr{radix} is a default radix that may be overridden
by an explicit radix prefix in \vr{string} (e.g. {\tt "\#o177"}).  If \vr{radix}
is not supplied, then the default radix is 10.  If \vr{string} is not
a syntactically valid notation for a number, then \ide{string\coerce{}number}
returns \schfalse{}.

\begin{scheme}
(string->number "100")        \ev  100
(string->number "100" 16)     \ev  256
(string->number "1e2")        \ev  100.0
(string->number "15\#\#")       \ev  1500.0%
\end{scheme}

\begin{note}
\label{stringtonumberdef}
An implementation may restrict the domain of
\ide{string->number} in the
following ways.  \ide{String->number} is permitted to return
\schfalse{} whenever \vr{string} contains an explicit radix prefix.
If all numbers supported by an implementation are real, then
\ide{string->number} is permitted to return \schfalse{} whenever
\vr{string} uses the polar or rectangular notations for complex
numbers.  If all numbers are integers, then
\ide{string->number} may return \schfalse{} whenever
the fractional notation is used.  If all numbers are exact, then
\ide{string->number} may return \schfalse{} whenever
an exponent marker or explicit exactness prefix is used, or if
a {\tt \#} appears in place of a digit.  If all inexact
numbers are integers, then
\ide{string->number} may return \schfalse{} whenever
a decimal point is used.
\end{note}

\end{entry}

\section{Characters}
\label{charactersection}

Characters are objects that represent printed characters such as
letters and digits.  
%There is no requirement that the data type of
%characters be disjoint from other data types; implementations are
%encouraged to have a separate character data type, but may choose to
%represent characters as integers, strings, or some other type.
Characters are written using the notation \sharpsign\backwhack\hyper{character}
or \sharpsign\backwhack\hyper{character name}.
For example:

$$
\begin{tabular}{ll}
{\tt \#\backwhack{}a}&; lower case letter\\
{\tt \#\backwhack{}A}&; upper case letter\\
{\tt \#\backwhack{}(}&; left parenthesis\\
{\tt \#\backwhack{}\visiblespace}&; the space character\\
{\tt \#\backwhack{}space}&; the preferred way to write a space\\
{\tt \#\backwhack{}newline}&; the newline character\\
\end{tabular}
$$

Case is significant in \sharpsign\backwhack\hyper{character}, but not in
\sharpsign\backwhack\hyper{character name}.  If \hyper{character} in
\sharpsign\backwhack\hyper{character} is alphabetic, then the character
following \hyper{character} must be a delimiter character such as a
space or parenthesis.  This rule resolves the ambiguous case where, for
example, the sequence of characters ``{\tt\sharpsign\backwhack space}''
could be taken to be either a representation of the space character or a
representation of the character ``{\tt\sharpsign\backwhack s}'' followed
by a representation of the symbol ``{\tt pace}.''

\todo{Fix}
Characters written in the \sharpsign\backwhack{} notation are self-evaluating.
That is, they do not have to be quoted in programs.

\vest Some of the procedures that operate on characters ignore the
difference between upper case and lower case.  The procedures that
ignore case have \hbox{``{\tt -ci}''} (for ``case
insensitive'') embedded in their names.


\begin{entry}{%
\proto{char?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is a character, otherwise returns \schfalse.

\end{entry}


\begin{entry}{%
\proto{char=?}{ \vari{char} \varii{char}}{procedure}
\proto{char<?}{ \vari{char} \varii{char}}{procedure}
\proto{char>?}{ \vari{char} \varii{char}}{procedure}
\proto{char<=?}{ \vari{char} \varii{char}}{procedure}
\proto{char>=?}{ \vari{char} \varii{char}}{procedure}}

\label{characterequality}
\nodomain{Both \vari{char} and \varii{char} must be characters.}
These procedures impose a total ordering on the set of characters.  It
is guaranteed that under this ordering:

\begin{itemize}
\item The upper case characters are in order.  For example,
\begin{scheme}
(char<?\ \#\backwhack{}A \#\backwhack{}B) \evalsto \schtrue%
\end{scheme}
\item The lower case characters are in order.  For example,
\begin{scheme}
(char<?\ \#\backwhack{}a \#\backwhack{}b) \evalsto \schtrue%
\end{scheme}
\item The digits are in order.  For example,
\begin{scheme}
(char<?\ \#\backwhack{}0 \#\backwhack{}9) \evalsto \schtrue%
\end{scheme}
\item Either all the digits precede all the upper case letters, or vice versa.
\item Either all the digits precede all the lower case letters, or vice versa.
\end{itemize}

\end{entry}


\begin{entry}{%
\proto{char-ci=?}{ \vari{char} \varii{char}}{procedure}
\proto{char-ci<?}{ \vari{char} \varii{char}}{procedure}
\proto{char-ci>?}{ \vari{char} \varii{char}}{procedure}
\proto{char-ci<=?}{ \vari{char} \varii{char}}{procedure}
\proto{char-ci>=?}{ \vari{char} \varii{char}}{procedure}}

\nodomain{Both \vari{char} and \varii{char} must be characters.}
These procedures are similar to \ide{char=?}\ et cetera, but they treat
upper case and lower case letters as the same.  For example,

\begin{scheme}
(char-ci=?\ \#\backwhack{}A \#\backwhack{}a) \evalsto \schtrue%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{char-alphabetic?}{ char}{procedure}
\proto{char-numeric?}{ char}{procedure}
\proto{char-whitespace?}{ char}{procedure}
\proto{char-upper-case?}{ char}{procedure}
\proto{char-lower-case?}{ char}{procedure}}

These procedures return \schtrue{} if their arguments are alphabetic,
numeric, whitespace, upper case, or lower case characters, respectively,
otherwise they return \schfalse.  The following remarks, which are specific to
the ASCII character set, are intended only as a guide:  The alphabetic characters
are the 52 upper and lower case letters.  The numeric characters are the
ten decimal digits.  The whitespace characters are space, tab, line
feed, form feed, and carriage return.
\end{entry}


%%IEEE%% Deleted because we've decided to do what R4RS does here.
% \begin{entry}{%
% \proto{char->integer}{ char}{procedure}}
% 
% \vest Given a character, \ide{char\coerce{}integer} returns an exact
% nonnegative integer representation of the character.  This procedure
% implements a one-to-one map between the set of characters and some
% subset of the integers.  That is,
% 
% \begin{scheme}
% (eqv? (char=? \vr{a} \vr{b})
%       (= (char\coerce{}integer \vr{a})
%          (char\coerce{}integer \vr{b})))       \evalsto \schtrue
% \end{scheme}
% 
% \begin{note}
% The integer returned by \ide{char\coerce{}integer} is not guaranteed
% to be the ``standard'' character code for any particular character
% set; the choice of what number to return is left to the implementor.
% In particular, on operating systems that normally use the ASCII
% character set, the value of \ide{char\coerce{}integer} is not
% constrained to be the ASCII code for that character.
% \end{note}
% 
% \end{entry}

\begin{entry}{%
\proto{char->integer}{ char}{procedure}
\proto{integer->char}{ \vr{n}}{procedure}}

Given a character, \ide{char\coerce{}integer} returns an exact integer
representation of the character.  Given an exact integer that is the image of
a character under \ide{char\coerce{}integer}, \ide{integer\coerce{}char}
returns that character.  These procedures implement injective order isomorphisms
between the set of characters under the \ide{char<=?}\ ordering and some
subset of the integers under the \ide{<=}\ ordering.  That is, if

\begin{scheme}
(char<=? \vr{a} \vr{b}) \evalsto \schtrue  {\rm{}and}  %
(<= \vr{x} \vr{y}) \evalsto \schtrue%
\end{scheme}

\noindent and \vr{x} and \vr{y} are in the domain of
\ide{integer\coerce{}char}, then

\begin{scheme}
(<= (char\coerce{}integer \vr{a})
    (char\coerce{}integer \vr{b}))         \ev  \schtrue

(char<=? (integer\coerce{}char \vr{x})
         (integer\coerce{}char \vr{y}))     \ev  \schtrue%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{char-upcase}{ char}{procedure}
\proto{char-downcase}{ char}{procedure}}

\nodomain{\var{Char} must be a character.}
These procedures return a character \varii{char} such that {\cf
(char-ci=? \var{char} \varii{char})}.  In addition, if \var{char} is
alphabetic, then the result of \ide{char-upcase} is upper case and the
result of \ide{char-downcase} is lower case.

\end{entry}


\section{Strings}
\label{stringsection}

Strings are sequences of characters.  
%In some implementations of Scheme
%they are immutable; other implementations provide destructive procedures
%such as \ide{string-set!}\ that alter string objects.
\vest Strings are written as sequences of characters enclosed within doublequotes
({\cf "}).  A doublequote can be written inside a string only by escaping
it with a backslash (\backwhack{}), as in

\begin{scheme}
"The word \backwhack{}"recursion\backwhack{}" has many meanings."%
\end{scheme}

A backslash can be written inside a string only by escaping it with another
backslash.  This standard does not specify the effect of a backslash within a
string that is not followed by a doublequote or backslash.

\vest A string may continue from one line to the next, but in such a
situation the exact contents of the string are unspecified\index{unspecified}.

The \mainindex{length of string}{\em length} of a string is the number
of characters that it contains.  This number is a non-negative integer
that is fixed when the string is created.  The \defining{valid
indexes} of a string are the exact non-negative integers less than the
length of the string.  The first character of a string has index 0,
the second has index 1, and so on.

\vest Some of the procedures that operate on strings ignore the
difference between upper and lower case.  The versions that ignore case
have \hbox{``{\cf -ci}''} (for ``case insensitive'') embedded in their
names.


\begin{entry}{%
\proto{string?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is a string, otherwise returns \schfalse.
\end{entry}


\begin{entry}{%
\proto{make-string}{ \vr{k}}{procedure}
\protox{make-string}{ \vr{k} char}{procedure}}

\ide{Make-string} returns a newly allocated string of
length \vr{k}.  If \var{char} is given, then all elements of the string
are initialized to \var{char}, otherwise the contents of the
\var{string} are unspecified\index{unspecified}.

\end{entry}


\begin{entry}{%
\proto{string}{ \var{char} \dotsfoo}{procedure}}

Returns a newly allocated string composed of the arguments.

\end{entry}

\begin{entry}{%
\proto{string-length}{ string}{procedure}}

Returns the number of characters in the given \var{string}.
\end{entry}


\begin{entry}{%
\proto{string-ref}{ string \vr{k}}{procedure}}

\domain{\vr{k} must be a valid index of \var{string}.}
\ide{String-ref} returns character \vr{k} of \var{string} using zero-origin indexing.
\end{entry}


\begin{entry}{%
\proto{string-set!}{ string k char}{procedure}}

\domain{%\var{String} must be a string, 
\vr{k} must be a valid index of \var{string}%, and \var{char} must be a character
.}
\ide{String-set!} stores \var{char} in element \vr{k} of \var{string}
and returns an unspecified\index{unspecified} value.  % <!>

\begin{scheme}
(define (f) (make-string 3 \sharpsign\backwhack{}*))
(define (g) "***")
(string-set! (f) 0 \sharpsign\backwhack{}?)  \ev  \unspecified
(string-set! (g) 0 \sharpsign\backwhack{}?)  \lev  \scherror ; constant string
(string-set! (symbol->string 'immutable)
             0
             \sharpsign\backwhack{}?)  \ev  \scherror%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{string=?}{ \vari{string} \varii{string}}{procedure}
\proto{string-ci=?}{ \vari{string} \varii{string}}{procedure}}

Returns \schtrue{} if the two strings are the same length and contain the same
characters in the same positions, otherwise returns \schfalse.
\ide{String-ci=?}\ treats
upper and lower case letters as though they were the same character, but
\ide{string=?}\ treats upper and lower case as distinct characters.

\end{entry}


\begin{entry}{%
\proto{string<?}{ \vari{string} \varii{string}}{procedure}
\proto{string>?}{ \vari{string} \varii{string}}{procedure}
\proto{string<=?}{ \vari{string} \varii{string}}{procedure}
\proto{string>=?}{ \vari{string} \varii{string}}{procedure}
\proto{string-ci<?}{ \vari{string} \varii{string}}{procedure}
\proto{string-ci>?}{ \vari{string} \varii{string}}{procedure}
\proto{string-ci<=?}{ \vari{string} \varii{string}}{procedure}
\proto{string-ci>=?}{ \vari{string} \varii{string}}{procedure}}

These procedures are the lexicographic extensions to strings of the
corresponding orderings on characters.  For example, \ide{string<?}\ is
the lexicographic ordering on strings induced by the ordering
\ide{char<?}\ on characters.  If two strings differ in length but
are the same up to the length of the shorter string, the shorter string
is considered to be lexicographically less than the longer string.

\end{entry}


\begin{entry}{%
\proto{substring}{ string start end}{procedure}}

\domain{\var{String} must be a string, and \var{start} and \var{end}
must be exact integers satisfying
$$0 \leq \var{start} \leq \var{end} \leq \hbox{\tt(string-length \var{string})\rm.}$$}
\ide{Substring} returns a newly allocated string formed from the characters of
\var{string} beginning with index \var{start} (inclusive) and ending with index
\var{end} (exclusive).
\end{entry}


\begin{entry}{%
\proto{string-append}{ \var{string} \dotsfoo}{procedure}}

Returns a newly allocated string whose characters form the concatenation of the
given strings.

\end{entry}


\section{Vectors}
\label{vectorsection}

Vectors are heterogenous structures whose elements are indexed
by integers.  A vector typically occupies less space than a list
of the same length, and the average time required to access a randomly
chosen element is typically less for the vector than for the list.

The \mainindex{length of vector}{\em length} of a vector is the number
of elements that it contains.  This number is a non-negative integer
that is fixed when the vector is created.  The {\em valid
indexes}\index{valid indexes} of a vector are the exact non-negative
integers less than the length of the vector.  The first element in a
vector is indexed by zero, and the last element is indexed by one less
than the length of the vector.

Vectors are written using the notation {\tt\#(\var{obj} \dotsfoo)}.
For example, a vector of length 3 containing the number zero in element
0, the list {\cf(2 2 2 2)} in element 1, and the string {\cf "Anna"} in
element 2 can be written as following:

\begin{scheme}
\#(0 (2 2 2 2) "Anna")%
\end{scheme}

Note that this is the external representation of a vector, not an
expression evaluating to a vector.  Like list constants, vector
constants must be quoted:

\begin{scheme}
'\#(0 (2 2 2 2) "Anna")  \lev  \#(0 (2 2 2 2) "Anna")%
\end{scheme}

\todo{Pitman sez: The visual similarity to lists is bound to be confusing
to some.  Elaborate on the distinction.}


\begin{entry}{%
\proto{vector?}{ obj}{procedure}}
 
Returns \schtrue{} if \var{obj} is a vector, otherwise returns \schfalse.
\end{entry}


\begin{entry}{%
\proto{make-vector}{ k}{procedure}
\protox{make-vector}{ k fill}{procedure}}

Returns a newly allocated vector of \var{k} elements.  If a second
argument is given, then each element is initialized to \var{fill}.
Otherwise the initial contents of each element is unspecified\index{unspecified}.

\end{entry}


\begin{entry}{%
\proto{vector}{ obj \dotsfoo}{procedure}}

Returns a newly allocated vector whose elements contain the given
arguments.  Analogous to \ide{list}.

\begin{scheme}
(vector 'a 'b 'c)               \ev  \#(a b c)%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{vector-length}{ vector}{procedure}}

Returns the number of elements in \var{vector}.
\end{entry}


\begin{entry}{%
\proto{vector-ref}{ vector k}{procedure}}

\domain{\vr{k} must be a valid index of \var{vector}.}
\ide{Vector-ref} returns the contents of element \vr{k} of
\var{vector}.

\begin{scheme}
(vector-ref '\#(1 1 2 3 5 8 13 21)
            5)  \lev  8
(vector-ref '\#(1 1 2 3 5 8 13 21)
            (inexact->exact
              (round (* 2 (acos -1))))) \lev 13%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{vector-set!}{ vector k obj}{procedure}}

\domain{\vr{k} must be a valid index of \var{vector}.}
\ide{Vector-set!} stores \var{obj} in element \vr{k} of \var{vector}.
The value returned by \ide{vector-set!}\ is unspecified\index{unspecified}.  % <!>

\begin{scheme}
(let ((vec (vector 0 '(2 2 2 2) "Anna")))
  (vector-set! vec 1 '("Sue" "Sue"))
  vec)      \lev  \#(0 ("Sue" "Sue") "Anna")

(vector-set! '\#(0 1 2) 1 "doe")  \lev  \scherror  ; constant vector%
\end{scheme}
\end{entry}


\section{Control features}
\label{proceduresection}
 
% Intro flushed; not very a propos any more.
% Procedures should be discussed somewhere, however.

%This chapter describes various primitive procedures which control the
%flow of program execution in special ways.
%The \ide{procedure?}\ predicate is also described here.

\todo{{\tt Procedure?} doesn't belong in a section with the name
``control features.''  What to do?}

\begin{entry}{%
\proto{procedure?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is a procedure, otherwise returns \schfalse.

\begin{scheme}
(procedure? car)            \ev  \schtrue
(procedure? 'car)           \ev  \schfalse
(procedure? (lambda (x) (* x x)))   
                            \ev  \schtrue
(procedure? '(lambda (x) (* x x)))  
                            \ev  \schfalse
(call-with-current-continuation procedure?)
                            \ev  \schtrue%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{apply}{ proc \vari{obj} $\ldots$ args}{procedure}}

\domain{\var{Proc} must be a procedure and \var{args} must be a list.}
Calls \var{proc} with the elements of the list

\begin{scheme}
(append (list \vari{obj} \dotsfoo) \var{args})
\end{scheme}

as the actual arguments.

\begin{scheme}
(apply + (list 3 4))              \ev  7

(apply + 10 (list 3 4))           \ev  17

(define compose
  (lambda (f g)
    (lambda args
      (f (apply g args)))))

((compose sqrt *) 12 75)              \ev  30%
\end{scheme}
\end{entry}


\begin{entry}{%
\proto{map}{ proc \vari{list} \varii{list} \dotsfoo}{procedure}}

\domain{The \var{list}s must be lists, and \var{proc} must be a
procedure taking as many arguments as there are {\it list}s.  If more
than one \var{list} is given, then they must all be the same length.}
\ide{Map} applies \var{proc} element-wise to the elements of the
\var{list}s and returns a list of the results, in order from left to right.
The dynamic order in which \var{proc} is applied to the elements of the
\var{list}s is unspecified\index{unspecified}.

\begin{scheme}
(map cadr '((a b) (d e) (g h)))   \lev  (b e h)

(map (lambda (n) (expt n n))
     '(1 2 3 4 5))                \lev  (1 4 27 256 3125)

(map + '(1 2 3) '(4 5 6))         \ev  (5 7 9)

(let ((count 0))
  (map (lambda (ignored)
         (set! count (+ count 1))
         count)
       '(a b c)))                 \ev  \unspecified%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{for-each}{ proc \vari{list} \varii{list} \dotsfoo}{procedure}}

The arguments to \ide{for-each} are like the arguments to \ide{map}, but
\ide{for-each} calls \var{proc} for its side effects rather than for its
values.  Unlike \ide{map}, \ide{for-each} is guaranteed to call \var{proc} on
the elements of the \var{list}s in order from the first element to the
last, and the value returned by \ide{for-each} is unspecified\index{unspecified}.

\begin{scheme}
(let ((v (make-vector 5)))
  (for-each (lambda (i)
              (vector-set! v i (* i i)))
            '(0 1 2 3 4))
  v)                                \ev  \#(0 1 4 9 16)%
\end{scheme}

\end{entry}


\begin{entry}{%
\proto{call-with-current-continuation}{ proc}{procedure}}

\label{continuations}
\domain{\var{Proc} must be a procedure of one argument.} The procedure
\callcc{} packages up the current continuation (see the rationale
below) as an ``escape procedure''\mainindex{escape procedure} and
passes it as an argument to \var{proc}.  The escape procedure is a
Scheme procedure of one argument that, if it is later passed a value,
will ignore whatever continuation is in effect at that later time and
will give the value instead to the continuation that was in effect
when the escape procedure was created.

\vest The escape procedure that is passed to \var{proc} has
unlimited extent just like any other procedure in Scheme.  It may be stored
in variables or data structures and may be called as many times as desired.

\vest The following examples show only the most common uses of
\callcc{}.  If all
real programs were as simple as these examples, there would be no need
for a procedure with the power of
\callcc{}.

\begin{scheme}
(call-with-current-continuation
  (lambda (exit)
    (for-each (lambda (x)
                (if (negative? x)
                    (exit x)))
              '(54 0 37 -3 245 19))
    \schtrue))                        \ev  -3

(define (list-length obj)
  (call-with-current-continuation
    (lambda (return)
      (let r ((obj obj))
        (cond ((null? obj) 0)
              ((pair? obj) (+ 1 (r (cdr obj))))
              (else (return \schfalse)))))))

(list-length '(1 2 3 4))            \ev  4

(list-length '(a b . c))            \ev  \schfalse%
\end{scheme}

\begin{rationale}

A common use of \callcc{} is for
structured, non-local exits from loops or procedure bodies, but in fact
\callcc{} is extremely useful for implementing a
wide variety of advanced control structures.

\todo{The example in the following paragraph assumes a top level
read-eval-print loop.  This is probably a bad idea as it precludes a
non-interpretive Scheme system.  Create a new example.}

\vest Whenever a Scheme expression is evaluated there is a
\defining{continuation} wanting the result of the expression.  The continuation
represents an entire (default) future for the computation.  If the expression is
evaluated at the top level of a Scheme interpreter, for example, then the continuation will take the
result, print it on the screen, prompt for the next input, evaluate it, and
so on forever.  Most of the time the continuation includes actions
specified by user code, as in a continuation that will take the result,
multiply it by the value stored in a local variable, add seven, and give
the answer to the top level continuation to be printed.  Normally these
ubiquitous continuations are hidden behind the scenes and programmers don't
think much about them.  On rare occasions, however, a programmer may
need to deal with continuations explicitly.
\callcc{} allows Scheme programmers to do
that by creating a procedure that acts just like the current
continuation.

% \vest Most programming languages incorporate one or more special-purpose
% escape constructs with names like {\tt exit}, \hbox{\ide{return}}, or
% even {\tt goto}.  In 1965, however, Peter Landin~\cite{Landin65}
% invented a general purpose escape operator called the J-operator.  John
% Reynolds~\cite{Reynolds72} described a simpler but equally powerful
% construct in 1972.  The \ide{catch} special form described by Sussman
% and Steele in the 1975 report on Scheme is exactly the same as
% Reynolds's construct, though its name came from a less general construct
% in MacLisp.  Several Scheme implementors noticed that the full power of the
% \ide{catch} construct could be provided by a procedure instead of by a
% special syntactic construct, and the name
% \callcc{} was coined in 1982.  This name is
% descriptive, but opinions differ on the merits of such a long name, and
% some people use the name \ide{call/cc} instead.

\end{rationale}

\end{entry}


\section{Input and output}

\subsection{Ports}
\label{portsection}

Ports represent input and output devices.  To Scheme, an input port is a
Scheme object that can deliver characters upon command, while an output port
is a Scheme object that can accept characters. \mainindex{port}

\todo{Haase: Mention that there are alternatives to files?}

\begin{entry}{%
\proto{call-with-input-file}{ string proc}{procedure}
\proto{call-with-output-file}{ string proc}{procedure}}

\domain{\var{Proc} must be a procedure of one argument, and
\var{string} must be a string naming a file.  For
\ide{call-with-input-file}, the file must already exist; for
\ide{call-with-output-file}, the effect is unspecified\index{unspecified} if the file
already exists.} These procedures call \var{proc} with one argument: the
port obtained by opening the named file for input or output.  If the
file cannot be opened, an error\index{error} is signalled.  If the procedure returns,
then the port is closed automatically and the value yielded by the
procedure is returned.  If the procedure does not return, then 
the port will not be closed automatically unless it is possible to
prove that the port will never again be used for a read or write
operation.

\begin{rationale}
Because Scheme's escape procedures have unlimited extent, it is
possible to escape from the current continuation but later to escape
back in.  If implementations were permitted to close the port on any
escape from the current continuation, then it would be impossible to
write portable code using both
\callcc{} and
\ide{call-with-input-file} or \ide{call-with-output-file}.
\todo{Pitman wants more said here; maybe encourage users to call
{\tt close-foo-port}; maybe talk about process switches (?).}
\end{rationale} 
\end{entry}


\begin{entry}{%
\proto{input-port?}{ obj}{procedure}
\proto{output-port?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is an input port or output port
respectively, otherwise returns \schfalse.

\todo{Won't necessarily return true after port is closed.}

\end{entry}


\begin{entry}{%
\proto{current-input-port}{}{procedure}
\proto{current-output-port}{}{procedure}}
 
Returns the current default input or output port.

\end{entry}


\begin{entry}{%
\proto{open-input-file}{ filename}{procedure}}
 
Takes a string naming an existing file and returns an input port capable of
delivering characters from the file.  If the file cannot be opened, an error\index{error} is
signalled.

\end{entry}


\begin{entry}{%
\proto{open-output-file}{ filename}{procedure}}

Takes a string naming an output file to be created and returns an output
port capable of writing characters to a new file by that name.  If the file
cannot be opened, an error\index{error} is signalled.  If a file with the given name
already exists, the effect is unspecified\index{unspecified}.

\end{entry}


\begin{entry}{%
\proto{close-input-port}{ port}{procedure}
\proto{close-output-port}{ port}{procedure}}

Closes the file associated with \var{port}, rendering the \var{port}
incapable of delivering or accepting characters.  \todo{But maybe a no-op
on some ports, e.g. terminals or editor buffers.}
These routines have no effect if the file has already been closed.
The value returned is unspecified\index{unspecified}.

\todo{Ramsdell:  Some note is needed explaining why there are two
different close procedures.}

\todo{A port isn't necessarily still a port after it has been closed?}

\end{entry}


\subsection{Input}
\label{inputsection}

\noindent \hbox{ }  %???
\vspace{-5ex}
\todo{The input routines have some things in common, maybe explain here.}

\begin{entry}{%
\proto{eof-object?}{ obj}{procedure}}

Returns \schtrue{} if \var{obj} is an end of file object, otherwise returns
\schfalse.  The precise set of end of file objects will vary among
implementations, but in any case no end of file object will ever be an object
that can be read in using \ide{read}.

\end{entry}

\begin{entry}{%
\proto{read}{}{procedure}
\protox{read}{ port}{procedure}}

\ide{Read} converts external representations of Scheme objects into the
objects themselves.  That is, it is a parser for the nonterminal
\meta{datum} (see~\ref{datum} and \ref{listsection}).  However,
\ide{read} need not implement the full grammar for the nonterminal
\meta{number}: it may report the violation of an implementation
restriction instead.  See the note associated with
\ide{string->number} on page~\pageref{stringtonumberdef} for
permissible restrictions of numeric syntax.

\ide{Read} returns the next object parsable from the given input
\var{port}, updating \var{port} to point to the first character past
the end of the external representation of the object.  If an end of
file is encountered in the input before any characters are found that
can begin an object, then an end of file object is returned.  \todo{}
The port remains open, and further attempts to read will also return
an end of file object.  If an end of file is encountered after the
beginning of an object's external representation, but the external
representation is incomplete and therefore not parsable, an error\index{error} is
signalled.

The \var{port} argument may be omitted, in which case it defaults to the
value returned by \ide{current-input-port}.  It is an error\index{error} to read from
a closed port.
\end{entry}

\begin{entry}{%
\proto{read-char}{}{procedure}
\protox{read-char}{ port}{procedure}}

Returns the next character available from the input \var{port}, updating
the \var{port} to point to the following character.  If no more characters
are available, an end of file object is returned.  \var{Port} may be
omitted, in which case it defaults to the value returned by \ide{current-input-port}.

\end{entry}


\begin{entry}{%
\proto{peek-char}{}{procedure}
\protox{peek-char}{ port}{procedure}}

Returns the next character available from the input \var{port},
{\em without} updating
the \var{port} to point to the following character.  If no more characters
are available, an end of file object is returned.  \var{Port} may be
omitted, in which case it defaults to the value returned by \ide{current-input-port}.

\begin{note}
The value returned by a call to \ide{peek-char} is the same as the
value that would have been returned by a call to \ide{read-char} with the
same \var{port}.  The only difference is that the very next call to
\ide{read-char} or \ide{peek-char} on that \var{port} will return the
value returned by the preceding call to \ide{peek-char}.  In particular, a call
to \ide{peek-char} on an interactive port will hang waiting for input
whenever a call to \ide{read-char} would have hung.
\end{note}

\end{entry}


\subsection{Output}
\label{outputsection}

% We've got to put something here to fix the indentation!!
\noindent \hbox{}
\vspace{-5ex}

\begin{entry}{%
\proto{write}{ obj}{procedure}
\protox{write}{ obj port}{procedure}}

Writes a written representation of \var{obj} to the given \var{port}.
If \var{obj} has a standard external representation, then the written
representation generated by \ide{write} shall be parsable by
\ide{read} into an equivalent object; but see~\ref{symbolsection} for
an exception.  Thus strings that appear in the written representation
are enclosed in doublequotes, and within those strings backslash and
doublequote characters are escaped by backslashes.

\ide{Write} returns an unspecified\index{unspecified} value.  The \var{port} argument may
be omitted, in which case it defaults to the value returned by
\ide{current-output-port}.

\end{entry}


\begin{entry}{%
\proto{display}{ obj}{procedure}
\protox{display}{ obj port}{procedure}}

Writes a representation of \var{obj} to the given \var{port}.  Strings
that appear in the written representation are not enclosed in
doublequotes, and no characters are escaped within those strings.  Character
objects appear in the representation as if written by \ide{write-char}
instead of by \ide{write}.  \ide{Display} returns an unspecified\index{unspecified} value.
The \var{port} argument may be omitted, in which case it defaults to the
value returned by \ide{current-output-port}.

\begin{rationale}
\ide{Write} is intended
for producing mach\-ine-readable output and \ide{display} is for producing
human-readable output.
\end{rationale}
\end{entry}


\begin{entry}{%
\proto{newline}{}{procedure}
\protox{newline}{ port}{procedure}}

Writes an end of line to \var{port}.  Exactly how this is done differs
from one operating system to another.  Returns an unspecified\index{unspecified} value.
The \var{port} argument may be omitted, in which case it defaults to the
value returned by \ide{current-output-port}.

\end{entry}


\begin{entry}{%
\proto{write-char}{ char}{procedure}
\protox{write-char}{ char port}{procedure}}

Writes the character \var{char} (not an external representation of the
character) to the given \var{port} and returns an unspecified\index{unspecified} value.  The
\var{port} argument may be omitted, in which case it defaults to the value
returned by \ide{current-output-port}.

\end{entry}
