\chapter{Arrays}\label{XtenArrays}\index{arrays}

An array is a mapping from a distribution to a range data
type. Multiple arrays may be declared with the same underlying
distribution. 

Each array has a field {\tt a.distribution} which may be used to
obtain the underlying distribution.\index{arrays!distribution@\tt distribution}

The distribution underlying an array {\tt a} may be obtained 

\input{XtenRegions}
\input{XtenDistributions}
\section{Array initializer}\label{ArrayInitializer}\label{array!creation}
An array initializer creating a new array with distribution {\cf D} may
optionally take a parametrized block of the form {\cf (ind1,...,
indk)\{S\}}. Here, {\cf k} may be zero; in this case the statement is written
as just {\cf \{S\}}. For instance:
\begin{x10}
\_ data = new int value [1000] 
    (i)\{ return i*i; \};
\_ data2 = new int[1000->current]@threadlocal 
    \{ return 1; \};
\end{x10}

{}\noindent The first declaration stores in {\tt data} an (immutable)
array whose distribution is {\tt (1..1000)-> here}, which is created
{\tt here}, and which is initialized with the value {\tt i*i} at index
{\tt i}. 

The second declaration stores in {\tt data2} a reference to a mutable
array (allocated in the {\tt threadlocal} region of the current
activity) with {\tt 1000} elements each of which is located in the
same place as the array (hence is {\tt threadlocal}). Each array
component is initialized to {\tt 1}.


In general the expression
\begin{x10}
    \_ data =  new T[D]@P (ind1, ..., indk) \{ S \}
\end{x10}

\noindent should be thought of as creating a new array located at {\tt P} with a {\tt k}-dimensional distribution {\tt D} such that the elements of 
the array are initialized as if by execution of the code:

\begin{x10}
   ateach(ind1, ..., indk : D) \{
       A[ind1, ..., indk] = 
     (new Object \{ T val(D ind1,...,indk) \{S\}\})
     .val(ind1, ..., indk);
   \}
\end{x10}

Notice that in the method declaration {\cf D} is used as a type. 

Other examples:
\begin{x10}
\_ data = new int[1000](i)\{return i*i; \};
float[D] d = new float[D] (i)\{return 10.0*i; \}; 
float[D] d2 = new float[D] (i)\{return i*i; \};
float[D] result = new float[D] 
      (i) \{return d[i] + d2[i]; \};
\end{x10}

\section{Operations on arrays}
In the following let {\tt a} be an array with distribution {\tt D} and
base type {\tt T}. {\tt a} may be mutable or immutable, unless
indicated otherwise.

\subsection{Element operations}\index{array!access}
The value of {\tt a} at a point {\tt p} in its region of definition is
obtained by using the indexing operation {\tt a[p]}. This operation
may be used on the left hand side of an assignment operation to update
the value.

\subsection{Constant promotion}\label{ConstantArray}\index{arrays!constant promotion}

For a distribution {\tt D} and a constant or final variable {\tt v} of
type {\tt T} the expression {\tt D v} denotes the mutable array with
distribution {\tt D} and base type {\tt T} initialized with {\tt v}.

\subsection{Restriction of an array}\index{array!restriction}

Let {\tt D1} be a sub-distribution of {\tt D}. Then {\tt a[D1]}
represents the sub-array of {\tt a} with the distribution {\tt D1}.

Recall that a rich set of operators are available on distributions
(\S~\ref{XtenDistributions}) to obtain sub-distributions
(e.g. restricting to a sub-region, to a specific place etc).

\subsection{Assembling an array}
Let {\tt a1,a2} be arrays of the same base type {\tt T} defined over
distributions {\tt D1} and {\tt D2} respectively. Assume that both
arrays are value or reference arrays. 
\paragraph{Assembling arrays over disjoint regions}\index{array!union!disjoint}

If {\tt D1} and {\tt D2} are disjoint then the expression {\tt a1 ||
a2} denotes the unique array of base type {\tt T} defined over the
distribution {\tt D1 || D2} such that its value at point {\tt p} is
{\tt a1[p]} if {\tt p} lies in {\tt D1} and {\tt a2[p]}
otherwise. This array is a reference (value) array if {\tt a1} is.

\paragraph{Overlaying an array on another}\index{array!union!asymmetric}
The expression
{\tt a1.over(a2)} (read: the array {\tt a1} {\em overlaid on} {\tt a2})
represents an array whose underlying region is the union of that of
{\tt a1} and {\tt a2} and whose distribution maps each point {\tt p}
in this region to {\tt D1[p]} if that is defined and to {\tt D2[p]}
otherwise. The value {\tt a1.over(a2)[p]} is {\tt a1[p]} if it is defined and {\tt a2[p]} otherwise.

This array is a reference (value) array if {\tt a1} is.

\todo{Define Flooding of arrays}

\todo{Wrapping an array}

\todo{Extending an array in a given direction.}

\subsection{Global operations }
\paragraph{Pointwise operations}\label{ArrayPointwise}\index{array!pointwise operations}
Suppose that {\tt m} is an operation defined on type {\tt T} that
takes an argument of type {\tt S} and returns a value of type {\tt
R}. Such an operation can be lifted pointwise to operate on a {\tt T}
array and an {\tt S} array defined over the same distribution {\tt D}
to return an {\tt R} array defined over {\tt D}.

The syntax for such pointwise application is {\tt a.m(b)} where {\tt a} and
{\tt b} are {\tt D} arrays.

\paragraph{Reductions}\label{ArrayReductions}\index{array!reductions}

Let {\tt m} be a reduction operator (\S~\ref{ReductionOperator})
defined on type {\tt T}. Let {\tt a} be a value or reference array
over base type {\tt T}. Then the operation {\tt a>>m()} returns a
value of type {\tt T} obtained by performing {\tt m} on all points in
{\tt a} in some order.

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation guarantees that
only one value of type {\tt T} is communicated from a place as part of
this reduction process.

\paragraph{Scans}\label{ArrayScans}\index{array!scans}

Let {\tt m} be a reduction operator (\S~\ref{ReductionOperator})
defined on type {\tt T}. Let {\tt a} be a value or reference array
over base type {\tt T} and distribution {\tt D}. Then the operation
{\tt a||m()} returns an array of base type {\tt T} and distribution
{\tt D} whose {\tt i}th element (in canonical order) is obtained by
performing the reduction {\tt m} on the first {\tt i} elements of {\tt
a} (in canonical order).

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation will endeavour to
minimize the communication between places to implement this operation.

