\section{Coercions and conversions}
\label{XtenConversions}
\index{conversions}\index{coercions}

\XtenCurrVer{} supports the following coercions and conversions

\subsection{Coercions}

A {\em coercion} does not change object identity;
a coerced object may be explicitly coerced back to its original
type through a cast.

\paragraph{Subsumption coercion.}
A subtype may be implicitly coerced to any supertype.
\index{coercions!subsumption coercion}

\paragraph{Explicit coercion (casting with \xcd"as")}
A reference type may be explicitly coerced to any other
reference type using the \xcd"as" operation.
If the value coerced is not an instance of the target type,
a \xcd"TypeCastException" is thrown.  Casting to a constrained
type may require a run-time check that the constraint is
satisfied.
\index{coercions!explicit coercion}
\index{casting}
\index{\Xcd{as}}

\subsection{Conversions}

A {\em conversion} may change object identity if the type being
converted to is not the same as the type converted from.

\paragraph{Narrowing conversion.}
A value class may be explicitly converted to any superclass
using the \xcd"to" operation.

\index{conversions!narrowing conversions}

\paragraph{Widening numeric conversion.}
A numeric type may be implicitly converted to a wider
numeric type.
In particular, an implicit conversion may be performed between
a numeric type and a type to its right, below:
\begin{xten}
Byte < Short < Int < Long < Float < Double
\end{xten}

\index{conversions!widening conversions}
\index{conversions!numeric conversions}

\paragraph{Boxing conversion.}
A type \xcd"T" may be implicitly converted to the
boxed type \xcd"Box[T]".

\index{conversions!boxing conversions}
\index{autoboxing}
\index{boxing}

\paragraph{Unboxing conversion.}
A value type \xcd"Box[T]" may be implicitly converted to the
type \xcd"T".

\index{conversions!unboxing conversions}
\index{autoboxing}
\index{unboxing}

\paragraph{Explicit conversion (casting with \xcd"to")}
A type may be explicitly converted to any type to which it can be
implicitly converted or implicitly or explicitly coerced.

\index{conversions!explicit conversion}
\index{casting}
\index{\Xcd{to}}

\paragraph{String conversion.}
Any object that is an operand of the binary
\xcd"+" operator may
be converted to \xcd"String" if the other operand is a \xcd"String".
A conversion to \xcd"String" is performed by invoking the \xcd"toString()"
method of the object.

\index{conversions!string conversion}

