\chapter{Arrays}\label{XtenArrays}\index{arrays}

An array is a mapping from a region (set of points) to a range data
type distributed over one or more places.
Multiple arrays may be declared with the same underlying
distribution.
The distribution underlying an array \xcd"a" may be obtained through
the field \xcd"a.dist".
\index{arrays!distribution@{\tt distribution}}

\section{Points}\label{point-syntax}\index{point syntax}

Arrays are indexed by points--$n$-dimensional tuples of
integers, implemented by the class \xcd"x10.lang.Point".
\Xten{} specifies a simple syntax for the construction of points.
A rail constructor (\Sref{RailConstructors}) of type \xcd"ValRail[Int]"
%or
%\xcd"ValRail[Long]" array
of length $n$
can be implicitly coerced to a \xcd"Point" of rank $n$.  For
example, the following code initializes \xcd"p" to a point of
rank two using a rail constructor:

\begin{xten}
p: Point = [1,2];
\end{xten}

The \xcd"Point" constructor can take a rail constructor as
argument.  The assignment above can be written, without
implicit coercion, as:

\begin{xten}
p: Point = new Point([1,2]);
\end{xten}

Points implement the function type \xcd"(Int) => Int"; thus, the
\xcd"i"the element of a point \xcd"p" may be accessed as \xcd"p(i)".
If \xcd"i" is out of range, an
\xcd"ArrayIndexOutOfBoundsException" is thrown.

\input{XtenRegions}
\input{XtenDistributions}
\section{Array initializer}\label{ArrayInitializer}\label{array!creation}

Arrays are instantiated by invoking a factory method for the
class \xcd"Array".

\eat{
An array instantiation may be annotated
\xcd"unsafe"
if it is intended to be
allocated in an unmanaged region (e.g., for communication with native
code). A value array is an immutable array. An array creation
must take either an \xcd"Int" as an argument or a \xcd"Dist". In the first
case an array is created over the distribution \xcd"[0:N-1]->here";
in the second over the given distribution. 
}

An array creation operation may also specify an initializer
function.
The function is applied in parallel
at all points in the domain of the distribution. The array
construction operation terminates locally only when the array has been
fully created and initialized (at all places in the range of the
distribution).

For instance:
\begin{xten}
val data : Array[Int]
    = Array.make[Int](1000->here, Point(i) => i);
val data2 : Array[Int]
    = Array.make[Int]([1:1000,1:1000]->here, Point(i,j) => i*j);
\end{xten}

{}\noindent 
The first declaration stores in \xcd"data" a reference to a
array with \xcd"1000" elements each of which is located in the
same place as the array. Each array component is initialized to \xcd"i".

The second declaration stores in \xcd"data2" a
2-d array over \xcd"[1:1000, 1:1000]" initialized with \xcd"i*j"
at point \xcd"[i,j]". It uses a more abbreviated form to specify 
the array initializer function.

Other examples:
\begin{xten}
val data : Array[Int]
    = Array.make[Int](1000, ((i): Point) => i*i);
val d : Array[Float](D)
    = Array.make[Float](D, ((i): Point) => 10.0*i);
val result : Array[Float](D)
    = Array.make[Float](D, ((i,j): Point) => i+j);
\end{xten}

\section{Operations on arrays}
In the following let \xcd"a" be an array with distribution \xcd"D" and
base type \xcd"T". \xcd"a" may be mutable or immutable, unless
indicated otherwise.

\subsection{Element operations}\index{array!access}
The value of \xcd"a" at a point \xcd"p" in its region of definition is
obtained by using the indexing operation \xcd"a(p)". This operation
may be used on the left hand side of an assignment operation to update
the value. The operator assignments \xcd"a(i) op= e" are also available
in \Xten{}.

For array variables, the right-hand-side of an assignment must
have the same distribution \xcd"D" as an array being assigned. This
assignment involves
control communication between the sites hosting \xcd"D". Each
site performs the assignment(s) of array components locally. The
assignment terminates when assignment has terminated at all
sites hosting \xcd"D".

\subsection{Constant promotion}\label{ConstantArray}\index{arrays!constant promotion}

For a distribution \xcd"D" and a constant or final variable \xcd"v" of
type \xcd"T" the expression \xcd"Array.make[T](D, (p: Point) => v)"
denotes the mutable array with
distribution \xcd"D" and base type \xcd"T" initialized with \xcd"v"
at every point.

\subsection{Restriction of an array}\index{array!restriction}

Let \xcd"D1" be a sub-distribution of \xcd"D". Then \xcd"a | D1"
represents the sub-array of \xcd"a" with the distribution \xcd"D1".

Recall that a rich set of operators are available on distributions
(\Sref{XtenDistributions}) to obtain sub-distributions
(e.g. restricting to a sub-region, to a specific place etc).

\subsection{Assembling an array}
Let \xcd"a1,a2" be arrays of the same base type \xcd"T" defined over
distributions \xcd"D1" and \xcd"D2" respectively. Assume that both
arrays are value or reference arrays. 
\paragraph{Assembling arrays over disjoint regions}\index{array!union!disjoint}

If \xcd"D1" and \xcd"D2" are disjoint then the expression \xcd"a1 || a2" denotes the unique array of base type \xcd"T" defined over the
distribution \xcd"D1 || D2" such that its value at point \xcd"p" is
\xcd"a1(p)" if \xcd"p" lies in \xcd"D1" and \xcd"a2(p)"
otherwise. This array is a reference (value) array if \xcd"a1" is.

\paragraph{Overlaying an array on another}\index{array!union!asymmetric}
The expression
\xcd"a1.overlay(a2)" (read: the array \xcd"a1" {\em overlaid with} \xcd"a2")
represents an array whose underlying region is the union of that of
\xcd"a1" and \xcd"a2" and whose distribution maps each point \xcd"p"
in this region to \xcd"D2(p)" if that is defined and to \xcd"D1(p)"
otherwise. The value \xcd"a1.overlay(a2)(p)" is \xcd"a2(p)" if it is defined and \xcd"a1(p)" otherwise.

This array is a reference (value) array if \xcd"a1" is.

The expression \xcd"a1.update(a2)" updates the array \xcd"a1" in place
with the result of \xcd"a1.overlay(a2)".

\oldtodo{Define Flooding of arrays}

\oldtodo{Wrapping an array}

\oldtodo{Extending an array in a given direction.}

\subsection{Global operations }

\paragraph{Pointwise operations}\label{ArrayPointwise}\index{array!pointwise operations}
The unary \xcd"lift" operation applies a function to each element of
an array, returning a new array with the same distribution.
The \xcd"lift" operation is implemented by the following method
in \xcd"Array[T]":
\begin{xten}
def lift[S](f: (T) => S): Array[S](dist);
\end{xten}

The binary \xcd"lift" operation takes a binary function and
another
array over the same distribution and applies the function
pointwise to corresponding elements of the two arrays, returning
a new array with the same distribution.
The \xcd"lift" operation is implemented by the following method
in \xcd"Array[T]":
\begin{xten}
def lift[S,R](f: (T,S) => R, Array[S](dist)): Array[R](dist);
\end{xten}

\paragraph{Reductions}\label{ArrayReductions}\index{array!reductions}

Let \xcd"f" be a function of type \xcd"(T,T)=>T".  Let
\xcd"a" be a value or reference array over base type \xcd"T".
Let \xcd"unit" be a value of type \xcd"T".
Then the
operation \xcd"a.reduce(f, unit)" returns a value of type \xcd"T" obtained
by performing \xcd"f" on all points in \xcd"a" in some order, and in
parallel.  The function \xcd"f" must be associative and
commutative.  The value \xcd"unit" should satisfy
\xcd"f(unit,x)" \xcd"==" \xcd"x" \xcd"==" \xcd"f(x,unit)".

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation guarantees that
only one value of type \xcd"T" is communicated from a place as part of
this reduction process.

\paragraph{Scans}\label{ArrayScans}\index{array!scans}

Let \xcd"f" be a reduction operator defined on type \xcd"T". Let
\xcd"a" be a value or reference array over base type \xcd"T" and
distribution \xcd"D". Then the operation \xcd"a||f()" returns an array
of base type \xcd"T" and distribution \xcd"D" whose $i$th element
(in canonical order) is obtained by performing the reduction \xcd"f"
on the first $i$ elements of \xcd"a" (in canonical order).

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation will endeavour to
minimize the communication between places to implement this operation.

Other operations on arrays may be found in \xcd"x10.lang.Array" and
other related classes.
