\documentclass[natbib]{sigplanconf}
%\documentclass{llncs}
\usepackage{times}
\usepackage{ifthen}
\usepackage{bcprules}
\usepackage{listings}
\usepackage{url}

\newenvironment{code}{\begin{center}\begin{small}\begin{tt}\begin{tabbing}}
{\end{tabbing}\end{tt}\end{small}\end{center}}

\newcommand{\subf}[2]{\begin{minipage}[t]{3in} #2 \end{minipage}\vspace{-.1in}\\\begin{minipage}[t]{3in}\center (#1) \end{minipage}\\}

\newcommand{\twocolfig}[2]{\begin{minipage}[t]{3in} #1 \end{minipage}\hspace{.5in}\begin{minipage}[t]{3in} #2 \end{minipage}}

\newcommand{\bug}[1]{\typeout{BUG: #1}$\spadesuit$ {\bf #1}}

\newcommand{\todo}[1]{\textbf{[[{#1}]]}}

\begin{document}

\conferenceinfo{3rd Workshop on Refactoring Tools '09} {October 26, 2009, Orlando, FL}
\CopyrightYear{2009}
\copyrightdata{978-1-60558-909-1}

\title{Extracting Concurrency via Refactoring in X10}

\authorinfo{Shane A. Markstrum}{Bucknell University\\Computer Science Department\\Lewisburg, PA USA}{shane.markstrum@bucknell.edu}
\authorinfo{Robert M. Fuhrer}{IBM T.J. Watson Research Center\\Hawthorne, NY USA}{rfuhrer@us.ibm.com}

%\author{Shane Markstrum}

%\institute{IBM Research \\ \email{smarkst@us.ibm.com}}

\date{7/30/08}

\maketitle

\begin{abstract}
In this paper, we present our vision of refactoring 
support for languages with a partitioned global address space memory model as 
embodied in the X10 programming language. We examine a novel refactoring, 
\emph{extract concurrent}, that introduces additional concurrency 
into a program by arranging for selected code in a loop body to run in parallel
with future iterations of the loop. We discuss the mechanisms and algorithms required
to implement the refactoring; provide insights on how development of this
refactoring aids designing future concurrency refactorings; and outline new
strategies for determining the effectiveness of concurrent refactorings.
\end{abstract}

\category{D.1.3}{Programming Techniques}{Concurrent Programming}
\category{D.2.6}{Software Engineering}{Programming Environments}[Integrated
environments]
\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming
Languages}[Program analysis]

\terms
Algorithms, Languages


\input{intro}
\input{description}
\input{algorithm}
\input{evaluation}
\input{relatedwork}

\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem{TSF}
F.~Bodin, Y.~M{\'e}vel, and R.~Quiniou.
\newblock A user level program transformation tool.
\newblock In {\em International Conference on Supercomputing}, pages 180--187,
  1998.

\bibitem{Charles05}
P.~Charles, C.~Grothoff, V.~Saraswat, C.~Donawa, A.~Kielstra, K.~Ebcioglu,
  C.~von Praun, and V.~Sarkar.
\newblock X10: an object-oriented approach to non-uniform cluster computing.
\newblock In {\em OOPSLA '05: Proceedings of the 20th annual ACM SIGPLAN
  conference on Object oriented programming, systems, languages, and
  applications}, pages 519--538, New York, NY, USA, 2005. ACM Press.

\bibitem{eclipse}
{Eclipse} home page.
\newblock \url{http://www.eclipse.org}.

\bibitem{ElGhazawi03}
T.~A. El-Ghazawi, W.~W. Carlson, and J.~M. Draper.
\newblock {U}{P}{C} language specifications v1.1.1, October 2003.

\bibitem[Hall et~al.(1993)Hall, Harvey, Kennedy, McIntosh, McKinley, Oldham,
  Paleczny, and Roth]{Hall93}
M.~W. Hall, T.~Harvey, K.~Kennedy, N.~McIntosh, K.~S. M\raisebox{.2em}{c}Kinley,
J.~D. Oldham, M.~Paleczny, and G.~Roth.
\newblock Experiences using the {Para{Scope} {Editor}}: an interactive parallel
  programming tool.
\newblock In \emph{Proceedings of the Fourth {ACM} {SIGPLAN} Symposium on
  Principles and Practice of Parallel Programming}, San Diego, CA, 1993.
\newblock URL \url{citeseer.ist.psu.edu/hall93experiences.html}.

\bibitem{Kennedy91}
K.~Kennedy, K.~S. M\raisebox{.2em}{c}Kinley, and C.-W. Tseng.
\newblock Analysis and transformation in the {Para{Scope} {Editor}}.
\newblock In {\em Proceedings of the 1991 {ACM} International Conference on
  Supercomputing}, Cologne, Germany, 1991.

\bibitem{Liao99}
S.-W. Liao, A.~Diwan, J.~Robert P.~Bosch, A.~Ghuloum, and M.~S. Lam.
\newblock {S}{U}{I}{F} {Explorer}: An interactive and interprocedural
  parallelizer.
\newblock In {\em PPoPP '99: Proceedings of the Seventh ACM SIGPLAN Symposium
  on Principles and Practice of Parallel Programming}, pages 37--48, New York,
  NY, USA, 1999. ACM Press.

\bibitem[Muraoka(1971)]{loopdist}
Yoichi Muraoka.
\newblock \emph{Parallelism exposure and exploitation in programs}.
\newblock PhD thesis, University of Illinois at Urbana-Champaign, Champaign,
  IL, USA, 1971.

\bibitem{Overbey05}
J.~Overbey, S.~Xanthos, R.~Johnson, and B.~Foote.
\newblock Refactorings for {Fortran} and high-performance computing.
\newblock In {\em SE-HPCS '05: Proceedings of the 2nd International Workshop on
  Software Engineering for High Performance Computing System Applications},
  pages 37--39, New York, NY, USA, 2005. ACM Press.

\bibitem[{S}{U}{I}{F} compiler system()]{SUIF}
{S}{U}{I}{F} compiler system.
\newblock The {S}{U}{I}{F} compiler system.
\newblock http://suif.stanford.edu/suif/suif2/.

\bibitem{X10}
V.~Saraswat.
\newblock Report on the experimental language {X10} v0.41.
\newblock http://www.research.ibm.com/x10/.

\bibitem{Yelick98}
K.~Yelick, L.~Semenzato, G.~Pike, C.~Miyamoto, B.~Liblit, A.~Krishnamurthy,
  P.~Hilfinger, S.~Graham, D.~Gay, P.~Colella, and A.~Aiken.
\newblock {Titanium}: {A} high-performance {Java} dialect.
\newblock In {ACM}, editor, {\em {ACM} 1998 Workshop on Java for
  High-Performance Network Computing}, New York, NY 10036, USA, 1998. ACM
  Press.

\end{thebibliography}

%\small{
%\bibliography{references}
%\bibliographystyle{plainnat}
%}

\end{document}
