%\documentclass[10pt,twoside,twocolumn,notitlepage]{report}
\documentclass[12pt,twoside,notitlepage]{report}
\usepackage{x10}
\usepackage{tenv}
\def\Hat{{\tt \char`\^}}
\usepackage{url}
\usepackage{times}
\usepackage{txtt}
\usepackage{ifpdf}
\usepackage{tocloft}
\usepackage{bcprules}
\usepackage{xspace}

\newif\ifdraft
%\drafttrue
\draftfalse

\pagestyle{headings}
\showboxdepth=0
\makeindex

\usepackage{commands}

\usepackage[
pdfauthor={Vijay Saraswat, Bard Bloom, Igor Peshansky, Olivier Tardieu, and David Grove},
pdftitle={Report on the Language X10},
pdfcreator={pdftex},
pdfkeywords={X10},
linkcolor=blue,
citecolor=blue,
urlcolor=blue
]{hyperref}

\ifpdf
          \pdfinfo {
              /Author   (Vijay Saraswat, Bard Bloom, Igor Peshansky, Olivier Tardieu, and David Grove)
              /Title    (Report on the Language X10)
              /Keywords (X10)
              /Subject  ()
              /Creator  (TeX)
              /Producer (PDFLaTeX)
          }
\fi

\def\headertitle{The \XtenCurrVer{} Report (Draft) }
\def\integerversion{2.0.5}

% Sizes and dimensions

%\topmargin -.375in       %    Nominal distance from top of page to top of
                         %    box containing running head.
%\headsep 15pt            %    Space between running head and text.

%\textheight 9.0in        % Height of text (including footnotes and figures, 
                         % excluding running head and foot).

%\textwidth 5.5in         % Width of text line.
\columnsep 15pt          % Space between columns 
\columnseprule 0pt       % Width of rule between columns.

\parskip 5pt plus 2pt minus 2pt % Extra vertical space between paragraphs.
\parindent 0pt                  % Width of paragraph indentation.
%\topsep 0pt plus 2pt            % Extra vertical space, in addition to 
                                % \parskip, added above and below list and
                                % paragraphing environments.


\newif\iftwocolumn

\makeatletter
\twocolumnfalse
\if@twocolumn
\twocolumntrue
\fi
\makeatother

\iftwocolumn

\oddsidemargin  0in    % Left margin on odd-numbered pages.
\evensidemargin 0in    % Left margin on even-numbered pages.

\else

\oddsidemargin  .5in    % Left margin on odd-numbered pages.
\evensidemargin .5in    % Left margin on even-numbered pages.

\fi


\newtenv{example}{Example}[section]
\newtenv{planned}{Planned}[section]

\begin{document}

% \parindent 0pt %!! 15pt                    % Width of paragraph indentation.

%\hfil {\bf 7 Feb 2005}
%\hfil \today{}

\input{first} 

\clearpage

{\parskip 0pt
\addtolength{\cftsecnumwidth}{0.5em}
\addtolength{\cftsubsecnumwidth}{0.5em}
%\addtolength{\cftsecindent}{0.5em}
\addtolength{\cftsubsecindent}{0.5em}
\tableofcontents
}

\input{Intro}
\input{Overview}
\input{Lex}
\input{Types}	
\input{Vars}
\input{Packages}
\input{Interfaces}
\input{Classes}
\input{Structs}
\input{Functions2}
\input{Expressions}	
\input{Statements}	
\input{Places}	
\input{Activities}	
\input{Clocks}	
\input{Arrays}	
\input{Annotations}
\input{extern}
\input{Lost}
%\input{Grammar}
\renewcommand{\bibname}{References}
\bibliographystyle{plain}
\bibliography{master}

%\input{bib}	

% Adjustment to avoid having the last index entry on a page by itself.
%\addtolength{\baselineskip}{-0.1pt}

\clearpage
\input{x10.ind}

\appendix

\chapter{Change Log}

\section{Changes from \Xten{} v2.0}

\begin{itemize}
\item \Xcd{Any} is now the top of the type hierarchy (every object,
  struct and function has a type that is a subtype of
  \Xcd{Any}). \Xcd{Any} defines \Xcd{home}, \Xcd{at}, \Xcd{toString},
  \Xcd{typeName}, \Xcd{equals} and \Xcd{hashCode}. \Xcd{Any} also defines the methods
  of \Xcd{Equals}, so \Xcd{Equals} is not needed any more.
\item Revised discussion of incomplete types (\Sref{ProtoRules}).
\item The manual has been revised and brought into line with the current implementation. 
\end{itemize}
\section{Changes from \Xten{} v1.7}

The language has changed in the following way:
\begin{itemize}

\item {\bf Type system changes}: There are now three kinds of entities
  in an \Xten{} computation: objects, structs and functions. Their
  associated types are class types, struct types and function
  types. 

  Class and struct types are called {\em container types} in that they
  specify a collection of fields and methods. Container types have a
  name and a signature (the collection of members accessible on that
  type). Collection types support primitive equality \Xcd{==} and may
  support user-defined equality if they implement the {\tt
    x10.lang.Equals} interface. 

  Container types (and interface types) may be further qualified with
  constraints.

  A function type specifies a set of arguments and their type, the
  result type, and (optionally) a guard. A function application
  type-checks if the arguments are of the given type and the guard is
  satisfied, and the return value is of the given type.  A function
  type does not permit \Xcd{==} checks. Closure literals create
  instances of the corresponding function type.

  Container types may implement interfaces and zero or more function
  types.

  All types support a basic set of operations that return a string
  representation, a type name, and specify the home place of the entity.

  The type system is not unitary. However, any type may be used to
  instantiate a generic type. 

  There is no longer any notion of \Xcd{value} classes. \Xcd{value}
  classes must be re-written into structs or (reference) classes. 

\item {\bf Global object model}: Objects are instances of
  classes. Each object is associated with a globally unique
  identifier. Two objects are considered identical \Xcd{==} if their
  ids are identical. Classes may specify \Xcd{global} fields and
  methods. These can be accessed at any place. (\Xcd{global} fields
  must be immutable.)

\item{\bf Proto types.} For the decidability of dependent type
  checking it is necessary that the property graph is acyclic. This is
  ensured by enforcing rules on the leakage of \Xcd{this} in
  constructors. The rules are flexible enough to permit cycles to be
  created with normal fields, but not with properties.

\item{Place types.} Place types are now implemented. This means that
  non-global methods can be invoked on a variable, only if the
  variable's type is either a struct type or a function type, or a
  class type whose constraint specifies that the object is located in
  the current place.

  There is still no support for statically checking array access
  bounds, or performing place checks on array accesses.

\end{itemize}
{\em The \Xten{} language has been developed as part of the IBM PERCS
Project, which is supported in part by the Defense Advanced Research
Projects Agency (DARPA) under contract No. NBCH30390004.}

{\em Java and all Java-based trademarks are trademarks of Sun Microsystems,
Inc. in the United States, other countries, or both.}
\end{document}

