

\chapter{Functions}
\label{Functions}
\label{functions}
\index{functions}
\label{Closures}

\section{Overview}
Functions, the last of the three kinds of values in X10, encapsulate pieces of
code which can be applied to a vector of arguments to produce a value.
Functions, when applied, can do nearly anything that any other code could do:
fail to terminate, throw an exception, offer values, modify variables, spawn activities,
execute in several places, and so on. X10 functions are not mathematical
functions: the \xcd`f(1)` may return \xcd`true` on one call and \xcd`false` on
an immediately following call.

It is a limitation of \XtenCurrVer{} that functions do not support
type arguments. This limitation may be removed in future versions of
the language.

A \emph{function literal} \xcd"(x1:T1,..,xn:Tn){c}:T=>e" creates a function of
type\\ \xcd"(x1:T1,...,xn:Tn){c}=>T" (\Sref{FunctionType}).  For example, 
\xcd`(x:Int) => x*x` is a function literal describing the squaring function on
integers.   
\xcd`null` is also a function value.

Function application is written \xcd`f(a,b,c)`, following common mathematical
usage. 
\index{Exception!unchecked}
Function invocation may throw unchecked exceptions. 

The function body may be a block.  To compute integer squares by repeated
addition (inefficiently), one may write: 
%~~gen
% package Functions.Are.For.Spunctions;
% class Examplllll {
% static 
%~~vis
\begin{xten}
val sq: (Int) => Int 
      = (n:Int) => {
           var s : Int = 0;
           val abs_n = n < 0 ? -n : n;
           for ((i) in 1..abs_n) s += abs_n;
           s
        };
\end{xten}
%~~siv
%}
%~~neg




A function literal evaluates to a function entity {$\phi$}. When {$\phi$} is
applied to a suitable list of actual parameters \xcd`a1`-\xcd`an`, it
evaluates \xcd`e` with the formal parameters bound to the actual parameters.
So, the following are equivalent, where \xcd`e` is an expression involving
\xcd`x1` and \xcd`x2`\footnote{Strictly, there are a few other requirements;
  \eg, \xcd`result` must be a \xcd`var` of type \xcd`T` defined outside the
  outer block, the variables \xcd`a1` and \xcd`a2` had better not appear in
  \xcd`e`, and everything in sight had better typecheck properly.}

%~~gen
% package functions2.why.is.there.a.two;
% abstract class FunctionsTooManyFlippingFunctions[T, T1, T2]{
% abstract def arg1():T1;
% abstract def arg2():T2;
% def thing1(e:T) {var result:T;
%~~vis
\begin{xten}
{
  val f = (x1:T1,x2:T2){true}:T => e;
  val a1 : T1 = arg1();
  val a2 : T2 = arg2();
  result = f(a1,a2);
}
\end{xten}
%~~siv
%}}
%~~neg
and 
%~~gen
% package functions2.why.is.there.a.two.but.here.is.the.other.one;
% abstract class FunctionsTooManyFlippingFunctions[T, T1, T2]{
% abstract def arg1():T1;
% abstract def arg2():T2;
% def thing1(e:T) {var result:T;
%~~vis
\begin{xten}
{
  val a1 : T1 = arg1();
  val a2 : T2 = arg2();
  {
     val x1 : T1 = a1;
     val x2 : T2 = a2;
     result = e;
  }  
}
\end{xten}
%~~siv
%}}
%~~neg
\noindent
This doesn't quite work if the body is a statement rather than an expression.
A few language features are forbidden (\xcd`break` or \xcd`continue` of a loop
that surrounds the function literal) or mean something different (\xcd`return`
inside a function returns from the function). 





The \emph{method selector expression} \Xcd{e.m.(x1:T1,...,xn:Tn)} (\Sref{MethodSelectors})
permits the specification of the function underlying
the method \Xcd{m}, which takes arguments of type \Xcd{(x1:T1,..., xn:Tn)}.
Within this function, \Xcd{this} is bound to the result of evaluating \Xcd{e}.

Function types may be used in \Xcd{implements} clauses of class
definitions. Instances of such classes may be used as functions of the
given type.  Indeed, an object may behave like any (fixed) number of
functions, since the class it is an instance of may implement any
(fixed) number of function types.

%\section{Implementation Notes}
%\begin{itemize}
%
%\item Note that e.m.(T1,...,Tn) will evaluate e to create a
%  function. This function will be applied later to given
%  arguments. Thus this syntax can be used to evaluate the receiver of
%  a method call ahead of the actual invocation. The resulting function
%  can be used multiple times, of course.
%\end{itemize}


\section{Function Literals}
\index{literal!function}
\label{FunctionLiteral}

\Xten{} provides first-class, typed functions, including
\emph{closures}, \emph{operator functions}, and \emph{method
  selectors}.

\begin{grammar}
ClosureExpression \:
        \xcd"("
        Formals\opt
        \xcd")"
\\ &&
        Guard\opt
        ReturnType\opt
        Throws\opt
        Offers\opt
        \xcd"=>" ClosureBody \\
ClosureBody \:
        Expression \\
        \| \xcd"{" Statement\star \xcd"}" \\
        \| \xcd"{" Statement\star Expression \xcd"}" \\
\end{grammar}

Functions have zero or more formal parameters, an optional return type, an
optional set of exceptions throws by the body, and an optional type offered by
the body.  The body has the
same syntax as a method body; it may be either an expression, a block
of statements, or a block terminated by an expression to return. In
particular, a value may be returned from the body of the function
using a return statement (\Sref{ReturnStatement}). 

The type of a
function is a function type (\Sref{FunctionType}).  In some cases the
return type \Xcd{T} is also optional and defaults to the type of the
body. If a formal \Xcd{xi} does not occur in any
\Xcd{Tj}, \Xcd{c}, \Xcd{T} or \Xcd{e}, the declaration \Xcd{xi:Ti} may
be replaced by just \Xcd{Ti}: \xcd`(Int)=>7` is the integer function returning
7 for all inputs.

\label{ClosureGuard}

As with methods, a function may declare a guard to
constrain the actual parameters with which it may be invoked.
The guard may refer to the type parameters, formal parameters,
and any \xcd`val`s in scope at the function expression.

The body of the function is evaluated when the function is
invoked by a call expression (\Sref{Call}), not at the function's
place in the program text.

As with methods, a function with return type \xcd"Void" cannot
have a terminating expression. 
If the return type is omitted, it is inferred, as described in
\Sref{TypeInference}.
It is a static error if the return type cannot be inferred.  \Eg,
\xcd`(Int)=>null` is not well-defined; X10 does not know which type of
\xcd`null` is intended.  
%~~exp~~`~~`~~ ~~
But \xcd`(Int):Rail[Double] => null` is legal.


\begin{example}
The following method takes a function parameter and uses it to
test each element of the list, returning the first matching
element.  It returns \xcd`otherwise` if no element matches.

%~~gen
% package functions2.oh.no;
% import x10.util.*;
% class Finder {
% static 
%~~vis
\begin{xten}

def find[T](f: (T) => Boolean, xs: List[T]!, absent:T): T = {
  for (x: T in xs)
    if (f(x)) return x;
  absent
  }
\end{xten}
%~~siv
% }
%~~neg

The method may be invoked thus:
%~~gen
% package functions2.oh.no.my.ears;
% import x10.util.*;
% class Finderator {
% static def find[T](f: (T) => Boolean, xs: x10.util.List[T]!, absent:T): T = {
%  for (x: T in xs)
%    if (f(x)) return x;
%  absent
%}
% static def checkery() {
%~~vis
\begin{xten}
xs: List[Int]! = new ArrayList[Int]();
x: Int = find((x: Int) => x>0, xs, 0);
\end{xten}
%~~siv
%}}
%~~neg

\end{example}

As with a normal method, the function may have a \xcd"throws"
clause. It is a static error if the body of the function throws a
checked exception that is not declared in the function's \xcd"throws"
clause.

Similarly, it may have an \Xcd{offers T} clause; it is a static error if the
body offers any value not of type \Xcd{T}.

\subsection{Outer variable access}

In a function
\xcdmath"(x$_1$: T$_1$, $\dots$, x$_n$: T$_n$){c} => { s }"
the types \xcdmath"T$_i$", the guard \xcd"c" and the body \xcd"s"
may access many, though not all, sorts of variables from outer scopes.  
Specifically, they can access: 
\begin{itemize}
\item All fields of the enclosing object and class;
\item All type parameters;
\item All \xcd`val` variables;
\item \xcd`var` variables with the \xcd`shared` annotation. 
\end{itemize}


\limitation{\xcd`shared` is not currently supported.}

The function body may refer to instances of enclosing classes using
the syntax \xcd"C.this", where \xcd"C" is the name of the
enclosing class.  \xcd`this` refers to the instance of the immediately
enclosing class, as usual.

For example, the following is legal.  However, it would not be legal to add
\xcd`e` or \xcd`h` to the sum; they are non-\xcd`shared` \xcd`var`s from the
surrounding scope.

%%TODO -- this example uses 'shared', which is not currently available.
\begin{xten}
class Lambda {
   var a : Int = 0;
   val b = 0;
   def m(var c : Int, shared var d : Int,  val e : Int) {
      var f : Int = 0;
      shared var g : Int = 0;
      val h : Int = 0;
      val closure = (var i: Int, val j: Int) => {
    	  return a + b + d + g + i + j + this.a + Lambda.this.a;
      };
      return closure;
   }
}
\end{xten}


{\bf Rationale:} Non-\xcd`shared` \xcd`var`s like \xcd`e` and \xcd`h` are
excluded in X10, as in many other languages, for practical implementation
reasons. They are allocated on the stack, which is desirable for efficiency.
However, the closure may exist for long after the stack frame containing
\xcd`e` and \xcd`h` has been freed, so those storage locations are no longer
valid for those variables. \xcd`shared var`s are heap-allocated, which is less
efficient but allows them to exist after \xcd`m` returns. 


\xcd`shared` does not guarantee {\bf atomic} access to the shared variable. As
with any code that might mutate shared data concurrently, be sure to protect
references to mutable shared state with \xcd`atomic`. For example, the
following code returns a pair of closures which operate on the same shared
variable \xcd`a`, which are concurrency-safe---even if invoked many times
simultaneously. Without \xcd`atomic`, it would no longer be concurrency-safe.


%~fails~gen
% package Functions2.Are.All.Too.Much;
% class Fun2Frivols {
%~fails~vis
\begin{xten}
  def counters() {
      shared var a : Int = 0;
       return [
          () => {atomic a ++;},
          () => {atomic return a;}
          ];
   }
\end{xten}
%~fails~siv
%}
%
%~fails~neg


%SHARED% \begin{note}
%SHARED% The main activity may run in parallel with any
%SHARED% functions it creates. Hence even the read of an outer variable by the
%SHARED% body of a function may result in a race condition. Since functions are
%SHARED% first-class, the analysis of whether a function may execute in parallel
%SHARED% with the activity that created it may be difficult.
%SHARED% \end{note}

%% vj: This should be verified.
%\begin{note}
%The rule for accessing outer variables from function bodies
%should be the same as the rule for accessing outer variables from local
%or anonymous classes.
%\end{note}

\section{Method selectors}
\label{MethodSelectors}
\index{function!method selector}
\index{method!underlying function}

A method selector expression allows a method to be used as a
first-class function, without writing a function expression for it.
For example, consider a class \xcd`Span` defining ranges of integers.  

%~~gen
% package Functions2.Span;
%~~vis
\begin{xten}
class Span(low:Int, high:Int) {
  def this(low:Int, high:Int) {property(low,high);}
  def between(n:Int) = low <= n && n <= high;
  def example() {
    val digit = new Span(0,9);
    val isDigit : (Int) => Boolean = digit.between.(Int);
    if (isDigit(8)) x10.io.Console.OUT.println("8 is!");
  }
}
\end{xten}
%~~siv
%
%~~neg
\noindent


In \xcd`example()`, 
%~~exp~~`~~`~~ digit:Span!~~class Span(low:Int, high:Int) {def this(low:Int, high:Int) {property(low,high);} def between(n:Int) = low <= n && n <= high;}
\xcd`digit.between.(Int)` 
is a unary function testing whether its argument is between zero
and nine.  It could also be written 
%~~exp~~`~~`~~ digit:Span!~~class Span(low:Int, high:Int) {def this(low:Int, high:Int) {property(low,high);} def between(n:Int) = low <= n && n <= high;}
\xcd`(n:Int) => digit.between(n)`.

This is formalized thus:

\begin{grammar}
MethodSelector \:
        Primary \xcd"."
        MethodName \xcd"."
                TypeParameters\opt \xcd"(" Formals\opt \xcd")" \\
      \|
        TypeName \xcd"."
        MethodName \xcd"."
                TypeParameters\opt \xcd"(" Formals\opt \xcd")" \\
\end{grammar}

The \emph{method selector expression} \Xcd{e.m.(T1,...,Tn)} is type
correct only if  the static type of \Xcd{e} is a
class or struct or interface \xcd`V` with a method
\Xcd{m(x1:T1,...xn:Tn)\{c\}:T} defined on it (for some
\Xcd{x1,...,xn,c,T)}. At runtime the evaluation of this expression
evaluates \Xcd{e} to a value \Xcd{v} and creates a function \Xcd{f}
which, when applied to an argument list \Xcd{(a1,...,an)} (of the right
type) yields the value obtained by evaluating \Xcd{v.m(a1,...,an)}.

Thus, the method selector

\begin{xtenmath}
e.m.[X$_1$, $\dots$, X$_m$](T$_1$, $\dots$, T$_n$)
\end{xtenmath}
\noindent behaves as if it were the function
\begin{xtenmath}
((v:V)=>
  [X$_1$, $\dots$, X$_m$](x$_1$: T$_1$, $\dots$, x$_n$: T$_n$){c} 
  => v.m[X$_1$, $\dots$, X$_m$](x$_1$, $\dots$, x$_n$))
(e)
\end{xtenmath}


\limitation{X10 functions, including method selectors, do not currently accept
generic arguments.}

Because of overloading, a method name is not sufficient to
uniquely identify a function for a given class (in Java-like languages).
One needs the argument type information as well.
The selector syntax (dot) is used to distinguish \xcd"e.m()" (a
method invocation on \xcd"e" of method named \xcd"m" with no arguments)
from \xcd"e.m.()"
(the function bound to the method). 

A static method provides a binding from a name to a function that is
independent of any instance of a class; rather it is associated with the
class itself. The static function selector
\xcdmath"T.m.(T$_1$, $\dots$, T$_n$)" denotes the
function bound to the static method named \xcd"m", with argument types
\xcdmath"(T$_1$, $\dots$, T$_n$)" for the type \xcd"T". The return type
of the function is specified by the declaration of \xcd"T.m".

Users of a function type do not care whether a function was defined
directly (using the function syntax), or obtained via (static or
instance) function selectors.


\section{Operator functions}
\label{OperatorFunction}
\index{function!operator}
Every operator (e.g.,
\xcd"+",
\xcd"-",
\xcd"*",
\xcd"/",
\dots) has a family of functions, one for
each type on which the operator is defined. The function can be
selected using the "." syntax:

\begin{grammar}
OperatorFunction
        \: TypeName \xcd"." Operator \xcd"(" Formals\opt \xcd")" \\
        \| TypeName \xcd"." Operator \\
\end{grammar}

If an operator has more than one arity (\eg, unary and binary
\xcd"-"), the unary version may be selected by giving the
formal parameter types.  The binary version is selected by
default, or the types may be specified for clarity.
For example, the following equivalences hold:

\begin{xtenmath}
String.+             $\equiv$ (x: String, y: String): String => x + y
Long.-               $\equiv$ (x: Long, y: Long): Long => x - y
Float.-(Float,Float) $\equiv$ (x: Float, y: Float): Float => x - y
Int.-(Int)           $\equiv$ (x: Int): Int => -x
Boolean.&            $\equiv$ (x: Boolean, y: Boolean): Boolean => x & y
Boolean.!            $\equiv$ (x: Boolean): Boolean => !x
Int.<(Int,Int)       $\equiv$ (x: Int, y: Int): Boolean => x < y
Dist.|(Place)        $\equiv$ (d: Dist, p: Place): Dist => d | p
\end{xtenmath}


%%TODO -- fix commented-out lines!

%~~gen
% package Functions2.For.The.Lose;
% class TypecheckThatSillyExample {
%   def checker() {
%    val l1 : (String, String) => String = String.+;
%    val r1 : (String, String) => String = (x: String, y: String): String => x + y;
%    val l2 : (Long,Long) => Long = Long.-;
%    val r2 : (Long,Long) => Long = (x: Long, y: Long): Long => x - y;
%//var v1 : (Float,Float) => Float = Float.-(Float,Float) ;
%var v2 : (Float,Float) => Float = (x: Float, y: Float): Float => x - y;
%//var v3 : (Int) => Int =  Int.-(Int)     ;      ;
%var v4  : (Int) => Int  =  (x: Int): Int => -x;
%var v5 : (Boolean,Boolean) => Boolean = Boolean.&            ;
%var v6 : (Boolean,Boolean) => Boolean =  (x: Boolean, y: Boolean): Boolean => x & y;
%//var v7 : (Boolean) => Boolean = Boolean.!            ;
%var v8 : (Boolean) => Boolean =  (x: Boolean): Boolean => !x;
%//var v9 : (Int,Int) => Boolean = Int.<(Int,Int)       ;
%var v10: (Int,Int) => Boolean =  (x: Int, y: Int): Boolean => x < y;
%//var v11 : (Dist,Place)=>Dist = Dist.|(Place)        ;
%var v12 : (Dist,Place)=>Dist=  (d: Dist, p: Place): Dist => d | p;
%}
% }
%~~vis
%~~siv
%
%~~neg

Unary and binary promotion (\Sref{XtenPromotions}) is not performed
when invoking these
operations; instead, the operands are coerced individually via implicit
coercions (\Sref{XtenConversions}), as appropriate.


\begin{planned}

{\bf The following is not implemented in version 2.0.3:}

Additionally, for every expression \xcd"e" of a type \xcd"T" at which a binary
operator \xcd"OP" is defined, the expression \xcd"e.OP" or
\xcd"e.OP(T)" represents the function
defined by:

\begin{xten}
(x: T): T => { e OP x }
\end{xten}

\begin{grammar}
Primary \: Expr \xcd"." Operator \xcd"(" Formals\opt \xcd")" \\
        \| Expr \xcd"." Operator \\
\end{grammar}

%% For every expression \xcd"e" of a type \xcd"T" at which a unary
%%operator \xcd"OP" is defined, the expression \xcd"e.OP()"
%% represents the function defined by:

%% \begin{xten}
%% (): T => { OP e }
%% \end{xten}

For example,
one may write an expression that adds one to each member of a
list \xcd"xs" by:

%%TODO -- when this topic works, make the example wwork too.
%~x~gen
% package Functions2.Wants.A.Dinner.Reservation;
% import x10.util.*;
% class Reservation {
% def smerp() {
%   val xs = new ArrayList[Int]();
%~x~vis
\begin{xten}
xs.map(1.+);
\end{xten}
%~x~siv
% }
% }
%
%~x~neg
\end{planned}


\section{Functions as objects of type \Xcd{Any}}
\label{FunctionAnyMethods}

\label{FunctionEquality}
\index{function!equality} \index{equality!function} Two functions \Xcd{f} and
\Xcd{g} are equal (``\Xcd{==}'') if both are instances of classes and the same
object, or if both were obtained by the same evaluation of a function
literal.\footnote{A literal may occur in program text within a loop, and hence
  may be evaluated multiple times.} Further, it is guaranteed that if two
functions are equal then they refer to the same locations in the environment
and represent the same code, so their executions in an identical situation are
indistinguishable. (Specifically, if \xcd`f == g`, then \xcd`f(1)` can be
substituted for \xcd`g(1)` and the result will be identical. However, there is
no guarantee that \xcd`f(1)==g(1)` will evaluate to true. Indeed, there is no
guarantee that \xcd`f(1)==f(1)` will evaluate to true either, as \xcd`f` might
be a function which returns {$n$} on its {$n^{th}$} invocation. However,
\xcd`f(1)==f(1)` and \xcd`f(1)==g(1)` are interchangeable.)
\index{function!==}


Every function type implements all the methods of \Xcd{Any}.
b\xcd`f.equals(g)` is equivalent to \xcd`f==g`.  \xcd`f.hashCode()`, 
\xcd`f.toString()`, and \xcd`f.typeName()` are implementation-dependent, but
respect \xcd`equals` and the basic contracts of \xcd`Any`. 
\xcd`f.home()` returns \xcd`here` and \xcd`f.at(x)`
always returns true, as for structs.

\index{function!equals}
\index{function!hashCode}
\index{function!toString}
\index{function!typeName}
\index{function!home}
\index{function!at(Place)}
\index{function!at(Object)}


