\chapter{Arrays}\label{XtenArrays}\index{arrays}

An array is a mapping from a distribution to a range data
type. Multiple arrays may be declared with the same underlying
distribution. 

Each array has a field {\tt a.distribution} which may be used to
obtain the underlying distribution.\index{arrays!distribution@\tt distribution}

The distribution underlying an array {\tt a} may be obtained through
the field {\tt a.distribution}.

\input{XtenRegions}
\input{XtenDistributions}
\section{Array initializer}\label{ArrayInitializer}\label{array!creation}

\begin{x10}
450 ArrayCreationExpression ::= 
       new ArrayBaseType Unsafeopt [] 
         ArrayInitializer
451  | new ArrayBaseType Unsafeopt 
       [ Expression ]
452  | new ArrayBaseType Unsafeopt 
       [ Expression ] Expression
453  | new ArrayBaseType Unsafeopt 
       [ Expression ] 
       (FormalParameter) MethodBody
454  | new ArrayBaseType value Unsafeopt 
       [ Expression ]
455  | new ArrayBaseType value Unsafeopt 
       [ Expression ] Expression
456  | new ArrayBaseType value Unsafeopt 
       [ Expression ] 
       ( FormalParameter ) MethodBody
457 ArrayBaseType ::= PrimitiveType
458  | ClassOrInterfaceType
530 Unsafeopt ::=
531  | unsafe
\end{x10}

An array may be declared {\tt unsafe} if it is intended to be
allocated in an unmanaged region (e.g.{} for communication with native
code). A {\tt value} array is an immutable array. An array creation
must take either an int as an argument or a distribution. In the first
case an array is created over the distribution {\tt [0:N-1]-> here};
in the second over the given distribution. 

An array creation operation may also specify an initializer using the
abbreviated formalparameter/methodbody functional syntax. The formal
parameter may contain exploded parameters
(Section~\ref{exploded-syntax}). The function is applied in parallel
at all points in the domain of the distribution. The array
construction operation terminates locally only when the array has been
fully created and initialized (at all places in the range of the
distribution).

For instance:
\begin{x10}
int[.] data 
    = new int[1000->here] 
         new intArray.pointwiseOp()\{
             public int apply(point p[i])\{
               return i;
           \}
     \};
int[.] data2 
    = new int value [[1:1000,1:1000]->here] 
        (point p[i,j])\{ return i*j; \};
\end{x10}

{}\noindent 
The first declaration stores in {\tt data} a reference to a mutable
array with {\tt 1000} elements each of which is located in the
same place as the array. Each array component is initialized to {\tt i}.

The second declaration stores in {\tt data2} an (immutable)
2-d array over {\tt [1:1000, 1:1000]} initialized with {\tt i*j}
at point {\tt [i,j]}. It uses a more abbreviated form to specify 
the array initializer function.

Other examples:
\begin{x10}
int[.] data 
   = new int[1000](point [i])\{return i*i;\};
float[D] d 
   = new float[D] (point [i])\{return 10.0*i;\}; 
float[D] result 
   = new float[D] 
     (point [i,j]) \{return i + j; \};
\end{x10}

\section{Operations on arrays}
In the following let {\tt a} be an array with distribution {\tt D} and
base type {\tt T}. {\tt a} may be mutable or immutable, unless
indicated otherwise.

\subsection{Element operations}\index{array!access}
The value of {\tt a} at a point {\tt p} in its region of definition is
obtained by using the indexing operation {\tt a[p]}. This operation
may be used on the left hand side of an assignment operation to update
the value. The operator assignments {\tt a[i] op= e} are also available
in \Xten{}.

\subsection{Constant promotion}\label{ConstantArray}\index{arrays!constant promotion}

For a distribution {\tt D} and a constant or final variable {\tt v} of
type {\tt T} the expression {\tt new T[D](point p) \{ return v;\}}
{\tt D v} denotes the mutable array with
distribution {\tt D} and base type {\tt T} initialized with {\tt v}.

\subsection{Restriction of an array}\index{array!restriction}

Let {\tt D1} be a sub-distribution of {\tt D}. Then {\tt a | D1}
represents the sub-array of {\tt a} with the distribution {\tt D1}.

Recall that a rich set of operators are available on distributions
(\S~\ref{XtenDistributions}) to obtain sub-distributions
(e.g. restricting to a sub-region, to a specific place etc).

\subsection{Assembling an array}
Let {\tt a1,a2} be arrays of the same base type {\tt T} defined over
distributions {\tt D1} and {\tt D2} respectively. Assume that both
arrays are value or reference arrays. 
\paragraph{Assembling arrays over disjoint regions}\index{array!union!disjoint}

If {\tt D1} and {\tt D2} are disjoint then the expression {\tt a1 ||
a2} denotes the unique array of base type {\tt T} defined over the
distribution {\tt D1 || D2} such that its value at point {\tt p} is
{\tt a1[p]} if {\tt p} lies in {\tt D1} and {\tt a2[p]}
otherwise. This array is a reference (value) array if {\tt a1} is.

\paragraph{Overlaying an array on another}\index{array!union!asymmetric}
The expression
{\tt a1.overlay(a2)} (read: the array {\tt a1} {\em overlaid with} {\tt a2})
represents an array whose underlying region is the union of that of
{\tt a1} and {\tt a2} and whose distribution maps each point {\tt p}
in this region to {\tt D2[p]} if that is defined and to {\tt D1[p]}
otherwise. The value {\tt a1.overlay(a2)[p]} is {\tt a2[p]} if it is defined and {\tt a1[p]} otherwise.

This array is a reference (value) array if {\tt a1} is.

The expression {\tt a1.update(a2)} updates the array {\tt a1} in place
with the result of {\tt a1.overlay(a2)}.
\todo{Define Flooding of arrays}

\todo{Wrapping an array}

\todo{Extending an array in a given direction.}

\subsection{Global operations }
\paragraph{Pointwise operations}\label{ArrayPointwise}\index{array!pointwise operations}
Suppose that {\tt m} is an operation defined on type {\tt T} that
takes an argument of type {\tt S} and returns a value of type {\tt
R}. Such an operation can be lifted pointwise to operate on a {\tt T}
array and an {\tt S} array defined over the same distribution {\tt D}
to return an {\tt R} array defined over {\tt D}, using the
{\tt lift} operation, {\tt a.lift(f, b)}.


\paragraph{Reductions}\label{ArrayReductions}\index{array!reductions}

Let {\tt f} be a {\tt binaryOp} defined on type {\tt T} (e.g.{} see
the specification of the classes {\tt x10.lang.intArray}).  Let {\tt
a} be a value or reference array over base type {\tt T}. Then the
operation {\tt a.reduce(f)} returns a value of type {\tt T} obtained
by performing {\tt m} on all points in {\tt a} in some order, and in
parallel.

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation guarantees that
only one value of type {\tt T} is communicated from a place as part of
this reduction process.

\paragraph{Scans}\label{ArrayScans}\index{array!scans}

Let {\tt m} be a reduction operator defined on type {\tt T}. Let {\tt
a} be a value or reference array over base type {\tt T} and
distribution {\tt D}. Then the operation {\tt a||m()} returns an array
of base type {\tt T} and distribution {\tt D} whose {\tt i}th element
(in canonical order) is obtained by performing the reduction {\tt m}
on the first {\tt i} elements of {\tt a} (in canonical order).

This operation involves communication between the places over which
the array is distributed. The \Xten{} implementation will endeavour to
minimize the communication between places to implement this operation.

Other operations on arrays may be found in {\tt x10.lang.intArray} and
other related classes.
