\chapter{Interfaces}
\label{XtenInterfaces}\index{interface}

An interface specifies signatures for public methods, properties,
\xcd`static val`s, and an invariant. It may extend several interfaces, giving
X10 a large fraction of the power of multiple inheritance at a tiny fraction
of the cost.

The following puny example illustrates all these features: 
% TODO Well, it would if there weren't a compiler bug in the way.
%~~gen ^^^Interfaces_static_val
% package Interfaces_static_val;
% NOCOMPILE
%~~vis
\begin{xten}
interface Pushable(text:String, prio:Int) {
  def push(): void;
  static val MAX_PRIO = 100;
}
class MessageButton(text:String, prio:Int) 
  implements Pushable{self.prio==Pushable.MAX_PRIO} {
  public def push() { 
    x10.io.Console.OUT.println(text + " pushed");
  }
}
\end{xten}
%~~siv
%
%~~neg
\noindent
\xcd`Pushable` defines two properties, a method, and a static value.  
\xcd`MessageButton` implements a constrained version of \xcd`Pushable`,
\viz\ one with maximum priority.  It also has \xcd`Pushable`'s properties.  It
defines the \xcd`push()` method given in the interface, as a \xcd`public`
method---interface methods are implicitly \xcd`public`.

A concrete type---a class or struct---can {\em implement} an interface,
typically by having all the methods and properties that the interface
requires.

A variable may be declared to be of interface type.  Such a variable has all
the fields and methods declared (directly or indirectly) by the interface;
nothing else is statically available.  Values of a concrete type which
implement the interface may be stored in the variable.  


\label{DepType:Interface}

\begin{bbgrammar}
 NormalInterfaceDecl    \: Mods\opt \xcd"interface" Id TypeParamsWithVariance\opt Properties\opt WhereClause\opt ExtendsInterfaces\opt InterfaceBody & (\ref{prod:NormalInterfaceDecl})\\%<FROM #(prod:NormalInterfaceDecl)#
 TypeParamsWithVariance    \: \xcd"[" TypeParamWithVarianceList \xcd"]" & (\ref{prod:TypeParamsWithVariance})\\%<FROM #(prod:TypeParamsWithVariance)#
 Properties    \: \xcd"(" PropertyList \xcd")" & (\ref{prod:Properties})\\%<FROM #(prod:Properties)#
 PropertyList    \: Property & (\ref{prod:PropertyList})\\%<FROM #(prod:PropertyList)#
    \| PropertyList \xcd"," Property\\
 Property    \: Annotations\opt Id ResultType & (\ref{prod:Property})\\%<FROM #(prod:Property)#
 WhereClause    \: DepParams & (\ref{prod:WhereClause})\\%<FROM #(prod:WhereClause)#
 ExtendsInterfaces    \: \xcd"extends" Type & (\ref{prod:ExtendsInterfaces})\\%<FROM #(prod:ExtendsInterfaces)#
    \| ExtendsInterfaces \xcd"," Type\\
 InterfaceBody    \: \xcd"{" InterfaceMemberDecls\opt \xcd"}" & (\ref{prod:InterfaceBody})\\%<FROM #(prod:InterfaceBody)#
 InterfaceMemberDecls    \: InterfaceMemberDecl & (\ref{prod:InterfaceMemberDecls})\\%<FROM #(prod:InterfaceMemberDecls)#
    \| InterfaceMemberDecls InterfaceMemberDecl\\
 InterfaceMemberDecl    \: MethodDecl & (\ref{prod:InterfaceMemberDecl})\\%<FROM #(prod:InterfaceMemberDecl)#
    \| PropertyMethodDecl\\
    \| FieldDecl\\
    \| ClassDecl\\
    \| InterfaceDecl\\
    \| TypeDefDecl\\


\end{bbgrammar}

\noindent
The invariant associated with an interface is the conjunction of the
invariants associated with its superinterfaces and the invariant
defined at the interface. 

%%WRONG%% \begin{staticrule*}
%%WRONG%% The compiler declares an error if this constraint
%%WRONG%% is not consistent.  
%%WRONG%% \end{staticrule*}

Each interface implicitly defines a nullary getter method
\xcd"def p(): T" for each property \xcd"p: T". The interface may not have
another definition of a method \xcd`p()`. 



A class \xcd"C" is said to implement an interface \xcd"I" if
\begin{itemize}
\item \xcd`I`, or a subtype of \xcd`I`, appears in the \xcd`implements` list
      of \xcd`C`, 
\item \xcd`C`'s properties include all the properties of \xcd"I",
%%IMPLIED%% \item \xcd`C`'s class invariant $\mathit{inv}($\xcd"C"$)$ implies
%%IMPLIED%% $\mathit{inv}($\xcd"I"$)$.
\item Each method \xcd`m` defined by \xcd`I` is also a method of \xcd`C` --
      with the {\em  \xcd`public`} modifier added.   These methods may be
      \xcd`abstract` if \xcd`C` is \xcd`abstract`.
\end{itemize}

Note that, by definition of $\mathit{inv}(\cdot)$ (see \Sref{DepType:ClassGuardDef}, \xcd`C`'s class invariant $\mathit{inv}($\xcd"C"$)$ implies
$\mathit{inv}($\xcd"I"$)$.

\section{Field Definitions}
\index{interface!field definition in}

An interface may declare a \xcd`val` field, with a value.  This field is implicitly
\xcd`public static val`: 
%~~gen ^^^ Interfaces10
% package Interface.Field;
%~~vis
\begin{xten}
interface KnowsPi {
  PI = 3.14159265358;
}
\end{xten}
%~~siv
%
%~~neg

Classes and structs implementing such an interface get the interface's fields as
\xcd`public static` fields.  Unlike properties and methods, there is no need
for the implementing class to declare them. 
%~~gen ^^^ Interfaces20
% package Interface.Field.Two;
% interface KnowsPi {PI = 3.14159265358;}
%~~vis
\begin{xten}
class Circle implements KnowsPi {
  static def area(r:Double) = PI * r * r;
}
\end{xten}
%~~siv
%
%~~neg

\subsection{Fine Points of Fields}

It can happen that two parent interfaces give fields of the same name.  In
that case, those fields must be referred to by qualified names.
%~~gen ^^^ Interface_field_name_collision
% 
%~~vis
\begin{xten}
interface E1 {static val a = 1;}
interface E2 {static val a = 2;}
interface E3 extends E1, E2{}
class Example implements E3 {
  def example() = E1.a + E2.a;
}
\end{xten}
%~~siv
%
%~~neg

If the {\em same} field \xcd`a` is inherited through many paths, there is no need to
disambiguate it:
%~~gen ^^^ Interfaces_multi
% package Interfaces.Mult.Inher.Field;
%~~vis
\begin{xten}
interface I1 { static val a = 1;} 
interface I2 extends I1 {}
interface I3 extends I1 {}
interface I4 extends I2,I3 {}
class Example implements I4 {
  def example() = a;
}
\end{xten}
%~~siv
%
%~~neg



