\section{Implicit syntax}\label{ImplicitSyntax}\index{implicit syntax}

Recall that the explicit syntax for \Xten{} requires the programmer to use
asyncs/future to ensure the Locality Principle: An activity accesses
only those mutable locations that reside in the same place as the
activity. 

Explicit syntax has the advantage that the performance model for \Xten{}
is explicit from the syntax. It has the disadvantage that the
programmer has to manually reason about the placement of various
objects. If the programmer reasons incorrectly then computation may
abort at runtime with an exception.

The place-based type system enables the compiler to support the
Locality principle. The programmer uses the type system to establish
that the types of various objects are local. These assertions are
checked by the compiler (as a side-effect of checking dependent
types). Additionally, the programmer may now use normal variable
syntax to access (read/write) variables, and invoke methods on
objects. Suppose the type of the variable \xcd"v" is \xcd"C{c}".
If \xcd"c" establishes
\xcd"location==here" then the compiler generates code for
performing the relevant operation on the local variable (read,
write, method invocation) synchronously. 

Otherwise the compiler generates code in explicit syntax as
follows. If the operation is a read, the compiler generates code to
perform a future/force on the variable

\begin{xten}
future(v) { v }.force();  
\end{xten}

If the operation is a write \xcd"v=e", the compiler generates code to perform
the write synchronously:

\begin{xten}
val temp: T = e;
finish async (v) { v = w; }
\end{xten}

If the operation is a read on an array variable a[p] the compiler
generates the code:

\begin{xten}
future(a.dist(p)) { a(p) }.force();  
\end{xten}


If the operation is a write \xcd"a[p]=e", the compiler generates code to perform
the write synchronously:

\begin{xten}
val tp: point = p;
val t: T = e;
finish async (a.dist(tp)) { a(tp) = t; }
\end{xten}

If the operation is a method invocation \xcdmath"e.m(e$_1$,..., e$_n$)"
for a void method, the compiler generates code to
perform the method invocation synchronously:

\begin{xten}
val t: T! = e;
val t1: P1! = e1;
...
val tn: Pn! = en;
finish async (t) {
  t.m(t1,..., tn);
}
\end{xten}


If the operation is a method invocation
\xcdmath"e.m(e$_1$,...,e$_n$)" for a method that returns a value
of type E, then the compiler generates the following code:

XXX remove this -- do not define semantics by translation

\begin{xten}
new Runnable() {
  public E run() {
    val t: T! = e;
    val t1: P1! = e1;
    ...
    val tn: Pn! = en;
    return future(t){ t.m(t1,...tn) }.force();
  }}.run()
\end{xten}


