\chapter{Lexical and Grammatical structure}


Lexically a program consists of a stream of white space, comments,
identifiers, keywords, literals, separators and operators, all of them
composed of Unicode characters in the UTF-8 encoding. 

\paragraph{Whitespace}
\index{white space}
% Whitespace \index{whitespace} follows \java{} rules \cite[Chapter 3.6]{jls2}.
ASCII space, horizontal tab (HT), form feed (FF) and line
terminators constitute white space.

\paragraph{Comments}
\index{comment}
% Comments \index{comments} follows \java{} rules
% \cite[Chapter 3.7]{jls2}. 
All text included within the ASCII characters ``\xcd"/*"'' and
``\xcd"*/"'' is
considered a comment and ignored; nested comments are not
allowed.  All text from the ASCII characters
``\xcd"//"'' to the end of line is considered a comment and is ignored.

\paragraph{Identifiers}
\index{identifier}
\index{variable name}

Identifiers consist of a single letter followed by zero or more letters or
digits. The letters are the ASCII characters \xcd`a` through \xcd`z`, \xcd`A`
through \xcd`Z`, and \xcd`_`. Digits are defined as the ASCII characters
\xcd"0" through \xcd"9". Case is significant; \xcd`a` and \xcd`A` are distinct
identifiers, \xcd`as` is a keyword, but \xcd`As` and \xcd`AS` are identifiers.
(However, case is insignificant in the hexadecimal numbers, exponent markers,
and type-tags of numeric literals -- 0xbabe = 0XBABE.)

\paragraph{Keywords}
\index{keywords}
\Xten{} uses the following keywords:\\
{\tt 
\begin{tabular}{lll}
    {\bf abstract} &
    {\bf as}&
    {\bf assert}\\
    {\bf async}&
    {\bf at}&
    {\bf ateach}\\
    {\it atomic}&
    {\bf break}&
    {\bf case}\\
    {\bf catch}&
    {\bf class}&
    {\bf clocked}\\
    {\bf continue}&
    {\bf def}&
    {\bf default}\\
    {\bf do}&
    {\bf else}&
    {\bf extends}\\
    {\bf false}&
    {\bf final}&
    {\bf finally}\\
    {\bf finish}&
    {\bf for}&
    {\bf goto}\\
    {\bf haszero}&
    {\it here}&
    {\bf if}\\
    {\bf implements}&
    {\bf import}&
    {\bf in}\\
    {\bf instanceof}&
    {\bf interface}&
    {\bf native}\\
    {\bf new}&
    {\it next}&
    {\bf null}\\
    {\it offer}&
    {\bf offers}&
    {\bf operator}\\
    {\bf package}&
    {\bf private}&
    {\bf property}\\
    {\bf protected}&
    {\bf public}&
    {\it resume}\\
    {\bf return}&
    {\bf self}&
    {\bf static}\\
    {\bf struct}&
    {\bf super}&
    {\bf switch}\\
    {\bf this}&
    {\bf throw}&
    {\bf transient}\\
    {\bf true}&
    {\bf try}&
    {\bf type}\\
    {\bf val}&
    {\bf var}&
    {\bf void}\\
    {\bf when}& 
    {\bf while}&
\end{tabular}
}

Keywords in {\em italics} are {\em soft} keywords, which may also be
used as identifiers.  The other keywords, in {\bf bold}, are {\em hard}
keywords and  may not be used as identifiers.

Note that the primitive type names are not considered keywords.

\paragraph{Literals}\label{Literals}\index{literal}

Briefly, \XtenCurrVer{} uses fairly standard syntax for its literals:
integers, unsigned integers, floating point numbers, booleans, 
characters, strings, and \xcd"null".  The most exotic points are (1) unsigned
numbers are marked by a \xcd`u` and cannot have a sign; (2) \xcd`true` and
\xcd`false` are the literals for the booleans; and (3) floating point numbers
are \xcd`Double` unless marked with an \xcd`f` for \xcd`Float`. 

Less briefly, we use the following abbreviations: 
\begin{displaymath}
\begin{array}{rcll}
d &=& \mbox{one or more decimal digits}\\
d_8 &=& \mbox{one or more octal digits}\\
d_{16} &=& \mbox{one or more hexadecimal digits, using \xcd`a`-\xcd`f`
or \xcd`A`-\xcd`F`
for 10-15}\\
i &=& d 
        \mathbin{|} {\tt 0} d_8 
        \mathbin{|} {\tt 0x} d_{16}
        \mathbin{|} {\tt 0X} d_{16}
\\
s &=& \mbox{optional \xcd`+` or \xcd`-`}\\
b &=& d 
          \mathbin{|} d {\tt .}
          \mathbin{|} d {\tt .} d
          \mathbin{|}  {\tt .} d \\
x &=& ({\tt e } \mathbin{|} {\tt E})
         s
         d \\
f &=& b x
\end{array}
\end{displaymath}

\begin{itemize}

\item \xcd`true` and \xcd`false` are the \xcd`Boolean` literals. \index{Boolean!literal}\index{literal!Boolean}

\item \xcd`null` is a literal for the null value.  It has type
      \xcd`Any{self==null}`. \index{null} \index{object!literal}

\item \index{Int!literal}\index{literal!integer}
\xcd`Int` literals have the form {$si$}; \eg, \xcd`123`,
      \xcd`-321` are decimal \xcd`Int`s, \xcd`0123` and \xcd`-0321` are octal
      \xcd`Int`s, and \xcd`0x123`, \xcd`-0X321`,  \xcd`0xBED`, and \xcd`0XEBEC` are
      hexadecimal \xcd`Int`s.  

\item \xcd`Long` literals have the form {$si{\tt l}$} or
      {$si{\tt L}$}. \Eg, \xcd`1234567890L`  and \xcd`0xBABEL` are \xcd`Long` literals. 

\item \xcd`UInt` literals have the form {$i{\tt u}$} or {$i {\tt U}$}.
      \Eg, \xcd`123u`, \xcd`0123u`, and \xcd`0xBEAU` are \xcd`UInt` literals.

\item \xcd`ULong` literals have the form {$i {\tt ul}$} or {$i {\tt
      lu}$}, or capital versions of those.  For example, 
      \xcd`123ul`, \xcd`0124567012ul`,  \xcd`0xFLU`, \xcd`OXba1eful`, and \xcd`0xDecafC0ffeefUL` are \xcd`ULong`
      literals. 

\item \xcd`Short` literals have the form {$si{\tt s}$} or
      {$si{\tt S}$}. \Eg,  414S, \xcd`OxACES` and \xcd`7001s` are short
      literals. 

\item \xcd`UShort` literals  form {$i {\tt us}$} or {$i {\tt
      su}$}, or capital versions of those.  For example, \xcd`609US`, 
      \xcd`107us`, and \xcd`OxBeaus` are unsigned short literals.

\item \xcd`Byte` literals have the form  {$si{\tt y}$} or
      {$si{\tt Y}$}.  (The letter \xcd`B` cannot be used for bytes, as it is
      a hexadecimal digit.)  \xcd`50Y` and \xcd`OxBABY` are byte literals.

\item \xcd`UByte` literals have the form {$i {\tt uy}$} or {$i {\tt yu}$}, or
      capitalized versions of those.  For example, \xcd`9uy` and \xcd`OxBUY`
      are \xcd`UByte` literals.
      


\item \xcd`Float` literals have the form {$s f {\tt f}$} or  {$s
\index{float!literal}
\index{literal!float}
      f {\tt F}$}.  Note that the floating-point marker letter \xcd`f` is
      required: unmarked floating-point-looking literals are \xcd`Double`. 
      \Eg, \xcd`1f`, \xcd`6.023E+32f`, \xcd`6.626068E-34F` are \xcd`Float`
      literals. 

\item \xcd`Double` literals have the form {$s f$}\footnote{Except that
\index{double!literal}
\index{literal!double}
      literals like \xcd`1` 
      which match both {$i$} and {$f$} are counted as
      integers, not \xcd`Double`; \xcd`Double`s require a decimal
      point, an exponent, or the \xcd`d` marker.
      }, {$s f {\tt
      D}$}, and {$s f {\tt d}$}.  
      \Eg, \xcd`0.0`, \xcd`0e100`, \xcd`1.3D`,  \xcd`229792458d`, and \xcd`314159265e-8`
      are \xcd`Double` literals.

\item 
\index{char!literal}
\index{literal!char}
\xcd`Char` literals have one of the following forms: 
      \begin{itemize}
      \item \xcd`'`{\it c}\xcd`'` where {\em c} is any printing ASCII
            character other than 
            \xcd`\` or \xcd`'`, representing the character {\em c} itself; 
            \eg, \xcd`'!'`;
      \item \xcd`'\b'`, representing backspace;
      \item \xcd`'\t'`, representing tab;
      \item \xcd`'\n'`, representing newline;
      \item \xcd`'\f'`, representing form feed;
      \item \xcd`'\r'`, representing return;
      \item \xcd`'\''`, representing single-quote;
      \item \xcd`'\"'`, representing double-quote;
      \item \xcd`'\\'`, representing backslash;
      \item \xcd`'\`{\em dd}\xcd`'`, where {\em dd} is one or more octal
            digits, representing the one-byte character numbered {\em dd}; it
            is an error if {\em dd}{$>0377$}.      
      \end{itemize}

\item
\index{string!literal} 
\index{literal!string}
\xcd`String` literals consist of a double-quote \xcd`"`, followed by
      zero or more of the contents of a \xcd`Char` literal, followed by
      another double quote.  \Eg, \xcd`"hi!"`, \xcd`""`.


\end{itemize}



\paragraph{Separators}
\Xten{} has the following separators and delimiters:
\begin{xten}
( )  { }  [ ]  ;  ,  .
\end{xten}

\paragraph{Operators}
\index{operator}
\Xten{} has the following operator,  type constructor, and miscellaneous symbols.  (\xcd`?` and
\xcd`:` comprise a single ternary operator, but are written separately.)
\begin{xten}
==  !=  <   >   <=  >=
&&  ||  &   |   ^
<<  >>  >>>
+   -   *   /   %
++  --  !   ~
&=  |=  ^=
<<= >>= >>>=
+=  -=  *=  /=  %=
=   ?   :  =>  ->
<:  :>  @   ..
\end{xten}

The precedence of the operators is as follows.  Earlier rows of the
table have higher precedence than later rows, binding more tightly.  
For example, \xcd`a+b*c<d` parses as \xcd`(a+(b*c))<d`, 
and \xcd`-1 as Byte` parses as \xcd`-(1 as Byte)`. 

\begin{tabular}{l}
\xcd`as T`, postfix \xcd`++`, postfix \xcd`--`\\
unary \xcd`-`, unary \xcd`+`, prefix \xcd`++`, prefix \xcd`--`\\
\xcd`!`, \xcd`~`\\
\xcd`? :` \\
\xcd`||` \\
\xcd`&&` \\
\xcd`|` \\
\xcd`^` \\
\xcd`&` \\
\xcd`in` \\
\xcd`==    `   \xcd`!=` \\
\xcd`instanceof` \\
\xcd`>     ` \xcd`>=    ` \xcd`<    ` \xcd`<=` \\
\xcd`<<    ` \xcd`>>    ` \xcd`>>>   ` \xcd`->` \\
\xcd`+` \xcd`     -` \\
\xcd`*     `  \xcd`/     `  \xcd`%` \\
\xcd`..`\\
\end{tabular}


\section{Grammatical Notation}

In this manual, ordinary BNF notation is used to specify grammatical
constructions, with a few minor extensions. 
Grammatical rules look like this: \\
\begin{bbgrammar}
Adj \: Adv\opt \mbox{\xcd`happy`}\\
    \| Adv\opt \mbox{\xcd`sad`}\\ 
Adv \: \mbox{\xcd`very`}\\
    \| Adv Adv 
\end{bbgrammar}

Terms in {\em italics} are called {\bf non-terminals}.  They 
represent kinds of phrases; for example, {\em ForStatement}
(\ref{prod:ForStatement}) describes all \xcd`for` statements.  
Equation numbers refer to the full X10 grammar, in \Sref{Grammar}.  
The small example has two non-terminals, {\em Adv} and {\em Adj}.  

Terms in \xcd`fixed-width font` are {\bf terminals}.  They represent the words
and symbols of the language itself.  In X10, the terminals are the words
described in this chapter.

A single grammatical rule has the form {$ A \mathrel{::=} X_1 X_2 ... X_n $},
where the $X_i$'s are either terminals or nonterminals.  This indicates that
the non-terminal $A$ could be an instance of $X_1$, followed by an instance of
$X_2$, \ldots, followed by an instance of $X_n$.  Multiple rules for the same
$A$ are allowed, giving several possible phrasings of $A$'s.  For brevity,
two rules with the same left-hand side are written with the left-hand side
appearing once, and the right-hand sides 
separated by $|$.  

In the {\em Adj} example, there are two rules for {\it Adv}, 
$Adv \mathrel{::=} \mbox{\xcd`very`}$ and
$Adv \mathrel{::=} Adv~Adv$.  So, an adverb could be \xcd`very`, or (by three
uses of the rule) \xcd`very very`, or, one or more \xcd`very`s.  

The notation $A^{?}$ indicates an optional $A$.  This is an ordinary
non-terminal, defined by the rules:\\ 
\begin{bbgrammar}
A\opt \:  \\
    \| A
\end{bbgrammar}
The first rule says that $A^?$ can amount to nothing; the second, that it can
amount to an $A$.  This concept shows up so often that it is worth having a
separate notation for it.  In the {\em Adj} example, an adjective phrase may be
preceded by an optional adverb. Thus, it may be \xcd`happy`, or 
\xcd`very happy`, or \xcd`very very sad`, etc. 
