\chapter{Types}
\label{XtenTypes}\index{types}

{}\Xten{} is a {\em strongly typed} object-oriented language: every
variable and expression has a type that is known at compile-time.
Types limit the values that variables can hold.

{}\Xten{} supports three kinds of runtime entities, {\em objects},
{\em structs}, and {\em functions}. Objects are instances of {\em
  classes} (\Sref{ReferenceClasses}). They may contain zero or
more mutable fields, and a reference to the list of methods defined on them.

An object is represented by some (contiguous) memory chunk on the
heap. Entities (such as variables and fields) contain a {\em
  reference} to this chunk. That is, objects are represented through
an extra level of indirection.  A consequence of this flexibility is
that an entity containing a reference to an object \xcd{o} needs only
one word of memory to represent that reference, regardess of the
number of fields in \xcd{o}. An assignment to this entity simply
overwrites the reference with another reference (thus taking constant
time). Another consequence is that every class type contains the value
\Xcd{null} corresponding to the invalid reference. \Xcd{null} is often
useful as a default value. Further, two objects may be compared for
identity (\Xcd{==}) in constant time by simply comparing references to
the memory used to represent the objects. The default hash code for an
object is based on the value of this reference. A downside of this
flexibility is that the operations of accessing a field and invoking a
method are more expensive than simply reading a register and
invoking a static function.


Structs are instances of {\em struct types} (\Sref{StructClasses}).  A
struct is represented without the extra level of indirection, with a
memory chunk of size $N$ words precisely big enough to store the value
of every field of the struct (modulo alignment), plus whatever padding is needed. Thus structs cannot
be shared. Entities (such as variables and fields) refering to the
struct must allocate $N$ words to directly contain the chunk.  An
assignment to this entity must copy the $N$ words representing the
right hand side into the left hand side. Since there are no references
to structs, \Xcd{null} is not a legal value for a struct
type. Comparison for identity (\Xcd{==}) involves examining $N$
words. Additionally, structs do not have any mutable fields, hence
they can be freely copied. The payoff for these restrictions lies in
that fields can be stored in registers or local variables, and 
and method invocation is implemented by invoking a static function.

Functions, called closures or lambda-expressions in other languages, are
instances of {\em function types} (\Sref{Functions}). 
A function has zero or more {\em formal parameters} (or {\em arguments}) and a
{\em body}, which is 
an expression that can reference the formal parameters and also other
variables in the surrounding block. For instance, \xcd`(x:Int)=>x*y`
is a unary integer function which multiplies its argument by the
variable \xcd`y` from the surrounding block.  Functions may be freely
copied from place to place and may be repeatedly applied. 

These runtime entities are classified by {\em types}. Types are used in
variable declarations, coercions and  explicit conversions, object creation,
array creation, static state and method accessors, and
\xcd"instanceof" and \xcd`as` expressions.


The basic relationship between values and types is the {\em is an
element of} relation.  We also often say ``$e$ has type $T$'' to
mean ``$e$ is an element of type $T$''.  For example, \xcd`1` has type
\xcd`Int` (the type of all integers representible in 32 bits). It also
has type \xcd`Any` (since all entitites have type \xcd`Any`), type
\xcd`Int{self != 0}` (the type of nonzero integers), type
\xcd`Int{self == 1}` (the type of integers which are equal to \xcd`1`, which
contains only one element), and many others. 

The basic relationship between types is {\em subtyping}: \xcd`T <: U`
holds if every instance of \xcd`T` is also an instance of \xcd`U`. Two
important kinds of subtyping are {\em subclassing} and {\em
  strengthening}. Subclassing is a familiar notion from
object-oriented programming. Here we use it to refer to the
relationship between a class and another class it extends, and the
relationship between a class and another interface it implements. For
instance, in a class hierarchy with classes \xcd`Animal` and \xcd`Cat`
such that \xcd`Cat` extends \xcd`Mammal` and \xcd`Mammal` extends
\xcd`Animal`, every instance of \xcd`Cat` is by definition an instance
of \xcd`Animal` (and \xcd`Mammal`). We say that \xcd`Cat` is a
subclass of \xcd`Animal`, or \xcd`Cat <: Animal` by subclassing. If
\xcd`Animal` implements \xcd`Thing`, then \xcd`Cat` also implements
\xcd`Thing`, and we say \xcd`Cat <: Thing` by subclassing.
Strengthening is an equally familiar notion from logic.  The instances
of \xcd`Int{self == 1}` are all elements of \xcd`Int{self != 0}` as well,
because \xcd`self == 1` logically implies \xcd`self != 0`; so 
\xcd`Int{self  == 1} <: Int{self !=0}` by strengthening.  X10 uses both notions
of subtyping.  See \Sref{DepType:Equivalence} for the full definition
of subtyping in X10.

\subsection{Type System}
\index{type system}
The types in X10 are as follows.  

These are the {\em elementary} types. Other
syntactic forms for types exist, but they are simply abbreviations for types
in the following system.  For example, \xcd`Array[Int](1)` is the type of
one-dimensional integer-valued arrays; it is an abbreviation for
\xcd`Array[Int]{rank==1}`.\\

% remove \refstepcounter{equation}
% snag the argument of \label{X}
% change the (\arabic{equation}) into (\ref{X})

%##(Type FunctionType ConstrainedType
\begin{bbgrammar}
%(FROM #(prod:Type)#)
                Type \: FunctionType & (\ref{prod:Type}) \\
                    \| ConstrainedType \\
%(FROM #(prod:FunctionType)#)
        FunctionType \: TypeParams\opt \xcd"(" FormalParamList\opt \xcd")" WhereClause\opt Offers\opt \xcd"=>" Type & (\ref{prod:FunctionType}) \\
%(FROM #(prod:ConstrainedType)#)
     ConstrainedType \: NamedType & (\ref{prod:ConstrainedType}) \\
                    \| AnnotatedType \\
                    \| \xcd"(" Type \xcd")" \\
\end{bbgrammar}
%##)


Types may be given by name. 
For example, 
%~~type~~`~~`~~ ~~ ^^^ Types10
\xcd`Int`
is the type of 32-bit integers.
Given a class declaration 
%~~gen ^^^ Types20
%package Types.Core.TypeName; 
%~~vis
\begin{xten}
class Triple { /* ... */ }
\end{xten}
%~~siv
%
%~~neg
the identifier \xcd`Triple` may be used as a type.

The type {\em TypeName \xcd`[` Types{$^?$} \xcd`]`} is an instance of
a {\em generic} (or {\em parameterized}) type. 
 For example,
\xcd`Array[Int]` is the type of arrays of integers. 
\xcd`HashMap[String,Int]` is the type of hash maps from strings to
integers.

The type {\em Type \xcd`{` Constraint \xcd`}`} refers to a constrained type.
{\em Constraint} is a Boolean expression -- written in a {\em very} limited
subset of X10 -- describing the acceptable values of the constrained type.
%~~stmt~~`~~`~~ ~~ ^^^ Types30
For example, \xcd`var n : Int{self != 0};` guarantees that \xcd`n` is always a
non-zero integer. 
%~~stmt~~`~~`~~ ~~class Triple{} ^^^ Types40
Similarly, \xcd`var x : Triple{x != null};` defines a \xcd`Triple`-valued
variable \xcd`x` whose value is never null.

The qualified type {\em Type \xcd`.` Type} refers to an instance of a {\em
nested} type; that is, a class or struct defined inside of another class or
struct, and holding an implicit reference to the outer.  For example, given
the type declaration 
%~~gen ^^^ Types50
% package Types.Core.Hardcore.Qualified;
%~~vis
\begin{xten}
class Outer {
  class Inner { /* ... */ }
}
\end{xten}
%~~siv
%
%~~neg
then 
%~~exp~~`~~`~~ ~~ NOTEST class Outer {class Inner { /* ... */ }} ^^^ Types60
\xcd`(new Outer()).new Inner()` creates a value of type 
%~~type~~`~~`~~ ~~class Outer {class Inner { /* ... */ }} ^^^ Types70
\xcd`Outer.Inner`.

Type variables, {\em TypeVar}, refer to types that are parameters.  For
example, the following class defines a cell in a linked list.  
%~~gen ^^^ Types80
% package Types.Core.Bore.Lore;
%~~vis
\begin{xten}
class LinkedList[X] {
  val head : X;
  val tail : LinkedList[X];
  def this(head:X, tail:LinkedList[X]) {
     this.head = head; this.tail = tail;
  }
}
\end{xten}
%~~siv
%
%~~neg
It doesn't
matter what type the cell is, but it has to have {\em some} type.
\xcd`LinkedList[Int]` is a linked list of integers;
\xcd`LinkedList[LinkedList[String]]` a list of lists of strings.
Note that \xcd`LinkedList` is {\em not} a type -- it is missing a type parameter.



The function type 
{\em \xcd`(` Formals{$^?$} \xcd`) =>`  Type} 
refers to functions taking the
listed formal parameters and returning a result of {\em Type}.  In
\XtenCurrVer, function types may not be generic.
The closely-related void function type 
{\em \xcd`(` Formals{$^?$} \xcd`) =>`  \xcd`void`}  takes the listed
parameters and returns no value.
For example, 
%~~type~~`~~`~~ ~~ ^^^ Types90
\xcd`(x:Int) => Int{self != x}` 
is the type of integer-valued functions which have no fixed points.  
An example of such a function is \xcd`(x:Int) => x+1`.





\section{Classes, Structs,  and interfaces}
\label{ReferenceTypes}

\subsection{Class types}

\index{type!class}
\index{class}
\index{class declaration}
\index{declaration!class declaration}
\index{declaration!reference class declaration}

A {\em class declaration} (\Sref{XtenClasses}) declares a {\em class type},
giving its name, behavior, data, and relationships to other classes and
interfaces. 

\begin{ex}

The \xcd`Position` class below could describe the position of a slider
control

%~~gen ^^^ Types100
% package Types.By.Cripes.Classes;
%~~vis
\begin{xten}
class Position {
  private var x : Int = 0;
  public def move(dx:Int) { x += dx; }
  public def pos() : Int = x;
}
\end{xten}
%~~siv
%
%~~neg
\end{ex}
Class instances, also called objects, are created by constructor calls: 
\xcd`new Position()`. Class
instances have fields and methods, type members, and value properties bound at
construction time. In addition, classes have static members: static \xcd`val` fields,
methods, type definitions, and member classes and member interfaces.

Classes may be {\em generic}, \ie, defined with one or more type
parameters (\Sref{TypeParameters}).  

%~~gen ^^^ Types110
%~~vis
\begin{xten}
class Cell[T] {
  var contents : T;
  public def this(t:T) { contents = t;  }
  public def putIn(t:T) { contents = t; }
  public def get() = contents;
  }
\end{xten}
%~~siv
%~~neg


%TODO: Yoav: ``This reasoning is no longer true in the new object model''
%% Why not?
\Xten{} does not permit mutable static state. A fundamental principle of the
X10 model of computation is that all mutable state be local to some place
(\Sref{XtenPlaces}), and, as static variables are
globally available, they
cannot be mutable. When mutable global state is necessary, programmers should
use singleton classes, putting the state in an object and using place-shifting
commands (\Sref{AtStatement}) and atomicity (\Sref{AtomicBlocks}) as necessary
to mutate it safely.

\index{\Xcd{Object}}
\index{\Xcd{x10.lang.Object}}

Classes are structured in a single-inheritance hierarchy. All classes extend
the class \xcd"x10.lang.Object", directly or indirectly. Each class other than
\xcd`Object` extends a single parent class.  \xcd`Object` provides no behaviors
of its own, beyond those required by \xcd`Any`.

\index{class!reference class}
\index{reference class type}
\index{\Xcd{Object}}
\index{\Xcd{x10.lang.Object}}


\index{null}


The null value, represented by the literal
\xcd"null", is a value of every class type \xcd`C`. The type whose values are
all instances of \xcd`C` but not 
\xcd`null` can be defined as \xcd`C{self != null}`.

\subsection{Struct Types}

A {\em struct declaration} \Sref{XtenStructs} introduces a {\em struct type}
containing all instances of the struct.  The \xcd`Coords` struct below gives
an immutable position in 3-space: 
%~~gen ^^^ Types120
% package Types.Structs.Coords;
%~~vis
\begin{xten}
struct Position {
  public val x:Double, y:Double, z:Double; 
  def this(x:Double, y:Double, z:Double) {
     this.x = x; this.y = y; this.z = z;
  }
}
\end{xten}
%~~siv
%
%~~neg

Structs have many capabilities of classes: they can have methods, implement
interfaces, and be generic. However, they have certain restrictions; for
example, they cannot contain mutable (\xcd`val`) fields, or inherit from
superclasses. There is no \xcd`null` value for structs. Due to these
restrictions, structs may be implemented more efficiently than objects.


\subsection{Interface types}
\label{InterfaceTypes}

\index{type!interface}
\index{interface}
\index{interface declaration}
\index{declaration!interface declaration}

An {\em interface declaration} (\Sref{XtenInterfaces}) defines an {\em
interface type}, specifying a set of methods 
%type members, 
and properties which must be provided by any class declared to implement the
interface. 


Interfaces can also have static members: static fields, type
definitions, and member classes, structs and interfaces.  However,
interfaces cannot specify that implementing classes must provide
static members or constructors.

\begin{ex}
In the following interface, \xcd`PI` is a static field, 
\xcd`Vec` a static type definition, 
\xcd`Pair` a static member class.
It can't insist that implementations provide a static method 
like \xcd`meth`, or a nullary constructor.
%~~gen ^^^ Types2y3i
% NOTEST
% package Types2y3i;
%~~vis
\begin{xten}
interface Stat {
  static val PI = 3.14159; 
  static type R = Double;
  static class Pair(x:R, y:R) {}
  // ERROR: static def meth():Int;
  // ERROR: static def this();
}
class Example {
  static def example() {
     val p : Stat.Pair = new Stat.Pair(Stat.PI, Stat.PI);
  }
}
\end{xten}
%~~siv
%
%~~neg

\end{ex}

An interface may extend multiple interfaces.  
%~~gen ^^^ Types130
%package Types.For.Snipes.Interfaces;
%~~vis
\begin{xten}
interface Named {
  def name():String;
}
interface Mobile {
  def move(howFar:Int):void;
}
interface Person extends Named, Mobile {}
interface NamedPoint extends Named, Mobile {} 
\end{xten}
%~~siv
%
%~~neg


Classes and structs may be declared to implement multiple interfaces. Semantically, the
interface type is the set of all objects that are instances of classes
or structs that
implement the interface. A class or struct implements an interface if it is declared to
and if it concretely or abstractly implements all the methods and properties
defined in the interface. For example, \xcd`KimThePoint` implements
\xcd`Person`, and hence \xcd`Named` and \xcd`Mobile`. It would be a static
error if \xcd`KimThePoint` had no \xcd`name` method, unless \xcd`KimThePoint` were also
declared \xcd`abstract`.

%~~gen ^^^ Types140
%interface Named {
%   def name():String;
% }
% interface Mobile {
%   def move(howFar:Int):void;
% }
% interface Person extends Named, Mobile {}
% interface NamedPoint extends Named, Mobile{} 
%~~vis
\begin{xten}
class KimThePoint implements Person {
   var pos : Int = 0;
   public def name() = "Kim (" + pos + ")";
   public def move(dPos:Int) { pos += dPos; }
}
\end{xten}
%~~siv
%
%~~neg


\subsection{Properties}
\index{property}
\label{properties}

Classes, interfaces, and structs may have {\em properties}, specified in
parentheses after the type name. Properties are much like public \xcd`val`
instance fields. They have certain restrictions on their use, however, which
allows the compiler to understand them much better than other public \xcd`val`
fields. In particular, they can be used in types.  \Eg, the number of elements
in an array is a property of the array, and an X10 program can specify that
two arrays have the same number of elements.

\begin{ex}
The
following code declares a class named \xcd"Coords" with properties
\xcd"x" and \xcd"y" and a \xcd"move" method. The properties are bound
using the \xcd"property" statement in the constructor.

%~~gen ^^^ Types150
%package not.x10.lang;
%~~vis
\begin{xten}
class Coords(x: Int, y: Int) { 
  def this(x: Int, y: Int) :
    Coords{self.x==x, self.y==y} = { 
    property(x, y); 
  } 

  def move(dx: Int, dy: Int) = new Coords(x+dx, y+dy); 
}
\end{xten}
%~~siv
%~~neg
\end{ex}
Properties, unlike other public \xcd`val` fields, can be used  
at compile time in {constraints}. This allows us
to specify subtypes based on properties, by appending a boolean expression to
the type. For example, the type \xcd"Coords{x==0}" is the set of all points
whose \xcd"x" property is \xcd"0".  Details of this substantial topic are
found in \Sref{ConstrainedTypes}.



\section{Type Parameters and Generic Types}
\label{TypeParameters}

\index{type!parameter}
\index{method!parametrized}
\index{constructor!parametrized}
\index{closure!parametrized}
\label{Generics}
\index{type!generic}

A class, interface, method, or type definition  may have type
parameters.  Type parameters can be used as types, and will be bound to types
on instantiation.  For example, a generic stack class may be defined as 
\xcd`Stack[T]{...}`.  Stacks can hold values of any type; \eg, 
%~~type~~`~~`~~ ~~class Stack[T]{} ^^^ Types160
\xcd`Stack[Int]` is a stack of integers, and 
%~~type~~`~~`~~ ~~class Stack[T]{} ^^^ Types170
\xcd`Stack[Point {self!=null}]` is a stack of non-null \xcd`Point`s.
Generics {\em must} be instantiated when they are used: \xcd`Stack`, by
itself, is not a valid type.
Type parameters may be constrained by a guard on the declaration
(\Sref{TypeDefGuard},
\Sref{MethodGuard},\Sref{ClosureGuard}).

\index{type!concrete}
\index{concrete type}
A {\em generic class} (or struct, interface, or type definition) 
is a class (resp. struct, interface, or type definition) 
declared with $k \geq 1$ type parameters. 
A generic class (or struct, interface, or type definition) 
can be used to form a type by supplying $k$ types as type arguments within
\xcd`[` \ldots \xcd`]`.
%%When instantiated,
%%with concrete (\viz, non-generic) types for its parameters, 
%%a generic type becomes a concrete type and can be
%%used like any other type. 
For example,
\xcd`Stack` is a generic class, 
%~~type~~`~~`~~ ~~class Stack[T]{} ^^^ Types180
\xcd`Stack[Int]` is a type, and can be used as one: 
%~~stmt~~`~~`~~ ~~class Stack[T]{} ^^^ Types190
\xcd`var stack : Stack[Int];`

A \xcd`Cell[T]` is a generic object, capable of holding a value of type
\xcd`T`.  For example, a \xcd`Cell[Int]` can hold an \xcd`Int`, and a
\xcd`Cell[Cell[Int{self!=0}]]` can hold a \xcd`Cell` which in turn can
only hold non-zero numbers. 
%% vj: Dont know what this saying: bound immutably... but mutable?
%% \xcd`Cell`s are actually useful in situations
%%where values must be bound immutably for one reason, but need to be mutable.
%~~gen ^^^ Types200
% package ch4;
%~~vis
\begin{xten}
class Cell[T] {
    var x: T;
    def this(x: T) { this.x = x; }
    def get(): T = x;
    def set(x: T) = { this.x = x; }
}
\end{xten}
%~~siv
%~~neg


\xcd"Cell[Int]" is the type of \xcd`Int`-holding cells.  
The \xcd"get" method on a \xcd`Cell[Int]` returns an \xcd"Int"; the
\xcd"set" method takes an \xcd"Int" as argument.  Note that
\xcd"Cell" alone is not a legal type because the parameter is
not bound.


A class (whether generic or not) may have generic methods.
Below,
\xcd`NonGeneric` has a generic method 
\xcd`first[T](x:List[T])`. An invocation of such a method may supply
the type parameters explicitly (\eg, \xcd`first[Int](z)`).
 In certain cases (\eg, \xcd`first(z)`)
type parameters may
be omitted and are inferred by the compiler (\Sref{TypeInference}).

%~~gen ^^^ Types210
% package Types.For.Cripes.Sake.Generic.Methods;
% import x10.util.*;
%~~vis
\begin{xten}
class NonGeneric {
  static def first[T](x:List[T]):T = x(0);
  def m(z:List[Int]) {
    val f = first[Int](z);
    val g = first(z);
    return f == g;
  }
}
\end{xten}
%~~siv
%
%~~neg
\limitation{ \XtenCurrVer{}'s C++ back end requires generic methods to be
static or final; the Java back end can accomodate generic instance methods as well. }

Unlike other kinds of variables, type parameters may {\em not} be shadowed.  
If name \xcd`X` is in scope as a type, \xcd`X` may not be rebound as a type
variable.  
For example, neither \xcd`class B` nor \xcd`class C[B]` are allowed in the
following code, because they both shadow the type variable \xcd`B`.
%~~gen ^^^ TypesNoShadow
% package TypesNoShadow;
%~~vis
\begin{xten}
class A[B] {
  // ILLEGAL: class B{} 
  // ILLEGAL: class C[B]{} 
}
\end{xten}
%~~siv
%
%~~neg


\subsection{Use of Generics}

An unconstrained type variable \Xcd{X} can be instantiated any type. Within a
generic struct or class, all the operations of \Xcd{Any} are available on a
variable of type unconstrained \Xcd{X}. Additionally, variables of type
\Xcd{X} may be used with \Xcd{==, !=}, in \Xcd{instanceof}, and casts.  

If a type variable is constrained, the operations implied by its constraint
are available as well.

\begin{ex}
The interface \xcd`Named` describes entities which know their own name.  The
class \xcd`NameMap[T]` is a specialized map which stores and retrieves
\xcd`Named` entities by name.  The call \xcd`t.name()` in \xcd`put()` is only
valid because the constraint \xcd`{T <: Named}` implies that \xcd`T` is a
subtype of \xcd`Named`, and hence provides all the operations of \xcd`Named`. 
%~~gen ^^^ Types6e6x
% package Types6e6x;
% import x10.util.*;
%~~vis
\begin{xten}
interface Named { def name():String; }
class NameMap[T]{T <: Named} {
   val m = new HashMap[String, T]();
   def put(t:T) { m.put(t.name(), t); }
   def get(s:String):T = m.getOrThrow(s);
}
\end{xten}
%~~siv
%
%~~neg


\end{ex}


%%NO-VARIANCE%% \subsection{Variance of Type Parameters}
%%NO-VARIANCE%% \index{covariant}
%%NO-VARIANCE%% \index{contravariant}
%%NO-VARIANCE%% \index{invariant}
%%NO-VARIANCE%% \index{type parameter!covariant}
%%NO-VARIANCE%% \index{type parameter!contravariant}
%%NO-VARIANCE%% \index{type parameter!invariant}
%%NO-VARIANCE%% 
%%NO-VARIANCE%% % Uncomment this when the language implementation properly supports variance.
%%NO-VARIANCE%% %\input{Variance}
%%NO-VARIANCE%% 
%%NO-VARIANCE%% Class, struct and interface definitions are permitted to specify a {\em
%%NO-VARIANCE%%   variance} 
%%NO-VARIANCE%% for each type parameter. 
%%NO-VARIANCE%% There are three variance specifications: 
%%NO-VARIANCE%% \xcd`+` indicates {\em co-variance},  \xcd`-` indicates {\em
%%NO-VARIANCE%%   contravariance} and the absence of  \xcd`+` and 
%%NO-VARIANCE%%  \xcd`-` indicates {\em invariance}. For a class (or struct or
%%NO-VARIANCE%%  interface) \xcd`S` specifying that a particular parameter position
%%NO-VARIANCE%%  (say, \xcd`i`) is covariant means that 
%%NO-VARIANCE%% if \xcd`Si <: Ti` then
%%NO-VARIANCE%% \xcdmath"S[S1,$\ldots$,Sn] <: S[S1,$\ldots$, Si-1,Ti,Si+1,$\ldots$ Sn]".
%%NO-VARIANCE%% Similarly, saying that position \xcd`i` is is contravariant means
%%NO-VARIANCE%% that 
%%NO-VARIANCE%% if \xcd`Si <: Ti` then
%%NO-VARIANCE%% \xcdmath"S[S1,$\ldots$, Si-1,Ti,Si+1,$\ldots$ Sn] <: S[S1,$\ldots$,Sn]". If the
%%NO-VARIANCE%% position is invariant, then no such relationship is asserted between
%%NO-VARIANCE%% \xcd`Si <: Ti` 
%%NO-VARIANCE%% and
%%NO-VARIANCE%% \xcdmath"S[S1,$\ldots$, Si-1,Ti,Si+1,$\ldots$ Sn]". The compiler must perform
%%NO-VARIANCE%% several checks on the body of the class (or struct or interface) to
%%NO-VARIANCE%% establish that type parameters with a variance are used in a manner
%%NO-VARIANCE%% that is consistent with their semantics.
%%NO-VARIANCE%% 
%%NO-VARIANCE%% \limitation{} The implementation of variance specifications  suffers from
%%NO-VARIANCE%% various limitations in \XtenCurrVer. Users are strongly encouraged not
%%NO-VARIANCE%% to use variance. (Some classes, structs, and interfaces in the standard
%%NO-VARIANCE%% libraries use variance specifications in a careful way that
%%NO-VARIANCE%% circumvents these limitations.)
%%NO-VARIANCE%% 

\section{Type definitions}
\label{TypeDefs}

\index{type!definitions}
\index{declaration!type}
A type definition can be thought of as a type-valued function,
mapping type parameters and value parameters to a concrete type.

%##(TypeDefDecl TypeParams FormalParams WhereClause
\begin{bbgrammar}
%(FROM #(prod:TypeDefDecl)#)
         TypeDefDecl \: Mods\opt \xcd"type" Id TypeParams\opt FormalParams\opt WhereClause\opt \xcd"=" Type \xcd";" & (\ref{prod:TypeDefDecl}) \\
%(FROM #(prod:TypeParams)#)
          TypeParams \: \xcd"[" TypeParamList \xcd"]" & (\ref{prod:TypeParams}) \\
%(FROM #(prod:FormalParams)#)
        FormalParams \: \xcd"(" FormalParamList\opt \xcd")" & (\ref{prod:FormalParams}) \\
%(FROM #(prod:WhereClause)#)
         WhereClause \: DepParams & (\ref{prod:WhereClause}) \\
\end{bbgrammar}
%##)

\noindent 
During type-checking the compiler replaces the use of such a defined
type with its body, substituting the actual type and value parameters
in the call for the formals. This replacement is performed recursively
until the type no longer contains a defined type or a predetermined
compiler limit is reached (in which case the compiler declares an
error). Thus, recursive type definitions are not permitted.

Thus type definitions are considered applicative and not generative --
they do not define new types, only aliases for existing types.

\label{TypeDefGuard}
Type definitions may have guards: an invocation of a type definition
is illegal unless the guard is satisified when formal types and values
are replaced by the actual parameters.

Type definitions may be overloaded: two type definitions with
the same name are permitted provided that they have a different number
of type parameters or different number or type of value parameters.

Type definitions must appear as static members or in a block statement.

\paragraph{Use of type definitions in constructor invocations}
If a type definition has no type parameters and no value
parameters and is an alias for a class type, a \xcd"new"
expression may be used to create an instance of the class using
the type definition's name.
Given the following type definition:
%TODO: Yoav says ``I just opened a jira on it: [1918].  I don't think you
% should be able to have {c} on the typedef A if you want to use it in a 'new'
% expression. If we do allow it, then we should allow: new
% Array[Int]{rank==1}(0..2) and new Array[Int](1)(0..2).
\begin{xtenmath}
type A = C[T$_1$, $\dots$, T$_k$]{c};
\end{xtenmath}
where 
\xcdmath"C[T$_1$, $\dots$, T$_k$]" is a
class type, a constructor of \xcdmath"C" may be invoked with
\xcdmath"new A(e$_1$, $\dots$, e$_n$)", if the
invocation
\xcdmath"new C[T$_1$, $\dots$, T$_k$](e$_1$, $\dots$, e$_n$)" is
legal and if the constructor return type is a subtype of
\xcd"A".

\paragraph{Automatically imported type definitions}
\index{import,type definitions}
\label{X10LangUnderscore}

The collection of type definitions in
\xcdmath"x10.lang._" is automatically imported in every compilation unit.


\subsection{Motivation and use}
The primary purpose of type definitions is to provide a succinct,
meaningful name for complex types
and combinations of types. 
With value arguments, type arguments, and constraints, the syntax for \Xten{}
types can often be verbose. 
For example, a non-null list of non-null strings is \\
%~~type~~`~~`~~ ~~import x10.util.*; ^^^ Types220
\xcd`List[String{self!=null}]{self!=null}`.

We could name that type: 
%~~gen ^^^ Types230
% package TypeDefs.glip.first;
% import x10.util.*;
% class LnSn {
% 
%~~vis
\begin{xten}
static type LnSn = List[String{self!=null}]{self!=null};
\end{xten}
%~~siv
%}
%~~neg
Or, we could abstract it somewhat, defining a type constructor
\xcd`Nonnull[T]` for the type of \xcd`T`'s which are not null:
%~~gen ^^^ Types240
% package TypeDefs.glip.second;
% import x10.util.*;
% 
%~~vis
\begin{xten}
class Example {
  static type Nonnull[T]{T <: Object}  = T{self!=null};
  var example : Nonnull[Example] = new Example();
}
\end{xten}
%~~siv
%
%~~neg

Type definitions can also refer to values, in particular, inside 
constraints.  The type of \xcd`n`-element \xcd`Array[Int](1)`s  is 
%~~type~~`~~`~~n:Int ~~ ^^^ Types250
\xcd`Array[Int]{self.rank==1 && self.size == n}`
but it is often convenient to give a shorter name: 
%~~gen ^^^ Types260
% package TypeDefs.glip.third;
% class Xmpl {
% def example() {
%~~vis
\begin{xten}
type Vec(n:Int) = Array[Int]{self.rank==1 && self.size == n}; 
var example : Vec(78); 
\end{xten}
%~~siv
%}}
%~~neg

%
The following examples are legal type definitions, given \xcd`import x10.util.*`:
%~~gen ^^^ Types270
% package Types.TypeDef.Examples;
% import x10.util.*;
%~~vis
\begin{xten}
class TypeExamples {
  static type StringSet = Set[String];
  static type MapToList[K,V] = Map[K,List[V]];
  static type Int(x: Int) = Int{self==x};
  static type Dist(r: Int) = Dist{self.rank==r};
  static type Dist(r: Region) = Dist{self.region==r};
  static type Redund(n:Int, r:Region){r.rank==n} 
      = Dist{rank==n && region==r};
}
\end{xten}
%~~siv
% 
%~~neg

The following code illustrates that type definitions are applicative rather
than generative.  \xcd`B` and \xcd`C` are both aliases for \xcd`String`,
rather than new types, and so are interchangeable with each other and with
\xcd`String`. Similarly, \xcd`A` and \xcd`Int` are equivalent.
%~~gen ^^^ Types280
% package Types.TypeDef.Example.NonGenerative;
% import x10.util.*;
% class TypeDefNonGenerative {
%~~vis
\begin{xten}
def someTypeDefs () {
  type A = Int;
  type B = String;
  type C = String;
  a: A = 3;
  b: B = new C("Hi");
  c: C = b + ", Mom!";
  }
\end{xten}
%~~siv
% }
%~~neg
% An instance of a defined type with no type parameters and no
% value parameters may 


%%MEMBERSHIP%% All type definitions are members of their enclosing package or
%%MEMBERSHIP%% class.  A compilation unit may have one or more type definitions
%%MEMBERSHIP%% or class or interface declarations with the same name, as long
%%MEMBERSHIP%% as the definitions have distinct parameters according to the
%%MEMBERSHIP%% method overloading rules (\Sref{MethodOverload}).


\section{Constrained types}
\label{ConstrainedTypes}
\label{DepType:DepType}
\label{DepTypes}

\index{dependent type}
\index{type!dependent}
\index{constrained type}
\index{generic type}
\index{type!constrained}
\index{type!generic}


Basic types, like \xcd`Int` and \xcd`List[String]`, provide useful
descriptions of data.  

However, one frequently wants to say more.  One might want to know
that a \xcd`String` variable is not \xcd`null`, or that a matrix is
square, or that one matrix has the same number of columns as another
has rows (so they can be multiplied).  In the multicore setting, one
might wish to know that two values are located at the same processor,
or that one is located at the same place as the current computation.

In most languages, there is simply no way to say these things
statically.  Programmers must made do with comments, \xcd`assert`
statements, and dynamic tests.  X10 programs can do better, with {\em
  constraints} on types, and guards on class, method and type
definitions,

A constraint is a boolean expression \xcd`e` attached to a basic type \xcd`T`,
written \xcd`T{e}`.  (Only a limited selection of boolean expressions is
available.)  The values of type \xcd`T{e}` are the values of \xcd`T` for which
\xcd`e` is true.  For example: 

\begin{itemize}
%~~type~~`~~`~~ ~~ ^^^ Types290
\item \xcd`String{self != null}` is the type of non-null strings.  \xcd`self`
      is a special variable available only in constraints; it refers to the
      datum being constrained, and its type is the type to which the
      constraint is attached.
\item If \xcd`Matrix` has properties \xcd`rows` and \xcd`cols`, 
%~~type~~`~~`~~ ~~class Matrix(rows:Int,cols:Int){} ^^^ Types300
      \xcd`Matrix{self.rows == self.cols}` is the type of square matrices.
\item One way to say that \xcd`a` has the same number of columns that \xcd`b`
      has rows (so that \xcd`a*b` is a valid matrix product), one could say: 
%~~gen ^^^ Types310
% package Types.cripes.whered.you.get.those.gripes;
% class Matrix(rows:Int, cols:Int){
% public static def someMatrix(): Matrix = null;
% public static def example(){
%~~vis
\begin{xten}
  val a : Matrix = someMatrix() ;
  var b : Matrix{b.rows == a.cols} ;
\end{xten}
%~~siv
%}}
%~~neg



\index{self}When constraining a value of type \xcd`T`, \xcd`self` refers to the object of
type \xcd`T` which is being constrained.  For example, \xcd`Int{self == 4}` is
the type of \xcd`Int`s which are equal to 4 -- the best possible description
of \xcd`4`, and a very difficult type to express without using \xcd`self`.  
\end{itemize}

\xcd"T{e}" is a {\em dependent type}, that is, a type dependent on values. The
type \xcd"T" is called the {\em base type} and \xcd"e" is called the {\em
  constraint}. If the constraint is omitted, it is \xcd`true`---that is, the
  base type is unconstrained.

Constraints may refer to immutable values in the local environment: 
%~~gen ^^^ Types320
% class ConstraintsMayReferToValues {
% def thoseValues() {
%~~vis
\begin{xten}
     val n = 1;
     var p : Point{rank == n};
\end{xten}
%~~siv
%}}
%~~neg
In a variable declaration, the variable itself is in scope in its
type. For example, \xcd`val nz: Int{nz != 0} = 1;` declares a
non-zero variable \xcd`nz`.
\bard{This will need to be explained further once the language issues are
sorted out..}

%%TYPES-CONSTR-EXP%% We permit variable declarations \xcd"v: T" where \xcd"T" is obtained
%%TYPES-CONSTR-EXP%% from a dependent type \xcd"C{c}" by replacing one or more occurrences
%%TYPES-CONSTR-EXP%% of \xcd"self" in \xcd"c" by \xcd"v". (If such a declaration \xcd"v: T"
%%TYPES-CONSTR-EXP%% is type-correct, it must be the case that the variable \xcd"v" is not
%%TYPES-CONSTR-EXP%% visible at the type \xcd"T". Hence we can always recover the
%%TYPES-CONSTR-EXP%% underlying dependent type \xcd"C{c}" by replacing all occurrences of \xcd"v"
%%TYPES-CONSTR-EXP%% in the constraint of \xcd"T" by \xcd"self".)
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% For instance, \xcd"v: Int{v == 0}" is shorthand for \xcd"v: Int{self == 0}".
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% A variable occurring in the constraint \xcd"c" of a dependent type, other than
%%TYPES-CONSTR-EXP%% \xcd"self" or a property of \xcd"self", is said to be a {\em
%%TYPES-CONSTR-EXP%% parameter} of \xcd"c".\label{DepType:Parameter} \index{parameter}

A constrained type may be constrained further: the type \xcd`S{c}{d}`
is the same as the type \xcd`S{c,d}`.  Multiple constraints are equivalent to
conjoined constraints: \xcd`S{c,d}` in turn is the same as \xcd`S{c && d}`.

\subsection{Syntax of constraints}
\index{constraint!permitted}
\index{constraint!syntax}
\label{PermittedConstraints}
\index{constraint}
\index{expression!allowed in constraint}
\index{expression!constraint}

Only a few kinds of expressions can appear in constraints.  For fundamental
reasons of mathematical logic, the more kinds of expressions that can appear
in constraints, the harder it is to compute the essential properties of
constrained type -- in particular, the harder it is to compute 
\xcd`A{c} <: B{d}`.  It doesn't take much to make this basic fact undecidable.
In order to make sure that it stays decidable, X10 places stringent restrictions on
constraints.  

Only the following forms of expression are allowed in constraints.  

{\bf Value expressions in constraints} may be: 
\begin{enumerate}
\item Literal constants, like \xcd`3` and \xcd`true`;
\item Expressions computable at compile time, like \Xcd{3*4+5};
\item Accessible and immutable variables and parameters;
\item Accessible and immutable fields of objects;
\item Properties of the type being constrained;
\item \xcd`this`, if the constraint is in a place where \xcd`this` is defined;
\item \xcd`here`, if the constraint is in a place where \xcd`here` is defined;
\item \xcd`self`;
\item Calls to property methods, where the receiver and arguments must be
      value expressions acceptable in constraints;
\item \xcd`T haszero`, if \xcd`T` is any type expression.
\end{enumerate}
For an expression \xcd`self.p` to be legal in a constraint, 
\xcd`p` must be 
a property. However terms \xcd`t.f` may be
used in constraints (where \xcd`t` is a term other than \xcd`self` and
\xcd`f` is an immutable field.

{\bf Constraints}, and {\bf Boolean expressions in constraints}  may be any of
the following, where 
all value expressions are of the forms which may appear in constraints: 
\begin{enumerate}
\item Equalities \xcd`e == f`;
\item Inequalities of the form \xcd`e != f`;\footnote{Currently inequalities
      of the form \xcd`e < f` are not supported.}
\item Conjunctions of Boolean expressions that may appear in constraints (but
      only in top-level constraints, not in Boolean expressions in constraints);
\item Subtyping and supertyping expressions: \xcd`T <: U` and \xcd`T :> U`; 
\item Type equalities and inequalities: \xcd`T == U` and \xcd`T != U`; 
% \item Testing a type for a default: \Xcd{hasZero T}.
\end{enumerate}

All variables appearing in a constraint expression must be visible wherever
that expression can used.  \Eg, properties and public fields of an object are
always permitted, but private fields of an object can only constrain private
members.  (Consider a class \xcd`PriVio` with a private field \xcd`p` and a
public method \xcd`m(x: Int{self != p})`, and a call \xcd`ob.m(10)` made
outside of the class. Since \xcd`p` is only visible inside the class, there is
no way to tell if \xcd`10` is of type \xcd`Int{self != p}` at the call site.)

\limitation{
% Currently \Xcd{hasZero T} is not supported.  
Certain spurious syntactic forms -- such as \xcd`a+b,a*b,(c==a&&b),a <b` --
are accepted by the compiler but treated incorrectly.  
}



\subsubsection{Semantics of constraints}
\index{constraint!semantics}
\label{SemanticsOfConstraints}
An assignment of values to variables is said to be a {\em solution} for a
constraint \xcd`c` if under this assignment \xcd`c` evaluates to
\xcd`true`. For instance, the assignment that maps 
the variables \xcd`a` and \xcd`b` to a value \xcd`t` is a solution for
the constraint \xcd`a==b`. An assignment that maps \xcd`a` to 
\xcd`s` and \xcd`b` to a distinct value \xcd`t` is a solution for 
\xcd`a != b`. 

An instance \xcd"o" of \xcd"C" is said to be of type \xcd"C{c}" (or {\em
belong to} \xcd"C{c}") if the constraint \xcd"c" evaluates to \xcd"true" in
the current lexical environment augmented with the binding \xcd"self"
$\mapsto$ \xcd"o".

A constraint \xcd`c` is said to {\em entail} a
constraint \xcd`d` if every solution for \xcd`c` is also a solution
for \xcd`d`. For instance the constraint
\xcd`x==y && y==z && z !=a` entails \xcd`x != a`.

The constraint solver considers the assignment \xcd`a` to \xcd`null`
to  satisfy any constraint of the form \xcd`a.f==t`. Thus, for
instance, the assignment \xcd`var x:Tree{self.home==p}=null` does not
produce an error, since \xcd`self==null` is considered a solution for \xcd`self.home==p`.

To ensure that type-checking is decidable, we require that property graphs be
acyclic.  The property graph, at an instant in an X10 execution, is the graph
whose nodes are all objects in existence at that instance, with an edge from
{$x$} to {$y$} if {$x$} is an object with a property whose value is {$y$}. 
The rules for constructors guarantee this.

Constraints participate in the subtyping relationship in a natural way:
\xcdmath"S[S1,$\ldots$, Sm]{c}" 
is a subtype of 
\xcdmath"T[T1,$\ldots$, Tn]{d}" 
if \xcdmath"S[S1,$\ldots$,Sm]" is a subtype of \xcdmath"T[T1,$\ldots$,Tn]" and
\xcd"c" entails \xcd"d".

For examples of constraints and entailment, see (\Sref{ConstraintExamples})
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% \begin{grammar}
%%TYPES-CONSTR-EXP%% Constraint \: ValueArguments     Guard\opt \\
%%TYPES-CONSTR-EXP%%            \| ValueArguments\opt Guard     \\
%%TYPES-CONSTR-EXP%%            \\
%%TYPES-CONSTR-EXP%% ValueArguments   \:  \xcd"(" ArgumentList\opt \xcd")" \\
%%TYPES-CONSTR-EXP%% ArgumentList     \:  Expression ( \xcd"," Expression )\star \\
%%TYPES-CONSTR-EXP%% Guard            \: \xcd"{" DepExpression \xcd"}" \\
%%TYPES-CONSTR-EXP%% DepExpression    \: ( Formal \xcd";" )\star ArgumentList \\
%%TYPES-CONSTR-EXP%% \end{grammar}
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% In \XtenCurrVer{} value constraints may be equalities (\xcd"=="),
%%TYPES-CONSTR-EXP%% disequalities (\xcd"!=") and conjunctions thereof.  The terms over
%%TYPES-CONSTR-EXP%% which these constraints are specified include literals and
%%TYPES-CONSTR-EXP%% (accessible, immutable) variables and fields, property methods, and the special
%%TYPES-CONSTR-EXP%% constants {\tt here}, {\tt self}, and {\tt this}. Additionally, place
%%TYPES-CONSTR-EXP%% types are permitted (\Sref{PlaceTypes}).
%%TYPES-CONSTR-EXP%% 
%%TYPES-CONSTR-EXP%% \index{self}

%%TYPES-CONSTR-EXP%% Type constraints may be subtyping and supertyping (\xcd"<:" and
%%TYPES-CONSTR-EXP%% \xcd":>") expressions over types.

\subsection{Constraint solver: incompleteness and approximation}
\index{constraint solver!incompleteness}
\index{constraint!entailment}
\index{constraint!subtyping}



The constraint solver is sound in that if it claims that \xcd`c` entails \xcd`d`
then in fact it is the case that every value that satisfies \xcd`c`
satisfies \xcd`d`. 

\limitation{X10's Entailment Algorithm is Incomplete}
However, X10's constraint solver is incomplete. There are situations
in which \xcd`c` entails \xcd`d` but the solver cannot establish it. For
instance it cannot establish that \xcd`a != b && a != c && b != c`
entails \xcd`false` if \xcd`a`, \xcd`b`, and \xcd`c` are of type
\xcd`Boolean`.


Certain other constraint entailments are prohibitively expensive to calculate.  The issues
concern constraints that connect different levels of recursively-defined
types, such as the following.  
%~~gen ^^^ Types330
% package Types.Entailment.EntailFail;
%~~vis
\begin{xten}
class Listlike(x:Int) {
  val kid : Listlike{self.x == this.x};
  def this(x:Int, kid:Listlike) { 
     property(x); 
     this.kid = kid as Listlike{self.x == this.x};}
}
\end{xten}
%~~siv
%
%~~neg
There is nothing wrong with \xcd`Listlike` itself, or with most uses of it;
however, a sufficiently complicated use of it could, in principle, cause X10's
typechecker to fail. 
It is hard to give a plausible example of when X10's algorithm fails, as we
have not yet observed such a failure in practice for a correct program.  

The entailment algorithm of X10 2.0 imposes a certain limit on the number of
times such types will be unwound.   If this limit is exceeded, the compiler
will print a warning, and type-checking will fail in a situation where it is
semantically allowed.  In this case, insert a dynamic cast at the point where
type-checking failed.  

\limitation{ Support for comparisons of generic type variables is
  limited. This will be fixed in future releases.}
% //, and existential quantification over typed variables.

%%TYPES-CONSTR-EXP%% \emph{
%%TYPES-CONSTR-EXP%% Subsequent implementations are intended to support boolean algebra,
%%TYPES-CONSTR-EXP%% arithmetic, relational algebra, etc., to permit types over regions and
%%TYPES-CONSTR-EXP%% distributions. We envision this as a major step towards removing most,
%%TYPES-CONSTR-EXP%% if not all, dynamic array bounds and place checks from \Xten{}.
%%TYPES-CONSTR-EXP%% }




%%PLACE%%\subsection{Place constraints}
%%PLACE%%\label{PlaceTypes}
%%PLACE%%\label{PlaceType}
%%PLACE%%\index{place types}
%%PLACE%%\label{DepType:PlaceType}\index{placetype}
%%PLACE%%
%%PLACE%%An \Xten{} computation spans multiple places (\Sref{XtenPlaces}). Much data
%%PLACE%%can only be accessed from the proper place, and often it is preferable to
%%PLACE%%determine this statically. So, X10 has special syntax for working with places.
%%PLACE%%\xcd`T!` is a value of type \xcd`T` located at the right place for the current
%%PLACE%%computation, and \xcd`T!p` is one located at place \xcd`p`.
%%PLACE%%
%%PLACE%%\begin{grammar}
%%PLACE%%PlaceConstraint     \: \xcd"!" Place\opt \\
%%PLACE%%Place              \:   Expression \\
%%PLACE%%\end{grammar}
%%PLACE%%
%%PLACE%%More specifically, All \Xten{} classes extend the class \xcd"x10.lang.Object",
%%PLACE%%which defines a property \xcd"home" of type \xcd"Place".  \xcd`T!p`, when
%%PLACE%%\xcd`T` is a class, is \xcd`T{self.home==p}`.  If \xcd`p` is omitted, it
%%PLACE%%defaults to \xcd`here`.   \xcd`T!` is far and away the most common usage of
%%PLACE%%\xcd`!`. 
%%PLACE%%
%%PLACE%%Structs don't have \xcd`home`; they are available everywhere.  For structs, 
%%PLACE%%\xcd`T!` and \xcd`T!p` are synonyms for \xcd`T`. Since \xcd`T` is available
%%PLACE%%everywhere, it is available \xcd`here` and at \xcd`p`. 
%%PLACE%%
%%PLACE%%\xcd`!` may be combined with other constraints.  \xcd`T{c}!` is the type of
%%PLACE%%values of \xcd`T!` which satisfy \xcd`c`; it is \xcd`T{c && self.home==here}`
%%PLACE%%for an object type and \xcd`T{c}` for a struct type.  
%%PLACE%%\xcd`T{c}!p` is the type of
%%PLACE%%values of \xcd`T!p` which satisfy \xcd`c`; it is \xcd`T{c && self.home==p}`
%%PLACE%%for an object type and \xcd`T{c}` for a struct type.  
%%PLACE%%
%%PLACE%%
%%PLACE%%
%%PLACE%%% The place specifier \xcd"any" specifies that the object can be
%%PLACE%%% located anywhere.  Thus, the location is unconstrained; that is,
%%PLACE%%% \xcd"C{c}!any" is equivalent to \xcd"C{c}".
%%PLACE%%
%%PLACE%%% XXX ARRAY
%%PLACE%%%The place specifier \xcd"current" on an array base type
%%PLACE%%%specifies that an object with that type at point \xcd"p"
%%PLACE%%%in the array 
%%PLACE%%%is located at \xcd"dist(p)".  The \xcd"current" specifier can be
%%PLACE%%%used only with array types.
%%PLACE%%
%%PLACE%%



\subsection{Example of Constraints}
\label{ConstraintExamples}

Example of entailment and subtyping involving constraints.
\begin{itemize}
\item \xcd`Int{self == 3} <: Int{self != 14}`.  The only value of
      \xcd`Int{self ==3}` is $3$.  All integers but $14$ are members of
      \xcd`Int{self != 14}`, and in particular $3$ is.  
\item Suppose we have classes \xcd`Child <: Person`, and \xcd`Person` has a
      long \xcd`ssn` property.  If \xcd`rhys : Child{ssn == 123456789}`, then
      \xcd`rhys` is also a \xcd`Person` and still has \xcd`ssn==123456789`, so 
      \xcd`rhys : Person{ssn==123456789}` as well.  
      So, \xcd`Child{ssn == 123456789} <: Person{ssn == 123456789}`.
\item Furthermore, since \xcd`123456789 != 555555555`, 
      \xcd`rhys : Person{ssn != 555555555}`.  
      So, \xcd`Child{ssn == 123456789} <: Person{ssn != 555555555}`.  
\item \xcd`T{e} <: T` for any type \xcd`T`.  That is, if you have a value
      \xcd`v` of some base type \xcd`T` which satisfied \xcd`e`, then \xcd`v`
      is of that base type \xcd`T` (with the constraint ignored).
\item If \xcd`A <: B`, then \xcd`A{c} <: B{c}` for every constraint \xcd`{c}`
      for which \xcd`A{c}` and \xcd`B{c}` are defined.  That is, if every
      \xcd`A` is also a \xcd`B`, and \xcd`a : A{c}`, then 
      \xcd`a` is an \xcd`A` and \xcd`c` is true of it. So \xcd`a` is also a
      \xcd`B` (and \xcd`c` is still true of 
      it), so \xcd`a : B{c}`.  
\end{itemize}

Constraints can be used to express simple relationships between objects,
enforcing some class invariants statically.  For example, in geometry, a line
is determined by two {\em distinct} points; a \xcd`Line` struct can specify the
distinctness in a type constraint:\footnote{We call them
\xcd`Position` to avoid confusion with the built-in class \xcd`Point`. 
Also, \xcd`Position` is a struct rather than a class so that the non-equality
test \xcd`start != end` compares the coordinates.  If \xcd`Position` were a
class, \xcd`start != end` would check for different \xcd`Position` objects,
which might have the same coordinates.
}


%~~gen ^^^ Types340
% package triangleExample.partOne;
%~~vis
\begin{xten}
struct Position(x: Int, y: Int) {}
struct Line(start: Position, end: Position){start != end} {}
\end{xten}

%~~siv
%~~neg

Extending this concept, a \xcd`Triangle` can be defined as a figure with three
line segments which match up end-to-end.  Note that the degenerate case in
which two or three of the triangle's vertices coincide is excluded by the
constraint on \xcd`Line`.  However, not all degenerate cases can be excluded
by the type system; in particular, it is impossible to check that the three
vertices are not collinear. 

%~~gen ^^^ Types350
%package triangleExample.partTwo;
% struct Position(x: Int, y: Int) {
%    def this(x:Int,y:Int){property(x,y);}
%    }
% class Line(start: Position, 
%            end: Position{self != start}) {}
% 
%~~vis
\begin{xten}
struct Triangle 
 (a: Line, 
  b: Line{a.end == b.start}, 
  c: Line{b.end == c.start && c.end == a.start})  
 {}
\end{xten}
%~~siv
%
%~~neg

The \xcd`Triangle` class automatically gets a ternary constructor which takes
suitably constrained \xcd`a`, \xcd`b`, and \xcd`c` and produces a new
triangle. 

\section{Default Values}
\index{default value}
\index{type!default value}
\label{DefaultValues}

Some types have default values, and some do not. Default values are used in
situations where variables can legitimately be used without having been
initialized; types without default values cannot be used in such situations.
For example, a field of an object \xcd`var x:T` can be left uninitialized if
\xcd`T` has a default value; it cannot be if \xcd`T` does not. Similarly, a
transient (\Sref{TransientFields}) field \xcd`transient val x:T` is only
allowed if \xcd`T` has a default value.

Default values, or lack of them, is defined thus:
\begin{itemize}
\item The fundamental numeric types (\xcd`Int`, \xcd`UInt`,
      \xcd`Long`, \xcd`ULong`, 
%%limitation%%       \xcd`Short`, \xcd`UShort`, \xcd`Byte`,
%%limitation%%       \xcd`UByte`, 
      \xcd`Float`, \xcd`Double`) all have default value 0.
\item \xcd`Boolean` has default value \xcd`false`.
\item \xcd`Char` has default value \xcd`'\0'`.
\item Struct types other than those listed above have no default value.
\item A function type has a default value of \xcd`null`.
\item A class type has a default value of \xcd`null`.
\item The constrained type \xcd`T{c}` has the same default value as \xcd`T` if
      that default value satisfies \xcd`c`.  If the default value of \xcd`T`
      doesn't satisfy \xcd`c`, then \xcd`T{c}` has no default value.
\end{itemize}

For example, \xcd`var x: Int{x != 4}` has default value 0, which is allowed
because \xcd`0 != 4` satisfies the constraint on \xcd`x`. 
\xcd`var y : Int{y==4}` has no default value, because \xcd`0` does not satisfy \xcd`y==4`.
The fact that \xcd`Int{y==4}` has precisely one value, \viz{} 4, doesn't
matter; the only candidate for its default value, as for any subtype of
\xcd`Int`, is 0. \xcd`y` must be initialized before it is used.

The predicate \xcd`T haszero` tells if the type \xcd`T` has a default value.
\xcd`haszero` may be used in constraints. 

\begin{ex}
The following code defines a sort of cell holding a single value of type
\xcd`T`. The cell is initially empty -- that is, has \xcd`T`'s zero value --
but may be filled later. 
%~~gen ^^^ TypesHaszero10
% package TypesHaszero10;
%~~vis
\begin{xten}
class Cell0[T]{T haszero} {
  public var contents : T;
  public def put(t:T) { contents = t; }
}
\end{xten}
%~~siv
%
%~~neg
\end{ex}

The built-in type \xcd`Zero` has the method \xcd`get[T]()` which
returns the default value of type \xcd`T`.  

\begin{ex}
A variant \xcd`Cell1[T]` which can be initialized with a value of an arbitrary
type
\xcd`T`, or, if \xcd`T` has a default value, can be created with the default
value, is given below.  Note that \xcd`T haszero` is a constraint on one of
the constructors, not the whole type:  
%~~gen ^^^ TypesHaszero20
% package TypesHaszero20;
%~~vis
\begin{xten}
class Cell1[T] {
  public var contents: T;
  def this(t:T) { contents = t; }
  def this(){T haszero} { contents = Zero.get[T](); }
  public def put(t:T) {contents = t;}
}
\end{xten}
%~~siv
%
%~~neg

\end{ex}

\section{Function types}
\label{FunctionTypes}
\label{FunctionType}
\index{function!types}
\index{type!function}

%##(FunctionType
\begin{bbgrammar}
%(FROM #(prod:FunctionType)#)
        FunctionType \: TypeParams\opt \xcd"(" FormalParamList\opt \xcd")" WhereClause\opt Offers\opt \xcd"=>" Type & (\ref{prod:FunctionType}) \\
\end{bbgrammar}
%##)


For every sequence of types \xcd"T1,..., Tn,T", and \xcd"n" distinct variables
\xcd"x1,...,xn" and constraint \xcd"c", the expression
\xcd"(x1:T1,...,xn:Tn){c}=>T" is a \emph{function type}. It stands for
 the set of all functions \xcd"f" which can be applied to a
 list of values \xcd"(v1,...,vn)" provided that the constraint
 \xcd"c[v1,...,vn,p/x1,...,xn]" is true, and which returns a value of
 type \xcd"T[v1,...vn/x1,...,xn]". When \xcd"c" is true, the clause \xcd"{c}" can be
 omitted. When \xcd"x1,...,xn" do not occur in \xcd"c" or \xcd"T", they can be
 omitted. Thus the type \xcd"(T1,...,Tn)=>T" is actually shorthand for
 \xcd"(x1:T1,...,xn:Tn){true}=>T", for some variables \xcd"x1,...,xn".

\limitationx{}
Constraints on closures are not supported.  They parse, but are not checked.

X10 functions, like mathematical functions, take some arguments and produce a
result.  X10 functions, like other X10 code, can change mutable state and
throw exceptions.  Closures (\Sref{Closures}) and method
selectors (\Sref{MethodSelectors}) are of function type.
Typical functions are the reciprocal function: 
%~~gen ^^^ Types360
% package Types.Functions;
% class RecipEx {
% static 
%~~vis
\begin{xten}
val recip = (x : Double) => 1/x;
\end{xten}
%~~siv
%}
%~~neg
and a function which increments  element \xcd`i` of an array \xcd`r`, or throws an exception
if there is no such element, where, for the sake of example, we constrain the
type of \xcd`i`:  
%~~gen ^^^ Types_constraint_b
% package Types_constraint_b;
% NOTEST
% /*NONSTATIC*/class IncrElEx {
% static def example()  {
%~~vis
\begin{xten}
val inc = (r:Array[Int](1), i: Int{i != r.size}) => {
  if (i < 0 || i >= r.size) throw new DoomExn();
  r(i)++;
};
\end{xten}
%~~siv
%}}
%class DoomExn extends Exception{}
%~~neg

In general, a function type needs to list the types 
\xcdmath"T$_i$"
of all the formal parameters,
and their distinct names \xcdmath"x$_i$" in case other types refer to them; a
constraint 
\xcd"c" on the
function as a whole; a return type \xcd"T".

\begin{xtenmath}
(x$_1$: T$_1$, $\dots$, x$_n$: T$_n$){c} => T
\end{xtenmath}


The names \xcdmath"x$_i$" of the formal parameters are not relevant.  Types
which differ only in the names of formals (following the usual rules for
renaming of variables, as in {$\alpha$}-renaming in the {$\lambda$} calculus
\bard{cite something}) are considered equal.  \Eg, the two function types
%~~type~~`~~`~~ ~~ ^^^ Types370
\xcd`(a:Int, b:Array[String](1){b.size==a}) => Boolean`
and \\
%~~type~~`~~`~~ ~~ ^^^ Types380
\xcd`(b:Int, a:Array[String](1){a.size==b}) => Boolean`
are equivalent.

\limitation{
This is not currently implemented properly; these two types are presently
considered different.
}

The formal parameter names are in scope from the point of definition to the
end of the function type---they may be used in the types of other formal parameters
and in the return type. 
Value parameters names may be
omitted if they are not used; the type of the reciprocal function can be
written as
%~~type~~`~~`~~ ~~ ^^^ Types390
\xcd`(Double)=>Double`. 

A function type is covariant in its result type and contravariant in
each of its argument types. That is, let 
\xcd"S1,...,Sn,S,T1,...Tn,T" be any
types satisfying \xcd"Si <: Ti" and \xcd"S <: T". Then
\xcd"(x1:T1,...,xn:Tn){c}=>S" is a subtype of
\xcd"(x1:S1,...,xn:Sn){c}=>T".

A class or struct definition may use a function type 
\xcd"F = (x1:T1,...,xn:Tn){c}=>T" in its 
implements clause; 
this is equivalent to implementing an interface requiring the single method
\xcd"def apply(x1:T1,...,xn:Tn){c}:T". 
Similarly, an interface
definition may specify a function type \xcd"F" in its \xcd"extends" clause.
Values of a class or struct implementing \xcd`F` 
can be used as functions of type \xcd`F` in all ways.  
In particular, applying one to suitable arguments calls the \xcd`apply`
method. 

\limitationx{} A class or struct may not implement two different
instantiations of a generic interface. In particular, a class or
struct can implement only one function type.


A function type \xcd"F" is not a class type in that it does not extend any
type or implement any interfaces, or support equality tests. 
\xcd`F` may be implemented, but not extended, by a class or function type. 
Nor is it a struct type, for it has no predefined notion of equality.


\section{Annotated types}
\label{AnnotatedTypes}

\index{type!annotated}
\index{annotations!type annotations}

        Any \Xten{} type may be annotated with zero or more
        user-defined \emph{type annotations}
        (\Sref{XtenAnnotations}).  

        Annotations are defined as (constrained) interface types and are
        processed by compiler plugins, which may interpret the
        annotation symbolically.

        A type \xcd"T" is annotated by interface types
        \xcdmath"A$_1$", \dots,
        \xcdmath"A$_n$"
        using the syntax
        \xcdmath"@A$_1$ $\dots$ @A$_n$ T".

\section{Subtyping and type equivalence}\label{DepType:Equivalence}
\index{type equivalence}
\index{subtyping}

Intuitively, type \xcdmath"T$_1$" is a subtype of type \xcdmath"T$_2$", 
written \xcdmath"T$_1$ <: T$_2$", 
if
every instance of \xcdmath"T$_1$" is also an instance of \xcdmath"T$_2$".  For
example, \xcd`Child` is a subtype of \xcd`Person` (assuming a suitably defined
class hierarchy): every child is a person.  Similarly, \xcd`Int{self != 0}`
is a subtype of \xcd`Int` -- every non-zero integer is an integer.  

This section formalizes the concept of subtyping. Subtyping of types depends
on a {\em type context}, \viz. a set of constraints on type parameters
and variables that occur in the type.
For example: 

%~~gen ^^^ Types400
% package Types.subtyping.cons;
% NOCOMPILE
%~~vis
\begin{xten}
class ConsTy[T,U] {
   def upcast(t:T){T <: U} :U = t;
}
\end{xten}
%~~siv
%
%~~neg
\noindent
Inside \xcd`upcast`, \xcd`T` is constrained to be a subtype of \xcd`U`, and so
\xcd`T <: U` is true, and \xcd`t` can be treated as a value of type \xcd`U`.  
Outside of \xcd`upcast`, there is no reason to expect any relationship between
them, and \xcd`T <: U` may be false.
However, subtyping of types that have no free variables does not depend
on the context.    \xcd`Int{self != 0} <: Int` is always
true.

\limitation{Subtyping of type variables does not currently work.}


\begin{itemize}
\item {\bf Reflexivity:} Every type \xcd`T` is a subtype of itself: \xcd`T <: T`.

\item {\bf Transitivity:} If \xcd`T <: U` and \xcd`U <: V`, then \xcd`T <: V`. 

\iffalse
{\bf Class types:}  
Given the definition 
\xcd`class C[$\vec{X}$] extends D[$\vec{Y}$]{d} implements I1, ..., In {...}`
where {$\vec{X}$} is a vector of type variables, and 
{$\vec{Y$} a vector of types possibly involving variables from {$\vec{X}$}, 
and {$\vec{T$} an instantiation of {$\vec{X$} and {$\vec{U$} the corresponding
instantiation of {$\vec{Y$}, 
then 
\xcdmath"C[$\vec{T}$]`"is a subtype of \xcd`D[$\vec{U}$]{d}`, \xcd`I1`, ..., \xcd`In`. 

\item
{\bf Interface types:}  
Given the definition 
\xcdmath"interface I[$\vec{X}$] extends I1, ... In {...}`"
then \xcdmath"I` is a subtype of \xcd`"1`, ..., \xcd`In`.

\item 
{\bf Struct types:} 
Given the definition 
\xcdmath"struct S implements I1, ..., In {...}`"then \xcd`S` is a 
subtype of \xcd`I1`, ..., \xcd`In`. 
\fi

\item {\bf Direct Subclassing:} 
Let {$\vec{X}$} be a (possibly empty) vector of type variables, and
{$\vec{Y}$}, {$\vec{Y_i}$} be vectors of type terms over {$\vec{X}$}.
Let {$\vec{T}$} be an instantiation of {$\vec{X}$}, 
and {$\vec{U}$}, {$\vec{U_i}$} the corresponding instantiation of 
{$\vec{Y}$}, {$\vec{Y_i}$}.  Let \xcd`c` be a constraint, and \xcdmath"c$'$"
be the corresponding instantiation.
We elide properties, and interpret empty vectors as absence of the relevant
clauses. 
Suppose that \xcd`C` is declared by one of the
forms: 
\begin{enumerate}
\item \xcdmath"class C[$\vec{X}$]{c} extends D[$\vec{Y}$]{d} implements I$_1[\vec{Y_1}]${i$_1$},...,I$_n[\vec{Y_n}]${i$_n$}{"
\item \xcdmath"interface C[$\vec{X}$]{c} extends I$_1[\vec{Y_1}]${i$_1$},...,I$_n[\vec{Y_n}]${i$_n$}{"
\item \xcdmath"struct C[$\vec{X}$]{c} implements I$_1[\vec{Y_1}]${i$_1$},...,I$_n[\vec{Y_n}]${i$_n$}{"
\end{enumerate}
Then: 
\begin{enumerate}
\item \xcdmath"C[$\vec{T}$] <: D[$\vec{U}$]{d}" for a class
\item \xcdmath"C[$\vec{T}$] <: I$_i$[$\vec{U_i}$]{i$_i$}" for all cases.
\item \xcdmath"C[$\vec{T}$] <: C[$\vec{T}$]{c$'$}" for all cases.
\end{enumerate}


\item
{\bf Function types:}
\xcdmath"(x$_1$: T$_1$, $\dots$, x$_n$: T$_n$){c} => T"
is a  subtype of 
\xcdmath"(x$'_1$: T$'_1$, $\dots$, x$'_n$: T$'_n$){c$'$} => T$'$ "
if: 
\begin{enumerate}
\item Each \xcdmath"T$_i$ <: T$'_i$";
\item \xcd`c` entails \xcdmath"c$'$";
\item \xcdmath"T$'$ <: T";
\end{enumerate}

\item
{\bf Constrained types:}
\xcd`T{c}` is a subtype of \xcd`T{d}` if \xcd`c` entails \xcd`d`. 

\item {\bf Any:} 
Every type \xcd`T` is a subtype of \xcd`x10.lang.Any`.

\item 
{\bf Type Variables:}
Inside the scope of a constraint \xcd`c` which entails \xcd`A <: B`, we have
\xcd`A <: B`.  \eg, \xcd`upcast` above.


%%NO-VARIANCE%% \item 
%%NO-VARIANCE%% {\bf Covariant Generic Types:} 
%%NO-VARIANCE%% If \xcd`C` is a generic type whose {$i$}th type parameter is covariant, 
%%NO-VARIANCE%% and {\xcdmath"T$'_i$ <: T$_i$"}
%%NO-VARIANCE%% and  {\xcdmath"T$'_j$ == T$_j$"} for all {$j \ne i$}, 
%%NO-VARIANCE%% then {\xcdmath"C[T$'_1$, $\ldots$, T$'_n$] <: C[T$'_1$, $\ldots$, T$'_n$]"}.
%%NO-VARIANCE%% \Eg, \xcd`class C[T1, +T2, T3]` with {$i=2$}, and \xcd"U2 <: T2", then
%%NO-VARIANCE%% \xcd`C[T1,U2,T3] <: C[T1,T2,T3]`.
%%NO-VARIANCE%% 
%%NO-VARIANCE%% \item 
%%NO-VARIANCE%% {\bf Contravariant Generic Types:} 
%%NO-VARIANCE%% If \xcd`C` is a generic type whose {$i$}th type parameter is contravariant, 
%%NO-VARIANCE%% and \xcdmath"T$'_i$ <: T$_i$"
%%NO-VARIANCE%% and  \xcdmath"T$'_j$ == T$_j$" for all {$j \ne i$}, 
%%NO-VARIANCE%% then \xcdmath"C[T$'_1$, $\ldots$, T$'_n$] :> C[T$'_1$, $\ldots$, T$'_n$]".
%%NO-VARIANCE%% \Eg, \xcd`class C[T1, -T2, T3]` with {$i=2$}, and \xcdmath"U2 <: T2", then
%%NO-VARIANCE%% \xcd`C[T1,U2,T3] :> C[T1,T2,T3]`.
%%NO-VARIANCE%% 

\end{itemize}


Two types are {\em equivalent}, \xcd`T == U`, if \xcd`T <: U` and \xcd`U <: T`. 


\section{Common ancestors of types}
\label{LCA}

There are several situations where X10 must find a type \xcd`T` that describes
values of two or more different types.  This arises when X10 is trying to find
a good type to describe: 
\begin{itemize}
%~~exp~~`~~`~~test:Boolean ~~ ^^^ Types410
\item Conditional expressions, like \xcd`test ? 0 : "non-zero"` or even 
%~~exp~~`~~`~~test:Boolean ~~ ^^^ Types420
      \xcd`test ? 0 : 1`;
%~~exp~~`~~`~~ ~~ ^^^ Types430
\item Array construction, like \xcd`[0, "non-zero"]` and 
%~~exp~~`~~`~~ ~~ ^^^ Types440
      \xcd`[0,1]`;
\item Functions with multiple returns, like
%~~gen ^^^ Types450
% package Types_odd_inferred_return_type;
% class Examplerator {
%~~vis
\begin{xten}
def f(a:Int) {
  if (a == 0) return 0;
  else return "non-zero";
}
\end{xten}
%~~siv
%}
%~~neg
\end{itemize}

In some cases, there is a unique best type describing the expression.  For
example, if \xcd`B` and \xcd`C` are direct subclasses of \xcd`A`, \xcd`pick`
will have return type \xcd`A`: 
%~~gen ^^^ Types_uniq
% package Types.For.Gripes.About.Pipes.Full.Of.Wipes;
%  class A {} class B extends A{} class C extends A{}
% class D {
%~~vis
\begin{xten}
static def pick(t:Boolean, b:B, c:C) = t ? b : c;  
\end{xten}
%~~siv
%}
%~~neg

However, in many common cases, there is no unique best type describing the
expression.  For example, consider the expression {$E=$} \xcd`b ? 0 : 1`.  The
best type of \xcd`0` 
is \xcd`Int{self==0}`, and the best type of 1 is \xcd`Int{self==1}`.
Certainly {$E$} could be given the type \xcd`Int`, or even \xcd`Any`, and that
would describe all possible results.  However, we actually know more.
\xcd`Int{self != 2}` is a better description of the type of {$E$}---certainly
the result of {$E$} can never be \xcd`2`.   \xcd`Int{self != 2, self != 3}` is
an even better description; {$E$} can't be \xcd`3` either.  We can continue
this process forever, adding integers which {$E$} will definitely not return
and getting better and better approximations. (If the constraint
sublanguage had \xcd`||`, we could give it the type 
\xcd`Int{self == 0 || self == 1`, which would be nearly perfect.  But 
\xcd`||` makes typechecking far more expensive, so it is excluded.)
No X10 type is the best description of {$E$}; there is always a better one.

Similarly, consider two unrelated interfaces: 
%~~gen ^^^ Types460
% package Types.For.Gripes.About.Snipes;
%~~vis
\begin{xten}
interface I1 {}
interface I2 {}
class A implements I1, I2 {}
class B implements I1, I2 {}
class C {
  static def example(t:Boolean, a:A, b:B) = t ? a : b;
}
\end{xten}
%~~siv
%
%~~neg
\xcd`I1` and \xcd`I2` are both perfectly good descriptions of \xcd`t ? a : b`, 
but neither one is better than the other, and there is no single X10 type
which is better than both. (Some languages have {\em conjunctive
    types}, and could say that the return type of \xcd`example` was 
\xcd`I1 && I2`.  This, too, complicates typechecking.)


So, when confronted with expressions like this, X10 computes {\em some}
satisfactory type for the expression, but not necessarily the {\em best} type.  
X10 provides certain guarantees about the common type \xcd`V{v}` computed for 
\xcd`T{t}` and \xcd`U{u}`: 
\begin{itemize}
\item If \xcd`T{t} == U{u}`, then \xcd`V{v} == T{t} == U{u}`.  So, if X10's
      algorithm produces an utterly untenable type for \xcd`a ? b : c`, and
      you want the result to have type \xcd`T{t}`, you can 
      (in the worst case) rewrite it to 
      \xcd`a ? b as T{t} : c as T{t}`.
\item If \xcd`T == U`, then \xcd`V == T == U`.  For example, 
      X10 will compute the type of \xcd`b ? 0 : 1` as 
      \xcd`Int{c}` for some constraint \xcd`c`---perhaps simply 
      picking \xcd`Int{true}`, \viz, \xcd`Int`. 
\item X10 preserves place information about \xcd`GlobalRef`s, because it is so important. If both
      \xcd`t` and \xcd`u` entail \xcd`self.home==p`, then  
      \xcd`v` will also entail \xcd`self.home==p`.  
\item X10 similarly preserves nullity information.  If \xcd`t` and \xcd`u`
      both entail \xcd`x == null` or \xcd`x != null` for some variable
      \xcd`x`, then \xcd`v` will also entail it as well.

\end{itemize}

%\subsection{Syntactic abbreviations}\label{DepType:SyntaxAbbrev}

\section{Fundamental types}

Certain types are used in fundamental ways by X10.  

\subsection{The interface {\tt Any}}

It is quite convenient to have a type which all values are instances of; that
is, a supertype of all types.\footnote{Java, for one, suffers a number of
  inconveniences because some built-in types like \xcd`int` and \xcd`char`
  aren't subtypes of anything else.}  X10's universal supertype is the
  interface \xcd`Any`. 

\begin{xten}
package x10.lang;
public interface Any {
  def toString():String;
  def typeName():String;
  def equals(Any):Boolean;
  def hashCode():Int;
}
\end{xten}

\xcd`Any` provides a handful of essential methods that make sense and are
useful for everything. \xcd`a.toString()` produces a
string representation of \xcd`a`, and \xcd`a.typeName()` the string
representation of its type; both are useful for debugging.  \xcd`a.equals(b)`
is the programmer-overridable equality test, and \xcd`a.hashCode()` an integer
useful for hashing.  


\subsection{The class {\tt Object}}
\label{Object}
\index{\Xcd{Object}}
\index{\Xcd{x10.lang.Object}}

The class \xcd"x10.lang.Object" is the supertype of all classes.
A variable of this type can hold a reference to any object.
\xcd`Object` implements \xcd`Any`.



\section{Type inference}
\label{TypeInference}
\index{type!inference}
\index{type inference}

\XtenCurrVer{} supports limited local type inference, permitting
certain variable types and return types to be elided.
It is a static error if an omitted type cannot be inferred or
uniquely determined. Type inference does not consider coercions.

\subsection{Variable declarations}

The type of a \xcd`val` variable declaration can be omitted if the
declaration has an initializer.  The inferred type of the
variable is the computed type of the initializer.
For example, 
%~~stmt~~`~~`~~ ~~ ^^^ Types470
\xcd`val seven = 7;`
is identical to 
%~~stmt~~`~~`~~ ~~ ^^^ Types480
\xcd`val seven: Int{self==7} = 7;`
Note that type inference gives the most precise X10 type, which might be more
specific than the type that a programmer would write.



\limitation{At the moment,  \xcd`var` declarations may not have their types
elided in this way.  
}

\subsection{Return types}

The return type of a method can be omitted if the method has a body (\ie, is
not \xcd"abstract" or \xcd"native"). The inferred return type is the computed
type of the body.  In the following example, the return type inferred for
\xcd`isTriangle` is 
%~~type~~`~~`~~ ~~ ^^^ Types490
\xcd`Boolean{self==false}`
%~~gen ^^^ Types500
% package Types.Inferred.Return;
%~~vis
\begin{xten}
class Shape {
  def isTriangle() = false; 
}  
\end{xten}
%~~siv
%
%~~neg
Note that, as with other type inference, methods are given the most specific
type.  In many cases, this interferes with subtyping.  For example, if one
tried to write: 
\begin{xten}
class Triangle extends Shape {
  def isTriangle() = true;
}
\end{xten}
\noindent
the X10 compiler would reject this program for attempting to override
\xcd`isTriangle()` by a method with the wrong type, \viz,
\xcd`Boolean{self==true}`.  In this case, supply the type that is actually
intended for \xcd`isTriangle`, such as 
\xcd`def isTriangle() :Boolean =false;`. 

The return type of a closure can be omitted.
The inferred return type is the computed type of the body.

The return type of a constructor can be omitted if the
constructor has a body.
The inferred return type is the enclosing class type with
properties bound to the arguments in the constructor's \xcd"property"
statement, if any, or to the unconstrained class type.
For example, the \xcd`Spot` class has two constructors, the first of which has
inferred return type \xcd`Spot{x==0}` and the second of which has 
inferred return type \xcd`Spot{x==xx}`. 
%~~gen ^^^ Types510
% package Types.Inferred.By.Phone;
%~~vis
\begin{xten}
class Spot(x:Int) {
  def this() {property(0);}
  def this(xx: Int) { property(xx); }
}
\end{xten}
%~~siv
%class Confirm{ 
% static val s0 : Spot{x==0} = new Spot();
% static val s1 : Spot{x==1} = new Spot(1);
%}
%~~neg


\index{void}

A method or closure that has expression-free \xcd`return` statements
(\xcd`return;` rather than \xcd`return e;`) is said to return \xcd`void`.
\xcd`void` is not a type; there are no \xcd`void` values, nor can \xcd`void`
be used as the argument of a generic type. However, \xcd`void` takes the
syntactic place of a type. A method returning \xcd`void` can be specified by
\xcd`def m():void`: 

%~~gen ^^^ Types520
% package Types.voidd;
% class voidddd {
% static 
%~~vis
\begin{xten}
val f : () => void = () => {return;};
\end{xten}
%~~siv
%}
%~~neg

By a convenient abuse of language, \xcd`void` is sometimes
lumped in with types; \eg, we may say ``return type of a method'' rather than
the formally correct but rather more awkward ``return type of a method, or
\xcd`void`''.   Despite this informal usage, \xcd`void` is not a type.  For
example, given 
%~~gen ^^^ Types530
% package Types.void_is_not_a_type;
% class EEEEVil {
%~~vis
\begin{xten}
  static def eval[T] (f:()=>T):T = f();
\end{xten}
%~~siv
% }
%~~neg
\noindent
The call \xcd`eval[void](f)` does {\em not} typecheck; \xcd`void` is not a
type and thus cannot be used as a type argument.  There is no way in X10 to
write a generic function which works with both functions which return a value
and functions which do not.  In most cases, functions which have no sensible
return value can be provided with a dummy return value.

\subsection{Inferring Type Arguments}
\label{TypeParamInfer}


A call to a polymorphic method %, closure, or constructor 
may omit the
explicit type arguments.  If the method has a type parameter
\xcd"T", the type argument corresponding to \xcd"T" is inferred
to be a common ancestor of the types of any formal
parameters of type \xcd"T".

(Exception: it is an error if the method call provides no information about
a type parameter that must be inferred.  For example, given the method
definition \xcd`def m[T](){...}`, an invocation \xcd`m()` is considered a
static error.  The compiler has no idea what \xcd`T` the programmer intends.)


%TODO--check this!
Consider the following method, which chooses one of its arguments.  (A more
sophisticated one might sometimes choose the second argument, but that does
not matter for the sake of this example.)
\begin{xten}
static def choose[T](a: T, b: T): T = a; 
\end{xten}

The type argument \xcd`T` can always be supplied: 
\xcd`choose[Int](1, 2)` picks an integer, 
and \xcd`choose[Any](1, "yes")` picks a value that might be an integer or a
string.  
However, the type argument can be elided.  Suppose that \xcd`Sub <: Super`;
then the following compiles: 

%~~gen ^^^ Types540
% package Types.GenericInference;
% class Exampllll{ 
%~~vis
\begin{xten}
  static def choose[T](a: T, b: T): T = a; 
  static val j : Any = choose("string", 1);
  static val k : Super = choose(new Sub(), new Super());
\end{xten}
%~~siv
%}
% class Super {}
% class Sub extends Super {}
%~~neg


\subsubsection{Sketch of X10 Type Inference for Method Calls}

When the X10 compiler sees a method call 
\xcdmath"a.m(b$_1$, $\ldots$,b$_n$)", and attempts to infer type parameters to see if it could be use of a
method \xcdmath"def m[X$_1$, $\ldots$, X$_t$](y$_1$: S$_1$, $\ldots$, y$_n$:S$_n$)", 
it reasons as follows. 



Suppose that \xcdmath"b$_i$" has type \xcdmath"T$_i$".  Then, X10 is seeking a
set of type {$B$} bindings \xcdmath"X$_j$ = U$_j$", for $1 \le j \le t$, such that 
\xcdmath"T$_i$ <: S$^*_i$" for {$1 \le i \le n$}, where \xcdmath"S$^*$" is
\xcd`S` with each type variable \xcdmath"X$_j$" replaced by the corresponding
\xcdmath"U$_j$".  If it can find such a {$B$}, it has a usable choice of type
arguments and can do the type inference.  If it cannot find {$B$}, then it
cannot do type inference.    (Note that X10's type inference algorithm is
incomplete -- there may {\em be} such a {$B$} that X10 cannot find.  If this
occurs in your program, you will have to write down the type arguments
explicitly.) 

Let $B_0$ be the set {$\{ T_i \subtype S_i | 1 \le i \le n\}$}.  Let
{$B_{n+1}$} be {$B_n$} with one element {$F \subtype G$} or 
{$F \typeeq G$} removed, and
{$C(F \subtype G)$} 
or {$C(F \typeeq G)$} (defined below) added.  Repeat this until 
{$B_n$} consists entirely of comparisons with type variables (\viz, 
\xcdmath"Y$_j$ == U", 
\xcdmath"Y$_j$ <: U", and
\xcdmath"Y$_j$ :> U"), 
or until some {$n$} exceeds a predefined compiler limit. 

The candidate inferred types may be read off of {$B_n$}.  The guessed binding
for \xcdmath"X$_j$" is: 
\begin{itemize}
\item If there is an equality \xcdmath"X$_j$==W" in {$B_n$}, then guess the
      binding \xcdmath"X$_j$=W".  Note that there may be several such
      equalities with different choices of \xcd`W`, but, if the inference is
      to work, all the choices of \xcd`W` must be equal types anyways.
\item Otherwise, if there is one or more upper bounds 
\xcdmath"X$_j$ <: V$_k$" in {$B_n$}, guess the binding 
\xcdmath"X$_j$ = V$_+$", where 
\xcdmath"V$_+$" is the computed lower bound of all the \xcdmath"V$_k$"'s.
\item Otherwise, if there is one or more lower bounds 
\xcdmath"R$_k$ <: X$_j$", guess that
\xcdmath"X$_j$ = R$_+$", where 
\xcdmath"R$_+$" is the computed upper bound of all the \xcdmath"R$_k$"'s.
\end{itemize}
If this does not yield a binding for some variable \xcdmath"X$_j$", then type
inference fails.  Furthermore, if every variable \xcdmath"X$_j$" is given a
binding \xcdmath"U$_j$", but the 
bindings do not work --- 
that is, if 
\xcdmath"a.m[U$_1$, $\ldots$, U$_t$](b$_1$, $\ldots$,b$_n$)"
is not a call of 
the original method 
\xcdmath"def m[X$_1$, $\ldots$, X$_t$](y$_1$: S$_1$, $\ldots$, y$_n$:S$_n$)"
--- then type inference also fails.

\paragraph{Computation of the Replacement Elements}

Given a type relation
{$r$} of the form {$F \subtype G$}
or {$F \typeeq G$}, we compute the set {$C(r)$} of
replacement constraints.  There are a number of cases; we present only the
interesting ones. 

\begin{itemize}
\item If $F$ has the form \xcdmath"$F'${c}", then  
\xcdmath"$C(r)$" is defined to be
 \xcdmath"$F'$ == $G$" if $r$ is an equality, or 
 \xcdmath"$F'$ <: $G$" if {$r$} is a subtyping.
That is, we erase type constraints.  
Validity is not an issue at this point in the algorithm, as 
we check at the end that the result is valid.
However, in important cases, the replacement is valid, in the sense that the 
solutions of {$B_{k+1}$} are precisely the solutions of {$B_k$}.
Specifically, if the equation had the form \xcdmath"Z{c} == A", it could be
solved by \xcd`Z==A` or by \xcd`Z = A{c}`.  By dropping constraints in this
rule, we choose the former solution. 

\item Similarly, we drop constraints on {$G$} as well.

\item If {$F$} has the form \xcdmath"K[F$_1$, $\ldots$, F$_k$]"
and 
{$G$}
has the form \xcdmath"K[G$_1$, $\ldots$, G$_k$]", 
then {$C(r)$} has one type relation comparing each parameter of 
{$F$} with the corresponding one of {$G$}. 

If {$r$} is a type equality {$F \typeeq G$}, then 
{$C(r) = \{ F_l \typeeq G_l | 1 \le l \le k$}.

If {$r$} is a type comparison, and the {$l^{th}$} type parameter of \xcd`K` is
invariant,
%%NO-VARIANCE%%  (resp. covariant or contravariant)
then 
{$C(r)$} has {$F_l \typeeq G_l$}
(resp. {$F_l \subtype G_l$} or {$G_l \subtype F_l $}). 


For example, the constraint \xcdmath"List[X] == List[Y]" produces the
constraint \xcd`X==Y`.

\item Other cases are fairly routine.  \Eg, if {$F$} is a \xcd`type`-defined
      abbreviation, it is expanded.

\end{itemize}

\begin{ex}

%~~gen ^^^ Types1s4y
% package Types1s4y;
%~~vis
\begin{xten}
import x10.util.*;
class Cl[C1, C2, C3]{}
class Example {
  static def me[X1, X2](Cl[Int, X1, X2]) = 
     new Cl[X1, X2, Point]();
  static def example() {
    val a = new Cl[Int, Boolean, String]();
    val b : Cl[Boolean, String, Point] = me[Boolean, String](a);
  }
}
\end{xten}
%~~siv
%
%~~neg

\end{ex}

\section{Type Dependencies}

Type definitions may not be circular, in the sense that no type may be its own
supertype, nor may it be a container for a supertype. This forbids interfaces
like \xcd`interface Loop extends Loop`, and indirect self-references such as
\xcd`interface A extends B.C` where \xcd`interface B extends A`.  

The formal definition of this is based on Java's.  

An {\em entity type} is a class, interface, or struct type.   

Entity type $E$ {\em directly depends on} entity type $F$ if $F$ is mentioned
in the \xcd`extends` or \xcd`implements` clause of $E$, either by itself or as
a qualifier within a super-entity-type name.  
\begin{eg}
In the following, \xcd`A` directly depends on \xcd`B`, \xcd`C`, \xcd`D`, 
\xcd`E`, and \xcd`F`.    It does not directly depend on \xcd`G`.
%~~gen ^^^ Types6a9m
% package Types6a9m;
% NOTEST
% class B{ static class C{}}
% class D{ static interface E{}}
% interface F[X]{}
% class G{}
%~~vis
\begin{xten}
class A extends B.C implements D.E, F[G] {}
\end{xten}
%~~siv
%
%~~neg
It is an ordinary programming idiom to use \xcd`A` as an argument to a generic
interface that \xcd`A` implements.  For example, \xcd`ComparableTo[T]`
describes things which can be compared to a value of type \xcd`T`. Saying that
\xcd`A` implements \xcd`ComparableTo[A]` means that one \xcd`A` can be
compared to another, which is reasonable and useful: 
%~~gen ^^^ Types2x6d
% package Types2x6d;
%~~vis
\begin{xten}
interface ComparableTo[T] {
  def eq(T):Boolean;
}
class A implements ComparableTo[A] {
  public def eq(other:A) = this.equals(other);
}
\end{xten}
%~~siv
%
%~~neg

\end{eg}

Entity type $E$ {\em depends on} entity type $F$ if
either $E$ directly depends on $F$, or $E$ directly depends on an entity type
that depends on $F$.   That is, the relation ``depends on'' is the transitive
closure of the relation ``directly depends on''.  

It is a static error if any entity type $E$ depends on itself.
