\documentclass{article}
\input{../../../../vj/res/pagesizes}
\input{../../../../vj/res/vijay-macros}
\def\csharp{{\sf C\#}}
\def\implies{\withmath{\;\mbox{\tt =>}\;}}
\usepackage[pdftex]{graphicx}
\def\Xten{{\sf X10}}
\def\java{{\sf Java}}
\def\FXten{{\sf FX10}}
\begin{document}
\title{Formalizing Dependent Types for \Xten}
\author{
({\sc Draft Version 0.09})\\
(Please do not cite)\\
(Send comments to {\tt vsaraswa@us.ibm.com}.)
}

\date{April 19, 2006}
\maketitle

\begin{abstract}
We formalize the basic ideas of dependent-types for Java like
languages (introduced in \cite{DependentTypes}), in the context of
\FXten, a Featherweight version of \Xten{} \cite{X10}.  For now, we
focus on a sequential language; distribution and concurrency
constructs {\em a la} \Xten{} are to be added later. This note should
be taken as a companion to \cite{X10-concur} which formalizes the
dynamic semantics for \FXten. It is intended to provide the basis for
a place-based type-system for scalars and arrays.
\end{abstract}

\section{Introduction}

\section{Basic \FXten}
{\em 
TO BE DONE: Add the ability to specify a constraint on super's fields
when subclassing
{\tt  class C(f:X, var g:Y) extends D(\&c) \{M\}}
Here {\tt c} is a constraint on {\tt f},{\tt g} and the fields of {\tt D}.
}
The abstract syntax for \FXten{} is specified in
Table~\ref{Table:AST}. We consider a parametric version of the
language, with an underlying constraint system $\cal C$ \cite{cccc} being used to
specify dependent types. 

\begin{table}
  \begin{tabular}{|lllll|}
\hline
%% IMPLICIT CTOR CHANGE
 (Classes) & CL &{::=}& {\tt class C($\tt \bar{\tt f}:\bar{\tt X}$, var $\tt\bar{\tt g}:\bar{\tt Y}$)}&\\
&&& \tt extends D \{$\bar{\tt M}$\} & \\
 (Method) & M &{::=}& {\tt def m($\tt\bar{\tt x}:\bar{\tt X}$\&c):T = e} & \\
 (Expression) & e,r,s &{::=}&  {\tt new C($\tt \bar{x}$)}  & (New object) \\
%%    &&& \alt {\tt \{val $\tt\bar{\tt x}:\bar{\tt X}=\bar{\tt e}$; e\}} \\
    &&& \alt {\tt \{val x:X=e; e\}} & (Local variable declaration)\\
    &&& \alt {\tt x.m($\tt \bar{\tt x}$)} & (Method Invocation) \\
    &&& \alt {\tt (T) e} & (Cast) \\
    &&& \alt {\tt x.f=e}  \alt {\tt e;e} & (Assignment, Sequencing) \\
    &&& \alt {\tt ce}  \alt {\tt ce \implies e : e} & \\
 (Type) & T,U,\ldots &{::=}& {\tt C(\&c)} \alt \tt nullable\ C(\&c) & \\
 (Constraint) & c,d &{::=}& ce \alt $\tt (\exists x:T)\, c$ & \\
 (Constraint Term) & ce &{::=}& \tt null \alt true \alt false \alt n &\\
       &&& \alt \tt x \alt x.f  \alt ce op ce \alt op ce & \\
\hline
  \end{tabular}
\caption{Abstract syntax for \FXten.}\label{Table:AST}
\end{table}

We follow \cite{FJ,MJ} in our treatment. Meta-variables {\tt C}, {\tt
D}, {\tt E} range over class names (including the ``built-in'' classes
{\tt int}, {\tt boolean} and {\tt Object}); {\tt f} and {\tt g} range over
field names; {\tt m} ranges over method names; {\tt x}, {\tt y}, {\tt
z} range over parameter and local constant names; other meta-variables
are specified in Table~\ref{Table:AST}.

We write $\tt \bar{e}$ as shorthand for $\tt e_1,\ldots, e_n$
(comma-separated sequence); this sequence may be empty ({\tt n=0}).
Similarly for $\bar{x}$.
$\bar{\tt M}$ and $\bar{\tt K}$ are similar except that no commas
separate the items in the sequence. We use the obvious abbreviation:
$\tt\bar{\tt f}:\bar{X},$ for $\tt f_1:X_1,\ldots, f_n:X_n$
({\tt n} may be zero).  {\tt var $\tt\bar{\tt g}:\bar{\tt Y}$}
abbreviates the sequence {\tt var $\tt g_1:Y_1, \ldots, g_n:Y_n$} if
$\tt n > 1$ and the empty sequence otherwise. Empty parameter
sequences may be omitted (like Scala, unlike Java). 

%%{\tt this.$\bar{\tt f}$=$\bar{\tt e}$;} abbreviates {$\tt this.f_1=e_1;\ldots
%%this.f_n=e_n$}.  {\tt \{val $\tt\bar{\tt x}:\bar{\tt X}=\bar{\tt
%%e};\}$} abbreviates {\tt \{val $\tt x_1: X_1= e_1;\ldots;
%%x_n:X_n=e_n;\}$}.  

Sequences of field declarations, parameter declarations, local
variable declarations, are assumed to not contain any duplicates.
Sequences of methods in a class must not contain two methods with the
same sequence of parameters types.  (\FXten{} permits {\em ad hoc}
polymorphism.)

%% IMPLICIT CTOR CHANGE
\paragraph{Class definition.}

A {\em class definition} specifies {\tt val} (immutable) fields, {\tt var}
(mutable fields) and their types, its super-class, its list of
methods, and an (implicit) constructor. Modulo minor syntactic changes
(and dependent types), one may view the declaration

$$
{\tt class\ C(\tt \bar{\tt f}:\bar{\tt X},\, var\ \tt\bar{\tt g}:\bar{\tt Y})\;extends\ D\;\{\bar{\tt M}\}}
$$

\noindent as shorthand for the Java class definition:

{\footnotesize
\begin{tabular}[t]{l}
\tt class C extends D \{ \\
\tt  \quad final $\bar{\tt X} \bar{\tt f}$;\\
 \tt \quad $\bar{\tt Y}$\ $\bar{\tt g}$; \\
 \tt \quad C($\bar{\tt X}$\ $\bar{\tt f}$,\,$\bar{\tt Y}$\ $\bar{\tt g}$,\,$\bar{\tt Z}$\ $\bar{\tt h}$) \{\\
  \tt \quad\quad super($\bar{\tt h}$); \\
  \tt \quad\quad this.$\bar{\tt f}$=$\bar{\tt f}$;\\
  \tt \quad\quad this.$\bar{\tt g}$=$\bar{\tt g}$;\\
 \quad \} \\
\tt \quad $\bar{\tt M}$\\
\tt \}
\end{tabular}}

\noindent where $\bar{\tt h}:\bar{\tt Z}$ is the list of argument types of the
constructor for the super class {\tt D}. 

\paragraph{Methods.}
In a method definition, {\tt c} is a constraint on the parameters
$\bar{\tt x}$ and the final fields of the current object.  This
constraint must be true (statically) for the method to be invoked.
One can think of {\tt c} as allowing the programmer to specify
conditions on the object which must be true for the method to be
invoked. This flexibility is particularly valuable for nullary methods
-- for such methods the constraint cannot be folded into the type of
arguments.

As usual for references to a field in the body of a class (not within
a dependent type, see below), the occurrence of a field {\tt f} in
{\tt c} stands for {\tt this.f}.

\paragraph{Expressions.}
For expressions, we assume the following precedence order (from less
tight to more tight): sequencing, type-cast, assignment, conditional,
field invocation, method invocation.

We also reserve the local constant names ``{\tt this}'' and ``{\tt
self}''. That is, no program may define a local constant or parameter
named {\tt this} or {\tt self}. 

We note that, somewhat unusually, field selection, assignment, method
invocation and constructor invocation take constants as arguments,
rather than expressions. This is necessary because we need a name for
the arguments so that the name can be substituted for the formal
argument in the resulting type. The version of these operations which
takes arbitrary expressionas as arguments can be obtained by combining
with the local variable combinator. Thus, {\tt e.m(e1)} is simply {\tt
\{val x:X=e; \{val y:Y=e1; x.m(y)\}\}}, where {\tt x} and {\tt y} are
new local variables, and the type of the expressions {\tt e} and {\tt
e1} is {\tt X} and {\tt Y} respectively. Below, when writing actual
programs we shall feel free to use the abbreviated {\tt e.m(e1)} form.

The two-armed conditional expression $\tt ce \implies e : e$ is a
typecase expression: it permits the compiler to reason conditionally
about the expression, by propagating the constraint down the positive
branch and its negation down the negative branch.

\paragraph{Types.}
We reserve the class names ``{\tt Object}'', ``{\tt int}'' and ``{\tt
boolean}''. That is, the user may not define these
classes. \footnote{In a subsequent version of this document we will
introduce value types, and then {\tt int} will be just another value
type, defined with native methods.}

A {\em dependent type} is of the form {\tt C(\&c)} where {\tt c} is a
constraint.  The phrase ``{\tt \&c}'' is called a {\em where
clause}. We abbreviate {\tt \&true} to the empty string. Since
empty parameter lists can be omitted, this means that {\tt C(\&true)}
can be abbreviated to {\tt C}.

Intuitively, a type {\tt C(\&c)} is the type of all objects that are
instances of {\tt C} and satisfy the condition {\tt c}. Note that if
the condition {\tt c} is unsatisfiable, then the type is
empty. Variables/parameters cannot be declared at empty types; this is
checked statically.

{\tt c} may contain references to parameters and local constants
visible at the point of declaration of the type (including {\tt
this}), and the special constant {\tt self}.  {\tt self} refers to the
object whose type is being specified. A field {\tt f} of {\tt C} may
occur unqualified in {\tt c}, such a reference is supposed to be
shorthand for {\tt self.f}. References to the {\tt f} field of the
current object must be explicitly preceded by {\tt this}.\footnote{
Note that in general {\tt this.f} is different from {\tt self.f}.  For
instance the type {\tt B(\& f = this.g)} appearing in the body of the
definition of the class {\tt A} is the type of all instances of {\tt
B} whose {\tt f} field has the same value as the {\tt g} field of the
current {\tt A} object. (It is the same as the type {\tt B(\& self.f =
this.g}).)}

{\tt self} is often absent in types. It is particularly useful in {\em
singleton types}, e.g. {\tt Point(\&self=p)} which is satisfied by any
object that is an instance of {\tt Point} and is the same as {\tt p},
and in {\em subrange types}, e.g. {\tt int(\&self >= 0)}.

%%The special variable {\tt this} may {\em not} be referenced in a typei
%%in the definition of a constructor (e.g.{} in the constraint in the
%%parameter list of the constructor, or in the return type of the
%%constructor). This recognizes the fact that an object does not exist
%%until it is created. 

The type {\tt nullable C(\&c)} is the type {\tt C(\&c)} together with
the special value {\tt null}. Thus the type {\tt nullable C(\&c)} is
never empty: if {\tt c} is inconsistent it permits precisely the value
{\tt null}.

The terms {\tt ce} in a constraint are drawn from an underlying
constraint system, $\cal C$ \cite{cccc}. {\em For now we take the
constraint system to be fixed, but it makes sense to permit the
programmer to extend the constraint system with new value types, and
operations over them, provided that a constraint solver is supplied
for these types.} $\cal C$ specifies a collection of value types
(e.g.{} {\tt int}, {\tt bool}, constants of those types and operations
on those types. 

We also permit the shorthand {\tt C($\overline{\tt ce}$)} for {\tt
C(\&$\overline{\tt f}=\overline{\tt ce}$)} where $\tt \overline f$ is the textual
order enumeration of the {\tt val} fields of {\tt C}. If the class has
no field, then we use the shorthand {\tt C(ce)} for {\tt
C(\&self=ce)}. Thus {\tt int(0)} is satisfied precisely by the value
{\tt 0}.

Finally, we permit the shorthand $\tt (\exists x:T) C(\&c)$ for
$\tt C(\exists x:T)c)$.

{\em In a later version of this document we will introduce type definitions.
This will let us use, for instance, the abbreviation {\tt nat}
for the type {\tt int(\&self >= 0)}. 
}

\paragraph{Program.}
A {\em program} is a pair of a set of classes and an expression. For
simplicity we shall leave the set of classes implicit. We shall assume
that every class name used in the program (except {\tt Object}) is
defined exactly once in the program. We assume that the class
hierarchy, defined by $\leq$ in the next section is acyclic
(anti-symmetric).

\subsection{Remarks on the syntax}
%%In a class declaration, $\tt\bar{\tt f}$ represents the val (final)
%%fields of the class, $\tt\bar{\tt g}$ the var (mutable)
%%fields.\footnote{ The addition of {\tt var} constructor and method
%%parameters and local variables is routine and omitted for
%%brevity. Similarly for initializers of fields, static fields and
%%methods. We may add exceptions later since they play an integral part in the
%%semantics of concurrent constructs.} 

%% $\tt c$ is the {\em class
%% invariant} (a constraint on the fields of the class true for all
%% instances of the class); this must be checked when validating the
%% constructor.\footnote{In the current formulation there is only one
%% constructor, so the class invariant can be folded into the return type
%% of the constructor. However, it makes sense to permit multiple
%% constructors in the future, so we shall retain the separate syntax for
%% the class invariant.}

After Scala \cite{scala}, we have chosen the ML-style variable
declaration (type comes after the variable, and is separated by a
colon), as opposed to conventional Java-style declarations, so that
the return type of a method may appear after the declaration of the
parameters of the method. This permits parameters to appear in the
return type, while respecting the ``define before use''
meta-rule. This syntax also permits types (for parameters, variables
and methods) to be optional, while still retaining readability.  Note
that the language above does not permit mutable constructor or method
parameters.

%%In a constructor declaration, {\tt c} is a constraint on the
%%constructor parameters $\bar{\tt x}$ that must be true for the
%%constructor to be invoked.  A constructor may specify the return type
%%(useful for specifying constraints on the fields of the object
%%returned by the constructor).  

The abstract syntax differs from the syntax in \cite{DependentTypes}
in many notational respects. It differs substantively in that there is
no distinction between parameters of classes and fields, and
parameters of methods and arguments to the method. All arguments to a
method are considered final. There are no implicit parameters for
classes.  All final fields of a class may be used in defining a
dependent type on that class.

\subsection{Example}
\begin{example}[List]\label{List}
Consider the class {\tt List}. We shall use generic syntax for
type parameters; this will be formalized in a subsequent version 
of this note in a fashion similar to \cite{FJ}. For now, 
the reader should understand generic syntax in the spirit of \cite{FJ}.

{\footnotesize
\begin{verbatim}
  class List<X>( n:int(&self>=0), 
                 var node: nullable X, 
                     rest: nullable List<X>(n-1)) extends Object {
    def makeList:List<X>(0)=new List(0,null, null)
    def makeList(node:X):List<X>(1)=new List(1,node,makeList)
    def makeList{node:X, rest:nullable List<X>):List<X>(rest.n+1)=
     rest==null => makeList(node) : new List(rest.n+1,node,rest)
    def EmptyListInt:List<int(&self>=0)>(0)=new List<int(&self>=0)>(0,null,null)

    def append(arg:List<X>):List<X>(n+arg.n)  =
         (n == 0) => arg : new List<X>(node, rest.append(arg))
    def rplacd(arg:List<X>(n-1)):List<X>(n)  = {
         this.rest=arg;
         this
    }
    def rev : List<X>(n) = rev(new List<X>())
    def rev(arg:List<X>):List<X>(n+arg.n) = 
         (n == 0) => arg : rest.rev(new List<X>( node, arg))
    def filter(f: fun<X,boolean>):List<X> = 
         (n==0) => this
         : (f(node) => 
           new List<X>(node, rest.filter(f));
            : rest.filter(f))
    /** Return a list of m numbers from o..m-1. */
    def gen(m:int(&self>= 0)):List<int(&self>=0)>(m) = gen(0,m)
    /** Return a list of (m-i) elements, from i to m-1. */
    def gen(i:int(&self>=0), m:int(&self>=i)):List<int(&self>=0)>(m-i) = 
       (i == m) => EmptyListInt : EmptyListInt.makeList(i, gen(i+1,m))
  }  
\end{verbatim}}

The class {\tt List} has three fields, the {\tt val} field {\tt n},
and the {\tt var} fields {\tt node} and {\tt rest}. {\tt rest} is
required to be a {\tt List} whose {\tt n} field has the value {\tt
this.n-1}.  {\tt n} is required to be non-negative. Note that {\tt
rest} is forced to be {\tt null} if {\tt n=0}, since {\tt
List<X>(n-1)} will be empty in this case.

Three make methods are provided that call the implicit constructor
with different arguments. While the first two return lists of
constant size, the third takes  returns a list of a size
that depends on one of its arguments. 

Functions that append one list to another or that reverse a list can
be defined quite naturally. In both cases the size of the
list returned is known statically (as a function of the size of the
list and the argument to the method).  The example also illustrates a
method {\tt filter} which returns a list whose size cannot be known
statically (it depends on properties of the argument function {\tt f}
which are not captured statically).

The {\tt gen} methods\footnote{These should really be {\tt static}.}
illustrate ``{\tt self}''-constraints. The first {\tt gen} method
takes a single argument {\tt m} that is required to be a non-negative
{\tt int}.  The second {\tt gen} method illustrates that the type of a
parameter can depend on the value of another parameter: {\tt m} is
required to be no less than {\tt i}. This assumption are necessary in
order to guarantee that the result type of the method is not empty,
that is, to guarantee {\tt m-i >= 0}.
\end{example}

{\em Note:
Language extensions to be made to get a fuller subset of \Xten:

\begin{itemize}
  \item Add multiple constructors.
  \item Permit arbitrary code in constructors.
  \item Add static state -- or at least distinguish between objects and traits.
  \item Permit fields to be overridden?
  \item Add functions.
  \item Add exceptions.
  \item Add places.
  \item Add async, finish, future.
  \item Add regions, distributions and arrays.
  \item Add generics with declaration-time
  covariance/contravariance/invariance, a la Scala.  Generics should
  be instantiable with arbitrary value types.
  \item Consider adding generic values (universal quantification).
  \item Arrays should be generic.
\end{itemize}
}

\subsection{Type system}

A {\em typing environment}, $\Gamma$, is a collection of
constraints. Constraints are taken from the underlying constraint
system $\cal C$, and extended to include tokens of the form {\tt var
x.g:T} (read as: {\em {\tt x} has a field {\tt g} of type {\tt T}}) 
and {\tt x:T} (read as: {\tt x} has type {\tt T}).

The constraint system satisfies the axiom:

$$
\tt \Gamma, x:C(\&c) \vdash c[x/self]
$$

\noindent Thus, for instance, we have:
$$
\tt x:List(\& n==3) \vdash x.n==3
$$
\noindent (asuming {\tt n} is a field in {\tt List}, and hence an abbreviation for {\tt self.n}). We also remind the reader that for any constraint system $\Gamma \vdash c\;\&\;d$ iff 
$\Gamma\vdash c$ and $\Gamma\vdash d$.

Given a program (collection of classes and an expression), we shall
also assume that the entailment relationship is closed under subtyping (see below):

$$
\from{\tt \Gamma \vdash x:C \ \ \ C \leq D }
\infer{\tt \Gamma \vdash x:D}
$$

\noindent and field selection. That is,

$$
\from{
\tt \Gamma \vdash x:C \ \ \theta= [x/this]\ \  \ class\ C(\bar{\tt f}:\bar{\tt X},\,var\ \bar{\tt g}:\bar{\tt Y}) \ldots }
\infer{
\begin{array}{l}
\tt \Gamma \vdash val\ x.\bar{\tt f}\;\&\; x.\bar{\tt f} : \bar{\tt X}\theta \\
\tt \Gamma \vdash var\ x.\bar{\tt g}\;\&\; x.\bar{\tt g} : \bar{\tt Y}\theta 
\end{array}
}
$$

\noindent (These assumptions enable us to state in a simple way that
the constraints on fields of a class should be consistent. For, by
requiring that {\tt x:C} be consistent, we require that the
conjunction of all the constraints on the fields of {\tt x} is
consistent, since a constraint is consistent iff its completion under
entailment is consistent.)

Typing judgements for expressions are of the form $\tt \Gamma \vdash
e:T$, read as ``Under the assumptions $\Gamma$, {\tt e} has type {\tt
T}'', and thought of as defining an inference relation that extends
the underlying constraint system.  We use sequence notation in the
obvious way: $\tt \Gamma \vdash \bar{x}:\bar{X}$ is shorthand for the
collection of typing judgements $\tt \Gamma \vdash x_1:X_1 \ldots
\Gamma \vdash x_n:X_n$.

\subsubsection{Type well-formedness rules}

We shall need the judgement $\tt \Gamma \vdash T\star$, read as ``Under
the assumptions $\Gamma$, {\tt T} is a valid type'', and the judgement
$\tt \Gamma \vdash val\ c$, read as ``Under the assumptions
$\Gamma$, {\tt c} is a constant term (not dependent on mutable fields).''


The rules for establishing {\tt val c} are straightforward:
$$
\begin{array}{llll}
\from{}\infer{
  \begin{array}{l}
    \tt \vdash val\ null\\
    \tt \vdash val\ true\\
    \tt \vdash val\ false\\
 \tt  \vdash val\ n\\
 \tt x:X \vdash val\ x 
  \end{array}}&
\from{\tt \Gamma \vdash X\star\ \ \Gamma \vdash val\ c}\infer{\tt \Gamma \vdash val\ (\exists x:X) c}&
\from{\tt \Gamma \vdash val\ e}\infer{\tt \Gamma \vdash val\ op\ e}&
\from{\tt \Gamma \vdash val\ e_0\;\&\; val\ e_1}\infer{\tt \Gamma \vdash val\ e_0\ op\ e_1}\\
\end{array}
$$

The only inference rule for establishing $\tt Z\star$ is:
$$
\from{\tt \Gamma, self:C \vdash c:boolean \;\&\;val\ c}
\infer{\tt \Gamma \vdash C(\&c)\ \star}
$$

\subsubsection{Subtyping rules}
We will also use the sub-typing judgement $\tt \Gamma \vdash X \leq Y$
on types, read as ``Under the assumptions $\Gamma$ the type $\tt X$ is
a sub-type of $\tt Y$.

$$
\begin{array}{llll}
\tt \Gamma \vdash C \leq C
& \tt 
\from{\tt \Gamma \vdash C \leq D \ \ \ \Gamma \vdash D \leq E}
\infer{\tt\Gamma \vdash C \leq E}
&
\from{\tt class\ C(\ldots)\ extends\ D\ \ldots}
\infer{\tt\Gamma \vdash C \leq D}
&
\from{\tt \Gamma \vdash C \leq D \ \ \ \Gamma, c \vdash d}
\infer{\tt\Gamma \vdash C(\&c) \leq D(\&d)}
\end{array}
$$

Under the assumptions of well-formed programs, this relation is acyclic.

\subsubsection{Static semantics rules}
First we cover the rules for literals:
$$
\begin{array}{|ll|}\hline
\rname{\sc\tt (T-Lit)}\from{}\infer{
  \begin{array}{l}
    \tt \vdash null: nullable\ T \\
    \tt \vdash n:int(n)    \\
    \tt \vdash true:boolan(true)  \\  
    \tt \vdash false:boolan(false)    
  \end{array}
}  &
\rname{\sc\tt (T-Nullable)}\from{\tt\Gamma \vdash e:T}\infer{\tt \Gamma \vdash e: nullable\ T}  \\
\quad &\\ \hline
\end{array}
$$

\noindent No rules are needed for typing field access or for
establishing {\tt x:T}; these are covered by the underlying constraint system.

$$
\begin{array}{|ll|}\hline
\rname{\sc\tt (T-New)}
\from{
  \begin{array}{l}
   \tt class\ C( \bar{\tt f}:\bar{\tt X},\, var\ \bar{\tt g}:\bar{\tt Y})\; extends\ D \ldots \\
   \tt \bar{\tt h}:\bar{\tt Z} = fields(D) \\
  \tt \Gamma \vdash \bar{\tt x},\bar{\tt y},\bar{\tt z}:\bar{\tt X}_1,\bar{\tt Y}_1,\bar{\tt Z}_1 \\ 
  \tt \theta= [ \bar{\tt x},\bar{\tt y},\bar{\tt z}/\bar{\tt f},\bar{\tt g},\bar{\tt h}] \\
  \tt \Gamma \vdash \bar{\tt X}_1,\bar{\tt Y}_1,\bar{\tt Z}_1 \leq 
                    \bar{\tt X}\theta,\bar{\tt Y}\theta,\bar{\tt Z}\theta  \\
  \end{array}}
\infer{\tt \Gamma \vdash new\ C(\bar{\tt x}, \bar{\tt y}, \bar{\tt z})
  :C(\&\bar{\tt f}, \bar{\tt g}, \bar{\tt h} == \bar{\tt x}, \bar{\tt y}, \bar{\tt z})
 }
&
\rname{\sc\tt (T-Invoke)}
\from{
  \begin{array}{l}
   \tt \Gamma \vdash x,\bar{y}:X,\bar{Y} \\
   \tt m(\bar{z}:\bar{Z}\&c):T \in mType(X) \\
   \tt \theta=[x/this,\bar{y}/\bar{z}] \\
   \begin{array}{ll}
   \tt \Gamma \vdash \bar{Y} \leq \bar{Z}\theta\;\&\; c\theta
   \end{array}
 \end{array}}
\infer{\tt \Gamma \vdash x.m(\bar{y}):T\theta}\\
\quad &\\
\rname{\sc\tt (T-Local)}
\from{
  \begin{array}{l}
    \tt \Gamma \vdash X\star \\
    \tt \Gamma,x:X\ \mbox{\em satisfiable}\\
    \tt \Gamma,x:X \vdash d:Y\; \&\; Y \leq X\; \&\; e:Z
  \end{array}
}
\infer{\tt \Gamma \vdash \{val\ x:X=d;e\}:\exists (x:X)\,Z} &\\
\quad&\\
\multicolumn{2}{|c|}{
\begin{array}{lll}
\rname{\sc\tt (T-UCast)}\from{\tt \Gamma \vdash Y\star \ \ \ \Gamma \vdash e:X \;\&\;X\leq Y} 
\infer{\tt \Gamma \vdash (Y) e: Y}  &
\rname{\sc\tt (T-GCast)}\from{\tt \Gamma \vdash Y\star \ \ \ \Gamma \vdash e:X \ \ \ \Gamma \not\vdash X\leq Y} 
\infer{\tt \Gamma \vdash (Y) e: Y}  &
\end{array}}

\\
\quad &\\
\multicolumn{2}{|c|}{
\rname{\sc\tt (T-Typecase)}
\from{
  \begin{array}{lll}
   \tt \Gamma \vdash ce:boolean &  \tt \Gamma, ce \vdash r:T &  \tt \Gamma, !ce \vdash s:T
 \end{array}}
\infer{\tt \Gamma \vdash (ce \implies r:s) :T}
}\\
\quad & \\
\rname{\sc\tt (T-Method)}
\from{
  \begin{array}{l}
  \tt M \in \mbox{methods}(C) \\
  \tt M= def\ m(\bar{\tt z}:\bar{\tt Z}\&c):T=e\\
   \tt this:C \vdash \bar{\tt Z}\star \\
   \tt \bar{\tt z}:\bar{\tt Z}, this:C \vdash c : boolean \;\&\;val\ c\\
   \tt \bar{\tt z}:\bar{\tt Z}, this:C, c\ \mbox{\em satisfiable}\\
   \tt \bar{\tt z}:\bar{\tt Z}, this:C, c\vdash e:T 
 \end{array}}
\infer{\tt M\ OK\ in\ C}
&
\rname{\sc\tt (T-Class)}
\from{
  \begin{array}{l}
%\tt \bar{\tt f}:\bar{\tt X},\,\tt\bar{\tt g}:\bar{\tt Y}\ \mbox{\em satisfiable}\\
\tt  this:C \vdash \bar{\tt X}\star\;\&\;\bar{\tt Y}\star\\
\tt \bar{\tt f}:\bar{\tt X},\,var\ \bar{\tt g}:\bar{\tt Y}\ \mbox{\em satisfiable}\\
\tt  D\ OK   \ \ \ \tt \bar{\tt M}\ OK \ in \ C\\
 \end{array}}
\infer{\tt class\ C(\bar{\tt f}:\bar{\tt X},\, var\, \tt\bar{\tt g}:\bar{\tt Y})\,
    extends\ D \{\bar{\tt M}\}\ OK} \\
\quad &\\ \hline
\end{array}
$$

Finally we cover the rules for field assignment and sequencing. These are standard.

$$
\begin{array}{|ll|}\hline
\rname{\sc\tt (T-Field-w)}
\from{
  \begin{array}{lll}
  \tt  \Gamma \vdash var\; x.g\;\&\;x.g:U &  \tt \Gamma \vdash e:V & \tt \Gamma \vdash V \leq U
 \end{array}}
\infer{\tt \Gamma \vdash x.g=e:U} &
\rname{\sc\tt (T-Seq)}
\from{
  \begin{array}{ll}
   \tt \Gamma \vdash d:S   & 
   \tt \Gamma \vdash e:T
 \end{array}}
\infer{\tt \Gamma \vdash d;e:T}\\ \hline
\end{array}
$$

%%\subsubsection{Auxiliary definitions}
%%The rules use some auxiliary definitions. 
%%For any type {\tt T}, {\tt fields(T)} is the set of typed fields for that type:
%%$$
%%\begin{array}{l}
%%\begin{array}{ll}
%%\tt fields(Object)=\emptyset & \tt fields(C(\&c))=fields(C)
%%\end{array} \\
%%\tt fields(C)=fields(D),\bar{\tt x}:\bar{\tt X},\bar{\tt y}:\bar{\tt Y}
%%\ \ \mbox{\em if 
%%{\tt class C($\tt \bar{\tt x}:\bar{\tt X},var\ \bar{\tt y}:\bar{\tt Y}$\&c) extends D\ldots}}  
%%\end{array}
%%$$
%%
%%%%\noindent Given two sets $S$ and $T$ of field type assertions, the set $S,T$
%%%%has all the elements of $T$ and all the elements of $S$ except those
%%%%whose variable is the same as a variable of an element in $T$. Thus
%%%%the above definitions implement field overriding: fields of {\tt C}
%%%%with the same name hide the fields of {\tt D} with the same name (even
%%%%if they have different types).
%%
%%For any type {\tt T}, {\tt varfields(T)} is the set of mutable fields for that type:
%%$$
%%\begin{array}{l}
%%\begin{array}{ll}
%%\tt varfields(Object)=\emptyset & \tt varfields(C(\&c))=varfields(C)
%%\end{array} \\
%%\tt varfields(C)=varfields(D)[\bar{y}:\bar{Y}]\ \ \mbox{\em if 
%%{\tt class C($\tt \bar{\tt x}:\bar{\tt X},var\ \bar{\tt y}:\bar{\tt Y}$\&c) extends D\ldots}}  
%%\end{array}
%%$$
%%
%%The set of constructors of a type {\tt T=C(\&c)} ({\tt ctor(C)}) is
%%the set of all elements {$\tt C(\bar{\tt x}:\bar{\tt X}\&c):T$} such
%%that the constructor {$\tt def C(\bar{\tt x}:\bar{\tt
%%X}\&c):T$=\ldots} occurs in the definition of the class {\tt C}. We omit a formal definition.
%%The set of methods of a type {\tt T=C(\&c)}, {\tt mType(T)}, is the
%%set of all elements {$\tt m(\bar{z}:\bar{Z}\&c):T$}  such that the method
%%{$\tt def m(\bar{z}:\bar{Z}\&c):T=e$} occurs in the definition of the class
%%{\tt C} or an ancestor class. We omit a formal definition.

\subsection{Example revisited}

Here we consider a concrete constraint system with the type {\tt int}
(with arithmetic operations).

{\em Show the List example type-checks.}

\subsection{Dynamic semantics}

We provide a dynamic semantics as a binary transition relation in the
usual structural operational semantics style.  We consider first the
assignment-free sublanguage (the language without field-assignment,
and sequencing).  For such a language there is no reason to introduce
heaps; the semantics can be specified by a transition relation on just
expressions.  The computation rules are not very different from
\cite{FJ}; in fact they are simpler because we permit only local
constants as receivers of method calls and arguments of method calls
and constructors.

First we define the notion of {\em canonical values}:

\begin{tabular}[t]{|llll|}\hline
 (Canonical Value) & v,w &{::=}& \tt null \alt true \alt false \alt n \alt new C($\bar{\tt v}$) 
\\ \hline
\end{tabular}

\vskip 10pt
We shall see that values are {\em terminal} in the dynamic semantics
(they cannot evolve).

$$
\begin{array}{|l|} \hline\\
  \begin{array}{ll}
    \rname{\tt\sc (R-Field)}
    \from{\tt fields(C)=\bar{\tt f}:\bar{\tt X}}
    \infer{\tt (new\ C(\bar{\tt v})).f_i \derives v_i} &
    \rname{\tt\sc (R-Invoke)}
    \from{\tt mbody(m,C)=(\bar{\tt x}, e)}
    \infer{\tt (new\ C(\bar{\tt v})).m(\bar{\tt w}) \derives e[\bar{\tt w}/\bar{\tt x},\,new\ C(\bar{\tt v})/this]}\end{array}\\
\quad\\
\begin{array}{ll}
  \rname{\tt\sc (R-Cast-Null)}\from{}\infer{\tt (nullable\ T)\; null \derives null} & 
  \rname{\tt\sc (R-Cast)}
  \from{
    \begin{array}{ll}
      \tt C \leq D & \tt c[new\ C(\bar{\tt v})/self]\starderives true      
    \end{array}}
  \infer{
    \begin{array}{l}
      \tt (D(\&c))(new\ C(\bar{\tt v})) \derives new\ C(\bar{\tt v}) \\
      \tt (nullable\ D(\&c))(new\ C(\bar{\tt v})) \derives new\ C(\bar{\tt v}) 
  \end{array}} 
\end{array}\\
\quad\\
  \rname{\tt\sc (R-Local)}
  \from{}
  \infer{\tt \{val\ x:T=v;\; e\} \derives e[v/x]} \\
\quad\\
\begin{array}{ll}
  \rname{\tt\sc (R-Typecase)}
  \from{}
  \infer{
    \begin{array}{l}
      \tt  true \implies e_0 : e_1 \derives e_0\\
      \tt  false \implies e_0 : e_1 \derives e_1\\
  \end{array}} &
  \rname{\tt\sc (R-Primitive)}
    \from{}
    \infer{\begin{array}{l}
	\tt v\ op\ w \derives z \ \ \ \mbox{({\tt z} is result of {\tt v op w})} \\
	  \tt op\ v  \derives  z \ \ \ \mbox{({\tt z} is result of {\tt op v})}
	\end{array}} \\
\end{array}\\ 
\hline
\end{array}
$$

Finally, we have the context rules:

$$
\begin{array}{|l|}\hline\\
\rname{\tt\sc (R-Context)}
\from{\tt r \derives r'}
\infer{
  \begin{array}{l}
    \tt \{val\ x:T=r;\, e\} \derives \{val\ x:T=r';\, e\} \\
    \tt r \implies e_0 : e_1  \derives r' \implies e_0 : e_1 \\
    \tt (T)\; r \derives (T)\; r'
  \end{array}}\\
\hline
\end{array}
$$

\subsection{Properties of the typing system}
Below, we say that $\tt e$ is {\em stuck} if there is no $\tt r$ such that
$e \derives r$.

{\em Proofs of these lemmas and theorems are being worked on.}

\begin{lemma}[Well-formedness] If $\Gamma \vdash e:T$ then $\Gamma$ is satisfiable, 
and $\Gamma \vdash T\star$.
\end{lemma}

\begin{lemma}[Term substitution preserves types] If $\tt \Gamma, x:X \vdash e:T$
and $\tt \Gamma \vdash r:Y$ such that $\tt\Gamma \vdash Y \leq X$ then 
$\tt\Gamma \vdash e[r/x]:T'$ for some $\tt T' \leq T$.
\end{lemma}

\begin{theorem}[Type soundness] $\Gamma \vdash e':T'$ whenever 
for some $T$ s.t. $\Gamma \vdash T' \leq T$, $\Gamma\vdash e:T$ and
$e \derives e'$.
\end{theorem}

Say that $\Gamma \vdash_s e$ (read: ``{\tt e}'' is safely well-typed in $\Gamma$) if
for some $T$, $\Gamma \vdash_e e:T$ and the derivation does not use rule
{\tt \sc (T-GCAST)}. (Invocations of the cast operation ``checked'' by such a rule can fail
dynamically.) 

\begin{theorem}[Safety] If $\Gamma \vdash_s  e$ and $e \vdash e'$ then
$\Gamma \vdash_s e'$. If $\tt \Gamma \vdash_s e$ and $\tt e$ is stuck then $\tt e$ is a value.
\end{theorem}

\section{\FXten{} with places}

{\em
  \begin{itemize}
    \item Follow the outline of my recent set of slides.
    \item Every reference object now has a final {\tt place} field, {\tt loc}.
    \item The constant {\tt here} evaluates to the current place.
    \item {\tt place} is a value type, with a fixed but unknown set of
      operations, we can assume {\tt .next:place}.
    \item The type {\tt T@!} is represented by {\tt T(\&loc=here}), and
      the type {\tt T@x} by {\tt T(\&loc=x.loc)}.
    \item The rules for field read/write and method invocation are changed to require that
      the subject be of type {\tt \_(\&loc=here)}.
    \item The new place-shifting control construct {\tt eval(p)e} is introduced.
  \end{itemize}
}

%\section{\FXten{} with regions}

%\section{\FXten{} with distributions}

\bibliographystyle{plain}
\bibliography{master}
\appendix
\section{An extended example}
{\footnotesize
\begin{verbatim}
/**
   A distributed binary tree.
   @author Satish Chandra 4/6/2006
   @author vj
 */
//                             ____P0
//                            |     |
//                            |     |
//                          _P2  __P0
//                         |  | |   |
//                         |  | |   |
//                        P3 P2 P1 P0
//                         *  *  *  *
// Right child is always on the same place as its parent; 
// left child is at a different place at the top few levels of the tree, 
// but at the same place as its parent at the lower levels.

class Tree(localLeft: boolean, 
           left: nullable Tree(& localLeft => loc=here),
           right: nullable Tree(& loc=here),
           next: nullable Tree) extends Object {
    def postOrder:Tree = {
        val result:Tree = this;
        if (right != null) {
            val result:Tree = right.postOrder();
            right.next = this;
            if (left != null) return left.postOrder(tt);
        } else if (left != null) return left.postOrder(tt);
        this
    }
    def postOrder(rest: Tree):Tree = {
        this.next = rest;
        postOrder
    }                    
    def sum:int = size + (right==null => 0 : right.sum()) + (left==null => 0 : left.sum)
}
value TreeMaker {
    // Create a binary tree on span places. 
    def build(count:int, span:int): nullable Tree(& localLeft==(span/2==0)) = {
        if (count == 0) return null;    
        {val ll:boolean = (span/2==0);
         new Tree(ll,  eval(ll => here : place.places(here.id+span/2)){build(count/2, span/2)},
           build(count/2, span/2),count)}
    }
}
\end{verbatim}}
\end{document}
