\documentclass[9pt]{sigplanconf}


\conferenceinfo{X10'11,} {June 4, 2011, San Jose, California, USA.}
\CopyrightYear{2011}
%\copyrightdata{}

% \stepcounter Doesn't work in a caption!
\newcounter{RuleCounter}
\stepcounter{RuleCounter}
\newcommand{\userule}[1]{\arabic{#1}}
\newcommand{\definerule}[1]{\newcounter{#1}\addtocounter{#1}{\arabic{RuleCounter}}\stepcounter{RuleCounter}}
\newcommand{\myrule}[2]{\textbf{Rule #1:} #2.}

\input{commands}

\begin{document}


\lstset{language=java,basicstyle=\ttfamily\small}


\title{Object Initialization in X10}


\authorinfo{Yoav Zibin \and David Cunningham \and Igor Peshansky \and Vijay Saraswat}
           {IBM research in TJ Watson}
           {yzibin$~|~$dcunnin$~|~$igorp$~|~$vsaraswa@us.ibm.com}


\maketitle


\begin{abstract}
\input{abstract}
\end{abstract}

%\category{D.3.3}{Programming Languages}{Language Constructs and Features}
%\category{D.1.5}{Programming Techniques}{Object-oriented Programming}

%\terms
%Asynchronous, Initialization, Types, X10

% Keywords are not required in the paper itself - only in the submission system's meta data.
% \keywords
% Immutability, Ownership, Java


\Section[introduction]{Introduction}
\input{introduction}

\Section[rules]{X10 Initialization Rules}
\input{rules}

\vspace{-0.3cm}
\Section[designs]{Alternative Initialization Designs}
\input{designs}

%\Section[implementation]{Implementation}
%\input{implementation}

%\Section[case-study]{Case Study}
%\input{case-study}

\Section[related-work]{Related Work}
\input{related-work}

\Section[conclusion]{Conclusion}
\input{conclusions}


\bibliographystyle{abbrv} %plain plainnat abbrvnat abbrv
\bibliography{x10-init}

\end{document}


Tested my serialization hypotheses using this Java code:
        ObjectOutput out = new ObjectOutputStream(new FileOutputStream("filename.ser"));
        out.writeObject("C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10"+"C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.tests,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.dist\\samples,C:\\cygwin\\home\\Yoav\\intellij\\sourceforge\\x10.runtime\\src-x10".substring(1));
        out.close(); // 2KB!!
        ObjectOutput out2 = new ObjectOutputStream(new FileOutputStream("filename2.ser"));
        out2.writeObject("C".substring(1));
        out2.close(); // 7bytes!


\subsection{Alternative property design}

Property design (what would happen if we use val fields instead of properties.)

Fields are partition into two: property fields and normal fields. Property fields are assigned together before any other field.
\begin{lstlisting}
class A(x:Int) {
  val y:Int{self!=x};
  val z = x*x;
}
\end{lstlisting}
Vs.
Fields are ordered, and all ctors need to assign in the same order as the fields declaration.
\begin{lstlisting}
class A {
  val x:Int;
  val y:Int{self!=x};
  val z = x*x;
}
\end{lstlisting}



\subsection{Static initialization}
X10 does not support dynamic class loading as opposed to Java,
    and all static fields in X10 are final.
Thus, initialization of static fields is a one-time phase, denoted the static-init phase,
    that is done before the \code{main} method is executed.

During the static-init phase we must finish writing to all static fields,
    and reading a static field \emph{waits} until the field is initialized
    (i.e., the current activity/thread blocks if the field was not written to,
    and it resumes after another activity writes to it).
Obviously, this may lead to deadlock as demonstrated by \Ref{Figure}{Static-init}.
However, in practice, deadlock is rare,
    and usually found quickly the first time a program is executed.

\begin{figure}
\begin{lstlisting}
class A {
  static val a:Int = B.b;
}
class B {
  static val b:Int = A.a;
}
\end{lstlisting}
\caption{Static initialization example:
    the program will deadlock at run-time
    during the static-init phase (before the \code{main} method is executed).
    }
\label{Figure:Static-init}
\end{figure}


\subsection{Memory model and constructor barrier}

todo: discuss final in Java and String class.
(people think that removing final may only hurt performance, but it may be semantic changing.)

type safety and the weak memory model:
* in Java if you don't use final fields correctly or leak this, you will simply see default values (you don't lose type-safety)
* in X10 it could break type safety (if we don't put a barrier at the end of a ctor).

class Box[T] {T haszero} {
  var value:T;
}
class A {
  static val box = new Box[A]();

  var f:Int{self!=0} = 1;
}

var a = new A();
a.f = 2;
A.box.value = a;

Suppose another activity reads A.box.value.
Should the writes to
a.f
and
A.box.value
be ordered? (I don't think we should order them without losing performance)

Therefore, X10 needs a synchronization barrier at the end of a ctor that guarantees that all writes to the fields (both VAL and VAR) of the object has finished before the handle is returned.
(This is different from Java that only promises this for final fields. And the barrier also happens again after deserialization - requiring this weird freeze operation and you could freeze a final field again even after the ctor ended due to deserialization.)


Bowen proposes this transformation in order to inline ctors:
class A {
  static val box = new Box[A]();

  var f:Int{self!=0} = 0;
}


var a = new A();
a.f = 1;
// INSERT BARRIER EXPLICITLY HERE
a.f = 2;
A.box.value = a;
