
\begin{abstract}

Obtaining practical efficient implementations for large, irregular
graph problems is challenging. Current software systems and commodity
multiprocessors do not support fine-grained, irregular parallelism
well. Implementing a custom framework for fine-grained parallelism for
each new graph algorithm is impractical.

We present \XWS{}, the \Xten{} Work Stealing framework.
\XWS{} is intended as an open-source runtime for the programming
language \Xten{}, a partitioned global address space language
supporting dynamic fine-grained concurrency.  \XWS{} is also intended
as a library to be used directly by application writers. \XWS{}
extends the Cilk work-stealing framework with several features
necessary to efficiently implement graph algorithms, viz., support for
improperly nested procedures, worker-specific data-structures, global
termination detection, and phased computation.

We present simple elegant programs using \XWS{} for different spanning
tree algortihms using a (pseudo-)depth first search, and breadth-first
search.  We evaluate these programs on a 32-way Niagara (moxie), and
an 8-way Opteron server (altair) and on three different bounded-degree
graphs: (i) graphs with randomly selected edges and (a) no degree
restrictions (b) fixed degree, and (ii) planar torus graphs.

We show the performance of BFS and pseudo-DFS search depends crucially
on the grannularity of parallel tasks. We show that the granularity
natural to the algorithms -- the examination of a single edge ---
leads to poor performance at scale. Instead, sets of of vertices must
be grouped into {\em batches}. We show that a fixed-size batching
scheme does not perform well. Instead we develop an adaptive batching
scheme that is sensitive to the instantaneous size of the work queue.
With this scheme, pseudo-DFS shows linear scaling on altair and moxie,
achieving peak performance of over 220 Million edges/s on moxie.  These
programs compare favorably with separate C and Cilk implementations.


We conclude that a fine-grained concurrent language such as \Xten{}
with a work-stealing based scheduler may provide an attractive
framework for the implementation of graph algorithms.

\end{abstract}

