We present a general framework for \emph{generic constrained types}
that captures the notion of value-dependent and
type-dependent (i.e., generic) type systems for object-oriented
languages.  Constraint systems formalize systems of partial
information.  Constrained types are formulas \Xcd{C\{c\}} where
\Xcd{C} is the name of a class or an interface and \Xcd{c} is a
constraint on the immutable state of an instance of \Xcd{C} (the
{\em properties}).

The basic idea is to formalize the essence of nominal
object-oriented types as a constraint system, and to permit both
value and type properties and parameters.  Type-generic
dependence is now expressed through constraints on these
properties and parameters.  Type-valued properties are required
to have a run-time representation---the run-time semantics is
not defined through erasure.

Many type systems for object-oriented languages developed over
the last decade can be thought of as constrained type systems in
this formulation.  This framework is parametrized by an
arbitrary constraint system $\cal C$ of interest.  It permits
the development of languages with pluggable type systems, and
supports dynamic code generation to check casts at run-time.

The paper makes the following contributions: (1) We show how to
accommodate generic object-oriented types within the framework of
constrained types. (1) We illustrate the type system with the
development of a formal calculus \FX{} and establish type
soundness. (3) We discuss the design and implementation of the
type system for \Xten{}, a modern object-oriented language, based on
constrained types.  The type system integrates and extends the
features of nominal types, virtual types, and
Scala's path-dependent types, as well as representing generic
types.
